### 2017-06-24 doc/manuscript/paper.tex
Starting the paper

\documentclass[oupdraft]{bio}
% \usepackage[colorlinks=true, urlcolor=citecolor, linkcolor=citecolor, citecolor=citecolor]{hyperref}
\usepackage{url}

% Add history information for the article if required
\history{Received XXX;
revised XXX;
accepted for publication XXX}

\newcommand{\todo}[1]{\textbf{[TODO: #1]}}
\newcommand{\note}[1]{\textbf{[NOTE: #1]}}
\newcommand{\fixme}[1]{\textbf{[FIXME: #1]}}

\begin{document}

% Title of paper
\title{Exploration of empirical Bayes hierarchical modeling for the
analysis of genome-wide association study data}

% List of authors, with corresponding author marked by asterisk
\author{Nelle Varoquaux$^\ast$,\\[4pt]
% Author addresses
\textit{Department of Statistics, University of California, Berkeley \\
Berkeley Data Science Fellow}
\\[2pt]
% E-mail address for correspondence
{eaheron@tcd.ie}}

% Running headers of paper:
\markboth%
% First field is the short list of authors
{Nelle Varoquaux}
% Second field is the short title of the paper
{XXX}

\maketitle

% Add a footnote for the corresponding author if one has been
% identified in the author list
\footnotetext{To whom correspondence should be addressed.}

\begin{abstract}
{My very cool abstract.}
{Hi-C; three-dimensional structure; and other cool stuff.
}
\end{abstract}

\section{Introduction}
\label{sec1}

\section{Methods}
\label{sec2}

\subsection{Analysing contact count data}
- compartment analysis
- extracting some relationships between counts genomic distances etc
- fit-HiC

\subsection{3D structure inference}
- MDS
- PO
- NB

\subsubsection{Structure stability}
- Structure stability
- 
\subsubsection{Volume exclusion models}
\subsection{Integrative analysis of gene expression and 3D structure using KernelCCA}

\subsection{Differential analysis}
\note{XXX only possible if Bunnik 2017 paper published.}

\section{Results}
\label{sec3}

\subsection{Simulation study application}

\section{Discussion}
\label{sec4}

\section{Software}
\label{sec5}

Software in the form of R code, together with a sample
input data set and complete documentation is available on
request from the corresponding author (eaheron@tcd.ie).

\section{Supplementary Material}
\label{sec6}

Supplementary material is available online at
% \href{http://biostatistics.oxfordjournals.org}%
% {http://biostatistics.oxfordjournals.org}.
\url{http://biostatistics.oxfordjournals.org}.

\section*{Acknowledgments}

The authors thank XXX

Funding for the project was provided by BIDS.
{\it Conflict of Interest}: None declared.

\bibliographystyle{biorefs}
\bibliography{refs}

\begin{figure}[!p]
\centering
%\includegraphics{fig1}
\caption{}
\label{Fig1}
\end{figure}

\end{document}

### 2017-06-24 scripts/structures/infer_structures_mds.py
Added MDS inference

@@ -7,10 +7,10 @@ from minorswing import dispersion
 #from minorswing._inference import negative_binomial
 from minorswing._inference import negative_binomial_structure
 #from pastis.optimization import negative_binomial_structure
-from minorswing._inference import poisson_inference
 from pastis.optimization import mds
 from minorswing._inference import utils
 import iced
+from utils import compute_wish_distances
 
 """
 Launches the inference of the 3D model on .matrix/.bed files"""
@@ -61,8 +61,9 @@ counts[np.arange(len(counts)), np.arange(len(counts))] = 0
 counts = iced.filter.filter_low_counts(
     counts, lengths=lengths,
     sparsity=False, percentage=0.03)
-normed, bias = iced.normalization.ICE_normalization(counts,
-                                                    output_bias=True)
+normed, bias = iced.normalization.ICE_normalization(
+    counts,
+    output_bias=True)
 
 counts[np.isnan(counts)] = 0
 normed[np.isnan(normed)] = 0
@@ -80,29 +81,17 @@ normed.eliminate_zeros()
 normed = normed.tocoo()
 
 # Compute starting point
-print "Initializing"
+print "Estimating structure"
+
 random_state = np.random.RandomState(args.seed)
-X = mds.estimate_X(normed, random_state=random_state)
-X[np.isnan(X)] = 0
+
+wd = compute_wish_distances(normed)
+X = mds.estimate_X(wd, random_state=random_state,
+                   precompute_distances=True)
 
 # PM2
 outname = filename.replace(
-    ".matrix", "_PO_%02d_structure.txt" % (args.seed, ))
-
-if os.path.exists(outname):
-    print "Already computed"
-    import sys
-    sys.exit()
-
-alpha = -3.
-max_iter = 1000000
-
-counts = counts.tocoo()
-X = poisson_structure.estimate_X(
-    counts, alpha=-3., beta=1., bias=bias,
-    verbose=1,
-    maxiter=max_iter,
-    ini=X.flatten())
+    ".matrix", "_MDS_%02d_structure.txt" % (args.seed, ))
 
 counts = np.array(counts.todense())
 mask = (np.array(counts.sum(axis=0)) +

### 2017-06-24 scripts/structures/utils.py

@@ -1,0 +1,14 @@
+from scipy import sparse
+import numpy as np
+
+
+def compute_wish_distances(counts, alpha=-3.0, beta=1., use_zero_counts=True):
+    counts = counts.toarray()
+    m = (counts.sum(axis=0) + counts.sum(axis=1)) == 0
+    wish_distances = counts.copy() / beta
+    wish_distances[wish_distances != 0] **= 1. / alpha
+    if use_zero_counts:
+        wish_distances[counts == 0] = wish_distances[counts != 0].max()
+    wish_distances[m] = 0
+    wish_distances[:, m] = 0
+    return sparse.coo_matrix(np.triu(wish_distances, 1))

### 2017-06-24 scripts/structures/cluster_scripts/slurm_mds.sh
Added cluster scripts to do MDS

@@ -1,0 +1,8 @@
+#! /bin/sh
+
+source activate minorswing
+cd $SLURM_SUBMIT_DIR
+
+filename=`data/ay2013/${DATASET}_10000_raw.matrix`
+python infer_structures_mds.py  $filename \
+  --seed ${SLURM_ARRAY_TASK_ID}

### 2017-06-24 scripts/structures/launch_all_jobs.sh

@@ -5,6 +5,12 @@ FOLDERS=(rings trophozoites schizonts)
 for FOLDER in ${FOLDERS[@]}; do
   sbatch --export=DATASET=${FOLDER} -v -p RM-shared --array 1-1 --mem 1500Mb \
     --job-name ${FOLDER} \
+    --time=8:00:00 cluster_scripts/slurm_mds.sh \
+    -N 1 \
+    --mail-type=ALL --mail-user=nelle@berkeley.edu;
+
+  sbatch --export=DATASET=${FOLDER} -v -p RM-shared --array 1-1 --mem 1500Mb \
+    --job-name ${FOLDER} \
     --time=8:00:00 cluster_scripts/slurm_nb0.sh \
     -N 1 \
     --mail-type=ALL --mail-user=nelle@berkeley.edu;

### 2017-06-24 scripts/structures/infer_structures_mds.py
FIX in mds

@@ -87,7 +87,7 @@ random_state = np.random.RandomState(args.seed)
 
 wd = compute_wish_distances(normed)
 X = mds.estimate_X(wd, random_state=random_state,
-                   precompute_distances=True)
+                   precompute_distances="precomputed")
 
 # PM2
 outname = filename.replace(

### 2017-06-24 scripts/structures/infer_structures_po.py
FIX structures PO

@@ -4,10 +4,7 @@ import os
 import argparse
 from pastis import fastio
 from minorswing import dispersion
-#from minorswing._inference import negative_binomial
-from minorswing._inference import negative_binomial_structure
-#from pastis.optimization import negative_binomial_structure
-from minorswing._inference import poisson_inference
+from minorswing._inference import poisson_structure
 from pastis.optimization import mds
 from minorswing._inference import utils
 import iced

### 2017-06-24 scripts/structures/launch_all_jobs.sh

@@ -4,19 +4,19 @@ FOLDERS=(rings trophozoites schizonts)
 # FOLDERS=(HP1_KD HP1_tagged pv_sporozoite)
 for FOLDER in ${FOLDERS[@]}; do
   sbatch --export=DATASET=${FOLDER} -v -p RM-shared --array 1-1 --mem 1500Mb \
-    --job-name ${FOLDER} \
+    --job-name ${FOLDER}_MDS \
     --time=8:00:00 cluster_scripts/slurm_mds.sh \
     -N 1 \
     --mail-type=ALL --mail-user=nelle@berkeley.edu;
 
   sbatch --export=DATASET=${FOLDER} -v -p RM-shared --array 1-1 --mem 1500Mb \
-    --job-name ${FOLDER} \
+    --job-name ${FOLDER}_NB0 \
     --time=8:00:00 cluster_scripts/slurm_nb0.sh \
     -N 1 \
     --mail-type=ALL --mail-user=nelle@berkeley.edu;
 
-  sbatch --export=DATASET=${FOLDER} -v -p RM-shared --array 1-100 --mem 1500Mb \
-    --job-name ${FOLDER} \
+  sbatch --export=DATASET=${FOLDER} -v -p RM-shared --array 1-1 --mem 1500Mb \
+    --job-name ${FOLDER}_PO \
     --time=8:00:00 cluster_scripts/slurm_po.sh \
     -N 1 \
     --mail-type=ALL --mail-user=nelle@berkeley.edu;

### 2017-06-24 scripts/structures/cluster_scripts/slurm_mds.sh
FIXes in cluster scripts

@@ -3,6 +3,8 @@
 source activate minorswing
 cd $SLURM_SUBMIT_DIR
 
-filename=`data/ay2013/${DATASET}_10000_raw.matrix`
-python infer_structures_mds.py  $filename \
+filename=data/ay2013/${DATASET}_10000_raw.matrix
+
+python infer_structures_mds.py $filename \
+  --lengths data/ay2013/${DATASET}_10000_raw.bed \
   --seed ${SLURM_ARRAY_TASK_ID}

### 2017-06-24 scripts/structures/cluster_scripts/slurm_nb0.sh

@@ -3,6 +3,7 @@
 source activate minorswing
 cd $SLURM_SUBMIT_DIR
 
-filename=`data/ay2013/${DATASET}_10000_raw.matrix`
+filename=data/ay2013/${DATASET}_10000_raw.matrix
 python infer_structures_nb.py  $filename \
+  --lengths data/ay2013/${DATASET}_10000_raw.bed \
   --seed ${SLURM_ARRAY_TASK_ID}

### 2017-06-24 scripts/structures/cluster_scripts/slurm_po.sh

@@ -3,6 +3,8 @@
 source activate minorswing
 cd $SLURM_SUBMIT_DIR
 
-filename=`data/ay2013/${DATASET}_10000_raw.matrix`
+filename=data/ay2013/${DATASET}_10000_raw.matrix
+
 python infer_structures_po.py  $filename \
+  --lengths data/ay2013/${DATASET}_10000_raw.bed \
   --seed ${SLURM_ARRAY_TASK_ID}

### 2017-06-24 scripts/structures/launch_all_jobs.sh
Cluster news

@@ -3,19 +3,19 @@
 FOLDERS=(rings trophozoites schizonts)
 # FOLDERS=(HP1_KD HP1_tagged pv_sporozoite)
 for FOLDER in ${FOLDERS[@]}; do
-  sbatch --export=DATASET=${FOLDER} -v -p RM-shared --array 1-1 --mem 1500Mb \
+  sbatch --export=DATASET=${FOLDER} -v -p RM-shared --array 2-100 --mem 1500Mb \
     --job-name ${FOLDER}_MDS \
     --time=8:00:00 cluster_scripts/slurm_mds.sh \
     -N 1 \
     --mail-type=ALL --mail-user=nelle@berkeley.edu;
 
-  sbatch --export=DATASET=${FOLDER} -v -p RM-shared --array 1-1 --mem 1500Mb \
+  sbatch --export=DATASET=${FOLDER} -v -p RM-shared --array 2-100 --mem 1500Mb \
     --job-name ${FOLDER}_NB0 \
     --time=8:00:00 cluster_scripts/slurm_nb0.sh \
     -N 1 \
     --mail-type=ALL --mail-user=nelle@berkeley.edu;
 
-  sbatch --export=DATASET=${FOLDER} -v -p RM-shared --array 1-1 --mem 1500Mb \
+  sbatch --export=DATASET=${FOLDER} -v -p RM-shared --array 2-100 --mem 1500Mb \
     --job-name ${FOLDER}_PO \
     --time=8:00:00 cluster_scripts/slurm_po.sh \
     -N 1 \

### 2017-06-24 doc/manuscript/paper.tex
Population of structures

@@
 
+Let us first introduce some notations. Hi-C data can be summarized as an $n
+\times n$ contact count matrix, where each row and column corresponds to a
+genomic window or loci, and each entry $c_{ij}$ refers to the number of time
+$i$ and $j$ have been seen interacting with one another.
+
+Hi-C data suffers from technical and biological biases that need to be
+corrected. We use ICE
+
 \subsection{Analysing contact count data}
@@
 - extracting some relationships between counts genomic distances etc
-- fit-HiC
+- Extracting significant interactions using fit-HiC
+
+Figure:
+  - log-linear relationships between counts and distances.
+  - significant interactions using fit-HiC ?
 
 \subsection{3D structure inference}
-- MDS
-- PO
-- NB
+
+A crucial step of our analysis of the \em{P. falciparum} genome relied on
+visual introspection of a consensus 3D model of the genome. Being able to
+quickly visualize how chromosome folded, where families of genes lied in the
+3D space or whether regions where enriched helped us integrate with other
+sources of data.
+
+Briefly, we cast the problem of finding a consensus 3D structure $X \in R^{n
+\times n}$ as an optimization problem, where we attempted to place each beads
+in 3D so that they match as closely sa possible a predefined
+``wish-distance``. We added constraints to our model based on prior knowledge,
+such as the size of the nucleus, or how far apart two adjacent beads should
+lie, though we showed these constraints where well satisfied even when
+performing an unconstrained optimized. In this section, I will review the
+inference method we used in \citet{ay:three-dimensional} as well as two newer
+methods, based on well-grounded statistical method.
+
+\subsubsection{Multidimensional scaling}
+
+\begin{equation}
+\end{equation}
+
+\subsubsection{Modeling contact counts as a Poisson distribution}
+
+\subsubsection{Modeling contact counts as a Negative Binomial distribution}
 
@@
 - Structure stability
+- at least 100 iterations
+- maybe compare results with chromSDE, PCA \& ShRec3D 
 - 
 \subsubsection{Volume exclusion models}
+
 \subsection{Integrative analysis of gene expression and 3D structure using KernelCCA}
+
+- Can we extend this to other data types? Chip-seq data available from our
+review? Can the results be interpretable?
 

### 2017-06-24 scripts/population_structures/clustering_across_stages_downsampled_distances.py

@@ -1,0 +1,128 @@
+import numpy as np
+import os
+import argparse
+from sklearn.decomposition import RandomizedPCA as PCA
+import matplotlib.pyplot as plt
+
+parser = argparse.ArgumentParser()
+parser.add_argument("--algorithm", "-a", default="MDS")
+args = parser.parse_args()
+
+datasets = ["rings", "schizonts", "trophozoites"]
+algorithm = args.algorithm
+
+labels = None
+X = None
+for i, dataset in enumerate(datasets):
+    x = np.load(
+        "results/ay2013/%s_10000_raw_%s_distances_downsampled.npy" % (
+            dataset, algorithm))
+    x = x[:100]
+    if labels is not None:
+        X[i, :len(x)] = x
+        labels = np.concatenate([labels, i * np.ones(len(x))])
+    else:
+        labels = i * np.ones(len(x))
+        X = np.nan * np.zeros((len(datasets), x.shape[0], x.shape[1]))
+        X[i, :len(x)] = x
+
+X = X.reshape(-1, x.shape[1])
+# Remove missing labels
+X = X[np.invert(np.isnan(X[:, 0]))]
+# Remove any bead with missing value
+X = X[:, np.invert(np.isnan(X.sum(axis=0)))]
+
+X /= (X).sum(axis=1)[:, np.newaxis] / 1000
+
+colors = ['#FFB773', '#FF7C00', "r",  # '#A65100',
+          '#04859D', 'skyblue',
+          '#015666', 'green', 'darkgreen']
+markers = [".", ".", ".", "D", "D", "D"]
+stages = ["Rin", "Sch", "Trop"]
+
+pca = PCA(n_components=3, copy=False, random_state=1)
+X = pca.fit_transform(X)
+fig, (ax1, ax2) = plt.subplots(
+    ncols=2,
+    figsize=(16, 6),
+    subplot_kw={"aspect": 1})
+
+for i in range(len(datasets)):
+    ax1.plot(
+        X[labels == i, 0], X[labels == i, 1],
+        marker=".",
+        markerfacecolor=colors[i],
+        markeredgecolor="white", linewidth=0, label=stages[i],
+        markeredgewidth=0,
+        markersize=5,
+        zorder=10,
+        alpha=0.5)
+    ax2.plot(
+        X[labels == i, 2], X[labels == i, 1],
+        marker=".",
+        markerfacecolor=colors[i], markeredgecolor="white",
+        linewidth=0, label=stages[i],
[…]

### 2017-06-24 scripts/population_structures/compute_distances.sh

@@ -1,0 +1,24 @@
+python compute_downsampled_distances.py \
+  structures/ay2013/trophozoites_10000_raw_MDS \
+  --lengths data/ay2013/rings_10000_raw.bed
+
+python compute_downsampled_distances.py \
+  structures/ay2013/trophozoites_10000_raw_UNB0 \
+  --lengths data/ay2013/rings_10000_raw.bed
+
+python compute_downsampled_distances.py \
+  structures/ay2013/schizonts_10000_raw_MDS \
+  --lengths data/ay2013/rings_10000_raw.bed
+
+python compute_downsampled_distances.py \
+  structures/ay2013/schizonts_10000_raw_UNB0 \
+  --lengths data/ay2013/rings_10000_raw.bed
+
+python compute_downsampled_distances.py \
+  structures/ay2013/rings_10000_raw_MDS \
+  --lengths data/ay2013/rings_10000_raw.bed
+
+python compute_downsampled_distances.py \
+  structures/ay2013/rings_10000_raw_UNB0 \
+  --lengths data/ay2013/rings_10000_raw.bed
+

### 2017-06-24 scripts/population_structures/compute_downsampled_distances.py

@@ -1,0 +1,61 @@
+import os
+import sys
+from glob import glob
+import argparse
+
+import numpy as np
+from sklearn.metrics import euclidean_distances
+from iced.utils import downsample_resolution
+from iced.io import load_lengths
+
+
+parser = argparse.ArgumentParser()
+parser.add_argument("directory")
+parser.add_argument("--lengths", "-l")
+args = parser.parse_args()
+
+
+if args.lengths is not None:
+    lengths = load_lengths(args.lengths)
+else:
+    lengths = load_lengths(os.path.join(
+        args.directory.replace("results", "data"),
+        "raw.bed"))
+
+distances = []
+
+filenames = glob(args.directory + "*_structure.txt")
+filenames.sort()
+filenames = filenames[:100]
+
+distances = []
+
+for i, filename in enumerate(filenames):
+    sys.stdout.write(
+        "\rAnalysing %0.2f %% files" % (100. * (i + 1) / len(filenames)))
+    sys.stdout.flush()
+    try:
+        X = np.loadtxt(filename)
+    except ValueError:
+        print(i, filename)
+        raise ValueError
+    mask = np.isnan(X[:, 0])
+    X[mask] = 0
+    dis = euclidean_distances(X)
+    dis[mask] = np.nan
+    dis[:, mask] = np.nan
+    dis, lengths_ = downsample_resolution(dis, lengths, factor=10)
+    distances.append(dis[np.triu_indices(dis.shape[0])][np.newaxis])
+
+distances = np.concatenate(distances)
+outfile = (
+    args.directory.replace(
+        "structures", "results") + "_distances_downsampled.npy")
+print("Saving in %s" % outfile)
+
+try:
+    os.makedirs(os.path.dirname(outfile))
+except OSError:
+    pass
+
+np.save(outfile, distances)

### 2017-06-24 scripts/population_structures/compute_pairwise_distances.py

@@ -1,0 +1,36 @@
+from __future__ import print_function
+
+import os
+from glob import glob
+import argparse
+
+import numpy as np
+from sklearn.metrics import euclidean_distances
+
+
+parser = argparse.ArgumentParser()
+parser.add_argument("directory")
+args = parser.parse_args()
+
+filenames = glob(args.directory + "*_structure.txt")
+filenames.sort()
+filenames = filenames[:100]
+
+distances = []
+for filename in filenames:
+    X = np.loadtxt(filename)
+    mask = np.isnan(X[:, 0])
+    X = X[np.invert(mask)]
+    dis = euclidean_distances(X)
+    distances.append(dis[np.triu_indices(dis.shape[0], 1)][np.newaxis])
+
+distances = np.concatenate(distances)
+outfile = args.directory.replace("structures", "results") + "_distances.npy"
+print("Saving in %s" % outfile)
+
+try:
+    os.makedirs(os.path.dirname(outfile))
+except OSError:
+    pass
+
+np.save(outfile, distances)

### 2017-06-24 scripts/population_structures/compute_pairwise_distances_chr.py

@@ -1,0 +1,40 @@
+import os
+from glob import glob
+import argparse
+
+import numpy as np
+from sklearn.metrics import euclidean_distances
+from iced.io import load_lengths
+
+
+parser = argparse.ArgumentParser()
+parser.add_argument("directory")
+args = parser.parse_args()
+
+filenames = glob(os.path.join(args.directory, "*NB0cst*_structure.txt"))
+filenames.sort()
+
+lengths = load_lengths(os.path.join(
+    args.directory.replace("results", "data"),
+    "raw.bed"))
+
+distances = [[] for _ in lengths]
+
+for filename in filenames:
+    begin, end = 0, 0
+    for l, length in enumerate(lengths):
+        end += length
+        X = np.loadtxt(filename)
+        X = X[begin:end]
+        mask = np.isnan(X[:, 0])
+        X[mask] = 0
+        dis = euclidean_distances(X)
+        dis[mask] = np.nan
+        dis[:, mask] = np.nan
+        distances[l].append(dis[np.triu_indices(dis.shape[0], 1)][np.newaxis])
+        begin = end
+
+for l, distances_chr in enumerate(distances):
+    distances_chr = np.concatenate(distances_chr)
+    np.save(os.path.join(args.directory, "distances_chr%02d.npy" % l),
+            distances_chr)

### 2017-06-24 scripts/structures/infer_structures_mds.py
Increased wall time for cluster

@@ -27,12 +27,12 @@ if args.filename.startswith("data"):
 else:
     filename = args.filename
 
-# Create name tag
-algo = "UNB"
-algo = algo + "0"
 
 outname = filename.replace(
-    ".matrix", "_%s_%02d_structure.txt" % (algo, args.seed))
+    ".matrix", "_MDS_%02d_structure.txt" % (args.seed, ))
+if os.path.exists(outname):
+    import sys
+    sys.exit(0)
 
 try:
     os.makedirs(os.path.dirname(filename))
@@ -88,10 +88,6 @@ random_state = np.random.RandomState(args.seed)
 wd = compute_wish_distances(normed)
 X = mds.estimate_X(wd, random_state=random_state,
                    precompute_distances="precomputed")
-
-# PM2
-outname = filename.replace(
-    ".matrix", "_MDS_%02d_structure.txt" % (args.seed, ))
 
 counts = np.array(counts.todense())
 mask = (np.array(counts.sum(axis=0)) +

### 2017-06-24 scripts/structures/infer_structures_nb.py

@@ -33,6 +33,10 @@ algo = algo + "0"
 outname = filename.replace(
     ".matrix", "_%s_%02d_structure.txt" % (algo, args.seed))
 
+if os.path.exists(outname):
+    import sys
+    sys.exit(0)
+
 try:
     os.makedirs(os.path.dirname(filename))
 except OSError:

### 2017-06-24 scripts/structures/infer_structures_po.py

@@ -24,12 +24,11 @@ if args.filename.startswith("data"):
 else:
     filename = args.filename
 
-# Create name tag
-algo = "UNB"
-algo = algo + "0"
-
 outname = filename.replace(
-    ".matrix", "_%s_%02d_structure.txt" % (algo, args.seed))
+    ".matrix", "_PO_%02d_structure.txt" % (algo, args.seed))
+if os.path.exists(outname):
+    import sys
+    sys.exit(0)
 
 try:
     os.makedirs(os.path.dirname(filename))

### 2017-06-24 scripts/structures/launch_all_jobs.sh

@@ -3,21 +3,21 @@
 FOLDERS=(rings trophozoites schizonts)
 # FOLDERS=(HP1_KD HP1_tagged pv_sporozoite)
 for FOLDER in ${FOLDERS[@]}; do
-  sbatch --export=DATASET=${FOLDER} -v -p RM-shared --array 2-100 --mem 1500Mb \
+  sbatch --export=DATASET=${FOLDER} -v -p RM-shared --array 1-100 --mem 1500Mb \
     --job-name ${FOLDER}_MDS \
-    --time=8:00:00 cluster_scripts/slurm_mds.sh \
+    --time=48:00:00 cluster_scripts/slurm_mds.sh \
     -N 1 \
     --mail-type=ALL --mail-user=nelle@berkeley.edu;
 
-  sbatch --export=DATASET=${FOLDER} -v -p RM-shared --array 2-100 --mem 1500Mb \
+  sbatch --export=DATASET=${FOLDER} -v -p RM-shared --array 1-100 --mem 1500Mb \
     --job-name ${FOLDER}_NB0 \
-    --time=8:00:00 cluster_scripts/slurm_nb0.sh \
+    --time=48:00:00 cluster_scripts/slurm_nb0.sh \
     -N 1 \
     --mail-type=ALL --mail-user=nelle@berkeley.edu;
 
-  sbatch --export=DATASET=${FOLDER} -v -p RM-shared --array 2-100 --mem 1500Mb \
+  sbatch --export=DATASET=${FOLDER} -v -p RM-shared --array 1-100 --mem 1500Mb \
     --job-name ${FOLDER}_PO \
-    --time=8:00:00 cluster_scripts/slurm_po.sh \
+    --time=48:00:00 cluster_scripts/slurm_po.sh \
     -N 1 \
     --mail-type=ALL --mail-user=nelle@berkeley.edu;
 

### 2017-06-25 doc/manuscript/paper.tex
Activate inference using 0 contact counts with PO

@@
 Let us first introduce some notations. Hi-C data can be summarized as an $n
-\times n$ contact count matrix, where each row and column corresponds to a
-genomic window or loci, and each entry $c_{ij}$ refers to the number of time
-$i$ and $j$ have been seen interacting with one another.
+\times n$ symmetric contact count matrix, where each row and column corresponds to a
+genomic window (or loci), and each entry $c_{ij}$ refers to the number of time
+$i$ and $j$ have been seen interacting with one another. The matrix is hollow
 
 Hi-C data suffers from technical and biological biases that need to be
-corrected. We use ICE
+corrected. \fixme{blha}
 
@@
 
-A crucial step of our analysis of the \em{P. falciparum} genome relied on
+A crucial step of our analysis of the \textit{P. falciparum} genome relied on
 visual introspection of a consensus 3D model of the genome. Being able to
@@
 Briefly, we cast the problem of finding a consensus 3D structure $X \in R^{n
-\times n}$ as an optimization problem, where we attempted to place each beads
-in 3D so that they match as closely sa possible a predefined
-``wish-distance``. We added constraints to our model based on prior knowledge,
+\times n}$ as an optimization problem, either as minimizing a cost function,
+(often measuring the error between euclidean distance of $\mathbf{X}$ and a
+series of ``wish distances'') or maximizing a likelihood. The optimization
+problem can either be unconstrained or constrained to reflect some prior
+knowledge on the structure (size and shape of the nucleas, distances between
+loci, etc).
+
+ We added constraints to our model based on prior knowledge,
 such as the size of the nucleus, or how far apart two adjacent beads should
@@
 
-\begin{equation}
+\begin{equation}\label{eq:mds1} 
+\renewcommand{\arraystretch}{2} 
+\begin{array}{ccll} 
+\underset{\mathbf{X}}{\text{minimize}} & & 
+\underset{(i,j) \in \mathcal{D}}{\sum} w_{ij} \big(d_{ij}(\mathbf{X}) - \delta_{ij}\big)^2
+\,,&\\ 
+\end{array} 
 \end{equation}
+
+where $w_{ij}$ corresponds to weight.
 

### 2017-06-25 scripts/population_structures/clustering_across_stages_downsampled_distances.py

@@ -6,6 +6,7 @@ import matplotlib.pyplot as plt
 
 parser = argparse.ArgumentParser()
 parser.add_argument("--algorithm", "-a", default="MDS")
+parser.add_argument("--chromosome", "-c", default=None, type=int)
 args = parser.parse_args()
 
 datasets = ["rings", "schizonts", "trophozoites"]
@@ -14,21 +15,27 @@ algorithm = args.algorithm
 labels = None
 X = None
 for i, dataset in enumerate(datasets):
-    x = np.load(
-        "results/ay2013/%s_10000_raw_%s_distances_downsampled.npy" % (
-            dataset, algorithm))
+    if args.chromosome is None:
+        filename = (
+            "results/ay2013/%s_10000_raw_%s_distances_downsampled.npy" % (
+                dataset, algorithm))
+    else:
+        filename = (
+            "results/ay2013/%s_10000_raw_%s_distances_chr%02d.npy" % (
+                dataset, algorithm, args.chromosome))
+    x = np.load(filename)
     x = x[:100]
     if labels is not None:
         X[i, :len(x)] = x
         labels = np.concatenate([labels, i * np.ones(len(x))])
     else:
         labels = i * np.ones(len(x))
-        X = np.nan * np.zeros((len(datasets), x.shape[0], x.shape[1]))
+        X = np.nan * np.zeros((len(datasets), 1000, x.shape[1]))
         X[i, :len(x)] = x
 
 X = X.reshape(-1, x.shape[1])
 # Remove missing labels
-X = X[np.invert(np.isnan(X[:, 0]))]
+X = X[np.invert(np.all(np.isnan(X), axis=1))]
 # Remove any bead with missing value
 X = X[:, np.invert(np.isnan(X.sum(axis=0)))]
 
@@ -54,18 +61,18 @@ for i in range(len(datasets)):
         markerfacecolor=colors[i],
         markeredgecolor="white", linewidth=0, label=stages[i],
         markeredgewidth=0,
-        markersize=5,
+        markersize=10,
         zorder=10,
-        alpha=0.5)
+        alpha=0.9)
     ax2.plot(
         X[labels == i, 2], X[labels == i, 1],
         marker=".",
         markerfacecolor=colors[i], markeredgecolor="white",
         linewidth=0, label=stages[i],
         markeredgewidth=0,
-        markersize=5,
+        markersize=10,
         zorder=10,
-        alpha=0.5)
[…]

### 2017-06-25 scripts/population_structures/compute_distances.sh

@@ -6,6 +6,15 @@ python compute_downsampled_distances.py \
   structures/ay2013/trophozoites_10000_raw_UNB0 \
   --lengths data/ay2013/rings_10000_raw.bed
 
+python compute_pairwise_distances_chr.py \
+  structures/ay2013/trophozoites_10000_raw_MDS \
+  --lengths data/ay2013/rings_10000_raw.bed
+
+python compute_pairwise_distances_chr.py \
+  structures/ay2013/trophozoites_10000_raw_UNB0 \
+  --lengths data/ay2013/rings_10000_raw.bed
+
+
 python compute_downsampled_distances.py \
   structures/ay2013/schizonts_10000_raw_MDS \
   --lengths data/ay2013/rings_10000_raw.bed
@@ -14,11 +23,28 @@ python compute_downsampled_distances.py \
   structures/ay2013/schizonts_10000_raw_UNB0 \
   --lengths data/ay2013/rings_10000_raw.bed
 
+python compute_pairwise_distances_chr.py \
+  structures/ay2013/schizonts_10000_raw_MDS \
+  --lengths data/ay2013/rings_10000_raw.bed
+
+python compute_pairwise_distances_chr.py \
+  structures/ay2013/schizonts_10000_raw_UNB0 \
+  --lengths data/ay2013/rings_10000_raw.bed
+
+
 python compute_downsampled_distances.py \
   structures/ay2013/rings_10000_raw_MDS \
   --lengths data/ay2013/rings_10000_raw.bed
 
 python compute_downsampled_distances.py \
+  structures/ay2013/rings_10000_raw_UNB0 \
+  --lengths data/ay2013/rings_10000_raw.bed
+
+python compute_pairwise_distances_chr.py \
+  structures/ay2013/rings_10000_raw_MDS \
+  --lengths data/ay2013/rings_10000_raw.bed
+
+python compute_pairwise_distances_chr.py \
   structures/ay2013/rings_10000_raw_UNB0 \
   --lengths data/ay2013/rings_10000_raw.bed
 

### 2017-06-25 scripts/population_structures/compute_pairwise_distances_chr.py

@@ -9,15 +9,18 @@ from iced.io import load_lengths
 
 parser = argparse.ArgumentParser()
 parser.add_argument("directory")
+parser.add_argument("--lengths")
 args = parser.parse_args()
 
-filenames = glob(os.path.join(args.directory, "*NB0cst*_structure.txt"))
+if args.lengths is not None:
+    lengths = load_lengths(args.lengths)
+else:
+    lengths = load_lengths(os.path.join(
+        args.directory.replace("results", "data"),
+        "raw.bed"))
+
+filenames = glob(args.directory + "*_structure.txt")
 filenames.sort()
-
-lengths = load_lengths(os.path.join(
-    args.directory.replace("results", "data"),
-    "raw.bed"))
-
 distances = [[] for _ in lengths]
 
 for filename in filenames:
@@ -36,5 +39,10 @@ for filename in filenames:
 
 for l, distances_chr in enumerate(distances):
     distances_chr = np.concatenate(distances_chr)
-    np.save(os.path.join(args.directory, "distances_chr%02d.npy" % l),
+    outfile = (
+        args.directory.replace(
+            "structures", "results") + "_distances_chr%02d.npy" % (l+1))
+    print("Saving in %s" % outfile)
+
+    np.save(outfile,
             distances_chr)

### 2017-06-25 scripts/structures/infer_structures_po.py

@@ -99,6 +99,7 @@ X = poisson_structure.estimate_X(
     counts, alpha=-3., beta=1., bias=bias,
     verbose=1,
     maxiter=max_iter,
+    use_zero_entries=True,
     ini=X.flatten())
 
 counts = np.array(counts.todense())

### 2017-06-25 scripts/structures/getStructures.sh
Detailed outline for the paper

@@ -1,0 +1,1 @@
+rsync -avz nvaroqua@nexus.gs.washington.edu:tmp/takefive/ results/

### 2017-06-25 doc/manuscript/paper.tex
Computing PCA automatically, for all algo and all chr

@@
 % Title of paper
-\title{Exploration of empirical Bayes hierarchical modeling for the
-analysis of genome-wide association study data}
+\title{My very cool title}
 
@@
 \subsection{Analysing contact count data}
-- compartment analysis
-- extracting some relationships between counts genomic distances etc
-- Extracting significant interactions using fit-HiC
 
-Figure:
-  - log-linear relationships between counts and distances.
-  - significant interactions using fit-HiC ?
+- Normalization
+- Quality assessment
+- Enrichment of counts
+- Anything else?
+    - Did we do a compartment analysis?
+- Can we extend it?
+  - Fit-Hic ?
 
@@
 
- We added constraints to our model based on prior knowledge,
-such as the size of the nucleus, or how far apart two adjacent beads should
-lie, though we showed these constraints where well satisfied even when
-performing an unconstrained optimized. In this section, I will review the
-inference method we used in \citet{ay:three-dimensional} as well as two newer
-methods, based on well-grounded statistical method.
+We added constraints to our model based on prior knowledge, such as the size
+of the nucleus, or how far apart two adjacent beads should lie, though we
+showed these constraints where well satisfied even when performing an
+unconstrained optimized. In this section, I will review the inference method
+we used in \citet{ay:three-dimensional} as well as two newer methods, based on
+well-grounded statistical method.
 

### 2017-06-25 scripts/population_structures/clustering_across_stages_downsampled_distances.py

@@ -1,7 +1,7 @@
 import numpy as np
 import os
 import argparse
-from sklearn.decomposition import RandomizedPCA as PCA
+from sklearn.decomposition import PCA
 import matplotlib.pyplot as plt
 
 parser = argparse.ArgumentParser()
@@ -47,7 +47,7 @@ colors = ['#FFB773', '#FF7C00', "r",  # '#A65100',
 markers = [".", ".", ".", "D", "D", "D"]
 stages = ["Rin", "Sch", "Trop"]
 
-pca = PCA(n_components=3, copy=False, random_state=1)
+pca = PCA(n_components=3, copy=False, random_state=1, svd_solver="randomized")
 X = pca.fit_transform(X)
 fig, (ax1, ax2) = plt.subplots(
     ncols=2,

### 2017-06-25 scripts/population_structures/compute_pca.sh

@@ -1,0 +1,13 @@
+
+for chr in {1..14}; do
+  python clustering_across_stages_downsampled_distances.py \
+    -c $chr
+
+  python clustering_across_stages_downsampled_distances.py \
+    -c $chr -a UNB0
+done;
+
+python clustering_across_stages_downsampled_distances.py
+python clustering_across_stages_downsampled_distances.py -a UNB0
+
+

### 2017-06-25 scripts/volume_exclusion/plasmo_landmark.py
Volume exclusion models

@@ -1,0 +1,408 @@
+"""Nuclear Landmark model for Plasmodium"""
+import os
+import sys
+import math
+import IMP
+import IMP.core
+import IMP.atom
+import IMP.display
+import IMP.algebra
+import numpy as np
+import time
+import argparse
+
+state = "RINGS"
+
+parser = argparse.ArgumentParser(description='Run volume exclusion model')
+parser.add_argument('seed', metavar='N', type=int,
+                    help='seed')
+args = parser.parse_args()
+
+# Set the seed and the random state
+random_state = np.random.RandomState(seed=args.seed)
+
+# Nuclear radius - this should change depending on the state of the plasmodium
+# Let's work with rings right now
+nuclear_rad = 350
+
+basedir = os.path.dirname(os.path.realpath(__file__))
+outname = os.path.join(basedir, "cluster", str(args.seed),
+                       '%s_plasmodium' % state)
+print 'output pdb:', outname
+if os.path.exists(outname):
+    sys.exit(0)
+
+try:
+    os.makedirs(os.path.dirname(outname))
+except OSError:
+    pass
+
+chr_seq = {"chr1": 640851, "chr2": 947102, "chr3": 1067971, "chr4": 1200490,
+           "chr5": 1343557, "chr6": 1418242, "chr7": 1445207, "chr8": 1472805,
+           "chr9": 1541735, "chr10": 1687656, "chr11": 2038340,
+           "chr12": 2271494,
+           "chr13": 2925236, "chr14": 3291936}
+
+# Centromers middle position
+chr_cen = {"chr1": 459191, "chr2": 448856, "chr3": 599144, "chr4": 643179,
+           "chr5": 456668, "chr6": 479886, "chr7": 810620, "chr8": 300205,
+           "chr9": 1243266, "chr10": 936752, "chr11": 833107,
+           "chr12": 1283731,
+           "chr13": 1169395, "chr14": 1073139}
+
+# Var genes mean position
+var_genes = {
+    "chr1": [62704, 555068],
+    "chr2": [46388, 892057],
+    "chr3": [52823, 1020563],
+    "chr4": [61257.5, 131090, 159800, 580028,
+             961239, 1140018],
+    "chr5": [31166, 1332682],
+    "chr6": [26775.5,
[…]

### 2017-06-25 scripts/volume_exclusion/plasmo_landmark_without_VRSM.py

@@ -1,0 +1,377 @@
+"""Nuclear Landmark model for Plasmodium"""
+import os
+import sys
+import math
+import IMP
+import IMP.core
+import IMP.atom
+import IMP.display
+import IMP.algebra
+import numpy as np
+import time
+import argparse
+
+state = "RINGS"
+
+parser = argparse.ArgumentParser(description='Run volume exclusion model')
+parser.add_argument('seed', metavar='N', type=int,
+                    help='seed')
+args = parser.parse_args()
+
+# Set the seed and the random state
+random_state = np.random.RandomState(seed=args.seed)
+
+# Nuclear radius - this should change depending on the state of the plasmodium
+# Let's work with rings right now
+nuclear_rad = 350
+
+basedir = os.path.dirname(os.path.realpath(__file__))
+outname = os.path.join(basedir, "cluster", str(args.seed),
+                       '%s_plasmodium_without_VRSM' % state)
+print 'output pdb:', outname
+if os.path.exists(outname):
+    sys.exit(0)
+
+try:
+    os.path.makedirs(os.path.dirname(outname))
+except OSError:
+    pass
+
+chr_seq = {"chr1": 640851, "chr2": 947102, "chr3": 1067971, "chr4": 1200490,
+           "chr5": 1343557, "chr6": 1418242, "chr7": 1445207, "chr8": 1472805,
+           "chr9": 1541735, "chr10": 1687656, "chr11": 2038340,
+           "chr12": 2271494,
+           "chr13": 2925236, "chr14": 3291936}
+
+# Centromers middle position
+chr_cen = {"chr1": 459191, "chr2": 448856, "chr3": 599144, "chr4": 643179,
+           "chr5": 456668, "chr6": 479886, "chr7": 810620, "chr8": 300205,
+           "chr9": 1243266, "chr10": 936752, "chr11": 833107,
+           "chr12": 1283731,
+           "chr13": 1169395, "chr14": 1073139}
+
+# Var genes mean position
+var_genes = {
+    "chr1": [62704, 555068],
+    "chr2": [46388, 892057],
+    "chr3": [52823, 1020563],
+    "chr4": [61257.5, 131090, 159800, 580028,
+             961239, 1140018],
+    "chr5": [31166, 1332682],
+    "chr6": [26775.5,
[…]

### 2017-06-26 scripts/volume_exclusion/plasmo_udpate.py
Updated VE model to use new version of IMP

@@ -1,0 +1,353 @@
+"""Nuclear Landmark model for Plasmodium"""
+import os
+import sys
+import math
+import IMP
+import IMP.core
+import IMP.atom
+import IMP.display
+import IMP.algebra
+import numpy as np
+import time
+import argparse
+
+state = "RINGS"
+
+parser = argparse.ArgumentParser(description='Run volume exclusion model')
+parser.add_argument('seed', metavar='N', type=int,
+                    help='seed')
+args = parser.parse_args()
+
+# Set the seed and the random state
+random_state = np.random.RandomState(seed=args.seed)
+
+# Nuclear radius - this should change depending on the state of the plasmodium
+# Let's work with rings right now
+nuclear_rad = 350
+
+basedir = os.path.dirname(os.path.realpath(__file__))
+outname = os.path.join(basedir, "cluster", str(args.seed),
+                       '%s_plasmodium_without_VRSM' % state)
+print('output pdb:', outname)
+if os.path.exists(outname):
+    sys.exit(0)
+
+try:
+    os.makedirs(os.path.dirname(outname))
+except OSError:
+    pass
+
+chr_seq = {"chr1": 640851, "chr2": 947102, "chr3": 1067971, "chr4": 1200490,
+           "chr5": 1343557, "chr6": 1418242, "chr7": 1445207, "chr8": 1472805,
+           "chr9": 1541735, "chr10": 1687656, "chr11": 2038340,
+           "chr12": 2271494,
+           "chr13": 2925236, "chr14": 3291936}
+
+# Centromers middle position
+chr_cen = {"chr1": 459191, "chr2": 448856, "chr3": 599144, "chr4": 643179,
+           "chr5": 456668, "chr6": 479886, "chr7": 810620, "chr8": 300205,
+           "chr9": 1243266, "chr10": 936752, "chr11": 833107,
+           "chr12": 1283731,
+           "chr13": 1169395, "chr14": 1073139}
+
+# Var genes mean position
+var_genes = {
+    "chr1": [62704, 555068],
+    "chr2": [46388, 892057],
+    "chr3": [52823, 1020563],
+    "chr4": [61257.5, 131090, 159800, 580028,
+             961239, 1140018],
+    "chr5": [31166, 1332682],
+    "chr6": [26775.5,
[…]

### 2017-06-26 scripts/volume_exclusion/plasmo_landmark_without_VRSM.py
FIX VE model

@@ -33,7 +33,7 @@ if os.path.exists(outname):
     sys.exit(0)
 
 try:
-    os.path.makedirs(os.path.dirname(outname))
+    os.makedirs(os.path.dirname(outname))
 except OSError:
     pass
 

### 2017-06-26 scripts/volume_exclusion/plasmo_udpate.py

@@ -292,7 +292,7 @@ constraints = [evr, br, rcell, rcentro, rt]
 mdstep(m, constraints, 500000, 5000)
 mdstep(m, constraints, 300000, 5000)
 mdstep(m, constraints, 5000, 10000)
-score = cgstep(n, constraints, 500)
+score = cgstep(m, constraints, 500)
 print 'before angle', score
 
 

### 2017-06-26 scripts/volume_exclusion/plasmo_udpate.py
easy switch between states

@@ -16,14 +16,23 @@ state = "RINGS"
 parser = argparse.ArgumentParser(description='Run volume exclusion model')
 parser.add_argument('seed', metavar='N', type=int,
                     help='seed')
+parser.add_argument("--state", default="rings")
 args = parser.parse_args()
 
 # Set the seed and the random state
+state  = args.state.lower()
 random_state = np.random.RandomState(seed=args.seed)
 
 # Nuclear radius - this should change depending on the state of the plasmodium
 # Let's work with rings right now
-nuclear_rad = 350
+if state == "rings" or state == "r":
+    nuclear_rad = 350
+elif state in ["schizont", "schizonts", "s"]:
+    nuclear_rad = 450
+elif state in ["trophozoites", "trophozoite", "t", "troph"]:
+    nuclear_rad = 850
+else:
+    raise ValueError("Unknown state %s" % state)
 
 basedir = os.path.dirname(os.path.realpath(__file__))
 outname = os.path.join(basedir, "cluster", str(args.seed),

### 2017-06-26 scripts/volume_exclusion/plasmo_udpate.py
Clean ups

@@ -10,6 +10,7 @@ import IMP.algebra
 import numpy as np
 import time
 import argparse
+from future import print_function
 
 state = "RINGS"
 
@@ -20,7 +21,7 @@ parser.add_argument("--state", default="rings")
 args = parser.parse_args()
 
 # Set the seed and the random state
-state  = args.state.lower()
+state = args.state.lower()
 random_state = np.random.RandomState(seed=args.seed)
 
 # Nuclear radius - this should change depending on the state of the plasmodium
@@ -35,9 +36,10 @@ else:
     raise ValueError("Unknown state %s" % state)
 
 basedir = os.path.dirname(os.path.realpath(__file__))
-outname = os.path.join(basedir, "cluster", str(args.seed),
-                       '%s_plasmodium_without_VRSM' % state)
+outname = os.path.join(basedir, "results",
+                       '%s_plasmodium_without_VRSM_%004d.pdb' % (state, args.seed))
 print('output pdb:', outname)
+
 if os.path.exists(outname):
     sys.exit(0)
 
@@ -153,11 +155,12 @@ def pdboutput(name=None):
         X.append([int(chr_num[3:]), p0.get_x(), p0.get_y(), p0.get_z()])
     return X
 
+
 def mdstep(model, constraints, t, step):
     sf = IMP.core.RestraintsScoringFunction(constraints)
     o = IMP.atom.MolecularDynamics(model)
     # replace 300 K with 500 K
-    #md = IMP.atom.VelocityScalingOptimizerState(xyzr, t, 10)
+    # md = IMP.atom.VelocityScalingOptimizerState(xyzr, t, 10)
     md = IMP.atom.VelocityScalingOptimizerState(model, xyzr, t)
     o.set_scoring_function(sf)
     o.add_optimizer_state(md)
@@ -226,11 +229,11 @@ for id in chr_seq.keys():
 # Restraint for bonds
 bss = IMP.atom.BondSingletonScore(IMP.core.Harmonic(0, 1))
 br = IMP.container.SingletonsRestraint(bss, bonds)
-#m.add_restraint(br)  # 0
+# m.add_restraint(br)  # 0
 
 # Set up excluded volume
 evr = IMP.core.ExcludedVolumeRestraint(chain)
-#m.add_restraint(evr)  # 1
[…]

### 2017-06-26 scripts/structures/launch_all_jobs.sh
Update scripts for cluster

@@ -3,19 +3,19 @@
 FOLDERS=(rings trophozoites schizonts)
 # FOLDERS=(HP1_KD HP1_tagged pv_sporozoite)
 for FOLDER in ${FOLDERS[@]}; do
-  sbatch --export=DATASET=${FOLDER} -v -p RM-shared --array 1-100 --mem 1500Mb \
-    --job-name ${FOLDER}_MDS \
-    --time=48:00:00 cluster_scripts/slurm_mds.sh \
-    -N 1 \
-    --mail-type=ALL --mail-user=nelle@berkeley.edu;
+#  sbatch --export=DATASET=${FOLDER} -v -p RM-shared --array 1-100 --mem 1500Mb \
+#    --job-name ${FOLDER}_MDS \
+#    --time=48:00:00 cluster_scripts/slurm_mds.sh \
+#    -N 1 \
+#    --mail-type=ALL --mail-user=nelle@berkeley.edu;
+#
+#  sbatch --export=DATASET=${FOLDER} -v -p RM-shared --array 1-100 --mem 1500Mb \
+#    --job-name ${FOLDER}_NB0 \
+#    --time=48:00:00 cluster_scripts/slurm_nb0.sh \
+#    -N 1 \
+#    --mail-type=ALL --mail-user=nelle@berkeley.edu;
 
-  sbatch --export=DATASET=${FOLDER} -v -p RM-shared --array 1-100 --mem 1500Mb \
-    --job-name ${FOLDER}_NB0 \
-    --time=48:00:00 cluster_scripts/slurm_nb0.sh \
-    -N 1 \
-    --mail-type=ALL --mail-user=nelle@berkeley.edu;
-
-  sbatch --export=DATASET=${FOLDER} -v -p RM-shared --array 1-100 --mem 1500Mb \
+  sbatch --export=DATASET=${FOLDER} -v -p RM-shared --array 101-1000 --mem 1500Mb \
     --job-name ${FOLDER}_PO \
     --time=48:00:00 cluster_scripts/slurm_po.sh \
     -N 1 \

### 2017-06-26 scripts/volume_exclusion/cluster_scripts/slurm_ve.sh

@@ -1,0 +1,6 @@
+#! /bin/sh
+
+source activate IMP
+cd $SLURM_SUBMIT_DIR
+
+python plasmo_udpate.py ${SLURM_ARRAY_TASK_ID} --state ${STATE}

### 2017-06-26 scripts/volume_exclusion/launch_all_jobs.sh

@@ -1,0 +1,14 @@
+#!/bin/bash
+
+STATES=(rings trophozoites schizonts)
+# FOLDERS=(HP1_KD HP1_tagged pv_sporozoite)
+for STATE in ${STATES[@]}; do
+
+  sbatch --export=STATE=${STATE} -v -p RM-shared --array 1-2 --mem 1500Mb \
+    --job-name ${STATE}_VE \
+    --time=4:00:00 cluster_scripts/slurm_ve.sh \
+    -N 1 \
+    --mail-type=ALL --mail-user=nelle@berkeley.edu;
+done;
+
+

### 2017-06-26 scripts/volume_exclusion/plasmo_landmark.py
Updated VE models to use VRSM genes or not

@@ -10,25 +10,34 @@ import IMP.algebra
 import numpy as np
 import time
 import argparse
-
-state = "RINGS"
+from future import print_function
 
 parser = argparse.ArgumentParser(description='Run volume exclusion model')
 parser.add_argument('seed', metavar='N', type=int,
                     help='seed')
+parser.add_argument("--state", default="rings")
 args = parser.parse_args()
 
 # Set the seed and the random state
+state = args.state.lower()
 random_state = np.random.RandomState(seed=args.seed)
 
 # Nuclear radius - this should change depending on the state of the plasmodium
 # Let's work with rings right now
-nuclear_rad = 350
+if state == "rings" or state == "r":
+    nuclear_rad = 350
+elif state in ["schizont", "schizonts", "s"]:
+    nuclear_rad = 450
+elif state in ["trophozoites", "trophozoite", "t", "troph"]:
+    nuclear_rad = 850
+else:
+    raise ValueError("Unknown state %s" % state)
 
 basedir = os.path.dirname(os.path.realpath(__file__))
-outname = os.path.join(basedir, "cluster", str(args.seed),
-                       '%s_plasmodium' % state)
-print 'output pdb:', outname
+outname = os.path.join(basedir, "results",
+                       '%s_plasmodium_without_VRSM_%004d.pdb' % (state, args.seed))
+print('output pdb:', outname)
+
 if os.path.exists(outname):
     sys.exit(0)
 
@@ -136,59 +145,22 @@ def find_bead_in_chr(bid):
     return i, order, genpos
 
 
-def pdboutput(name):
-    pdb = [[] for i in range(nbead)]
-    # -----------------------------
-    cen_pos = []
-    for k in chr_seq.keys():
-        j = bead_id(k, chr_cen[k])
-        cen_pos.append(j)
-    # ------------------------------------
+def pdboutput(name=None):
+    X = []
     for i in range(nbead):
         p0 = IMP.core.XYZR(chain.get_particle(i))
-        chr = find_chromosome(i)
-        # pdb[i].append('ATOM')
-        if i in cen_pos:
[…]

### 2017-06-26 scripts/volume_exclusion/plasmo_landmark_without_VRSM.py

@@ -10,25 +10,34 @@ import IMP.algebra
 import numpy as np
 import time
 import argparse
-
-state = "RINGS"
+from future import print_function
 
 parser = argparse.ArgumentParser(description='Run volume exclusion model')
 parser.add_argument('seed', metavar='N', type=int,
                     help='seed')
+parser.add_argument("--state", default="rings")
 args = parser.parse_args()
 
 # Set the seed and the random state
+state = args.state.lower()
 random_state = np.random.RandomState(seed=args.seed)
 
 # Nuclear radius - this should change depending on the state of the plasmodium
 # Let's work with rings right now
-nuclear_rad = 350
+if state == "rings" or state == "r":
+    nuclear_rad = 350
+elif state in ["schizont", "schizonts", "s"]:
+    nuclear_rad = 450
+elif state in ["trophozoites", "trophozoite", "t", "troph"]:
+    nuclear_rad = 850
+else:
+    raise ValueError("Unknown state %s" % state)
 
 basedir = os.path.dirname(os.path.realpath(__file__))
-outname = os.path.join(basedir, "cluster", str(args.seed),
-                       '%s_plasmodium_without_VRSM' % state)
-print 'output pdb:', outname
+outname = os.path.join(basedir, "results",
+                       '%s_plasmodium_without_VRSM_%004d.pdb' % (state, args.seed))
+print('output pdb:', outname)
+
 if os.path.exists(outname):
     sys.exit(0)
 
@@ -136,59 +145,22 @@ def find_bead_in_chr(bid):
     return i, order, genpos
 
 
-def pdboutput(name):
-    pdb = [[] for i in range(nbead)]
-    # -----------------------------
-    cen_pos = []
-    for k in chr_seq.keys():
-        j = bead_id(k, chr_cen[k])
-        cen_pos.append(j)
-    # ------------------------------------
+def pdboutput(name=None):
+    X = []
     for i in range(nbead):
         p0 = IMP.core.XYZR(chain.get_particle(i))
-        chr = find_chromosome(i)
-        # pdb[i].append('ATOM')
-        if i in cen_pos:
[…]

### 2017-06-26 scripts/volume_exclusion/cluster_scripts/slurm_ve.sh
Some cleaning up

@@ -3,4 +3,4 @@
 source activate IMP
 cd $SLURM_SUBMIT_DIR
 
-python plasmo_udpate.py ${SLURM_ARRAY_TASK_ID} --state ${STATE}
+python plasmo_landmark_without_VRSM.py ${SLURM_ARRAY_TASK_ID} --state ${STATE}

### 2017-06-26 scripts/volume_exclusion/launch_all_jobs.sh

@@ -4,7 +4,7 @@ STATES=(rings trophozoites schizonts)
 # FOLDERS=(HP1_KD HP1_tagged pv_sporozoite)
 for STATE in ${STATES[@]}; do
 
-  sbatch --export=STATE=${STATE} -v -p RM-shared --array 1-2 --mem 1500Mb \
+  sbatch --export=STATE=${STATE} -v -p RM-shared --array 1-100 --mem 1500Mb \
     --job-name ${STATE}_VE \
     --time=4:00:00 cluster_scripts/slurm_ve.sh \
     -N 1 \

### 2017-06-26 scripts/volume_exclusion/plasmo_landmark.py

@@ -1,4 +1,5 @@
 """Nuclear Landmark model for Plasmodium"""
+
 import os
 import sys
 import math
@@ -34,8 +35,8 @@ else:
     raise ValueError("Unknown state %s" % state)
 
 basedir = os.path.dirname(os.path.realpath(__file__))
-outname = os.path.join(basedir, "results",
-                       '%s_plasmodium_without_VRSM_%004d.pdb' % (state, args.seed))
+outname = os.path.join(basedir, "results", "with_VRMS",
+                       '%s_plasmodium_%004d.pdb' % (state, args.seed))
 print('output pdb:', outname)
 
 if os.path.exists(outname):

### 2017-06-26 scripts/volume_exclusion/plasmo_landmark_without_VRSM.py

@@ -34,8 +34,8 @@ else:
     raise ValueError("Unknown state %s" % state)
 
 basedir = os.path.dirname(os.path.realpath(__file__))
-outname = os.path.join(basedir, "results",
-                       '%s_plasmodium_without_VRSM_%004d.pdb' % (state, args.seed))
+outname = os.path.join(basedir, "results", "without_VRSM",
+                       '%s_plasmodium_%004d.pdb' % (state, args.seed))
 print('output pdb:', outname)
 
 if os.path.exists(outname):

### 2017-06-26 scripts/volume_exclusion/plasmo_landmark.py
More cleaning up

@@ -1,3 +1,5 @@
+from __future__ import print_function
+
 """Nuclear Landmark model for Plasmodium"""
 
 import os
@@ -11,12 +13,11 @@ import IMP.algebra
 import numpy as np
 import time
 import argparse
-from future import print_function
 
 parser = argparse.ArgumentParser(description='Run volume exclusion model')
 parser.add_argument('seed', metavar='N', type=int,
                     help='seed')
-parser.add_argument("--state", default="rings")
+parser.add_argument("--state", "-s", default="rings", type=str)
 args = parser.parse_args()
 
 # Set the seed and the random state

### 2017-06-26 scripts/volume_exclusion/plasmo_landmark_without_VRSM.py

@@ -1,4 +1,7 @@
+from __future__ import print_function
+
 """Nuclear Landmark model for Plasmodium"""
+
 import os
 import sys
 import math
@@ -10,12 +13,11 @@ import IMP.algebra
 import numpy as np
 import time
 import argparse
-from future import print_function
 
 parser = argparse.ArgumentParser(description='Run volume exclusion model')
 parser.add_argument('seed', metavar='N', type=int,
                     help='seed')
-parser.add_argument("--state", default="rings")
+parser.add_argument("--state", "-s", default="rings")
 args = parser.parse_args()
 
 # Set the seed and the random state

### 2017-06-26 scripts/volume_exclusion/plasmo_landmark_without_VRSM.py
more cleaning up

@@ -348,6 +348,7 @@ score = cgstep(m, constraints, 500)
 X = pdboutput(outname)
 X = np.array(X)
 np.savetxt(outname, X)
+np.savetxt(outname + ".score", score)
 
 # -------------------------
 

### 2017-06-26 scripts/structures/getStructures.sh
Tiny nitpicks

@@ -1,1 +1,1 @@
-rsync -avz nvaroqua@nexus.gs.washington.edu:tmp/takefive/ results/
+rsync -avz nvaroqua@nexus.gs.washington.edu:tmp/takefive/structures/ results/

### 2017-06-26 scripts/volume_exclusion/getStructures.sh

@@ -1,0 +1,1 @@
+rsync -avz nvaroqua@nexus.gs.washington.edu:tmp/takefive/VE/ results/

### 2017-06-26 scripts/volume_exclusion/plasmo_landmark.py

@@ -361,7 +361,7 @@ for k in var_genes.keys():
     for pos in var_genes[k]:
         j = bead_id(k, pos)
         pvar = chain.get_particle(j)
-        listvargenes.add_particle(pvar)
+        listvargenes.add(pvar)
 
 var_genes_center = random_state.randint(
     -nuclear_rad + 50,  nuclear_rad - 50, 3)
@@ -371,7 +371,7 @@ varcen = IMP.core.HarmonicUpperBound(var_genes_rad, 1.)
 ssvarcen = IMP.core.DistanceToSingletonScore(varcen, var_genes_center)
 rvar = IMP.container.SingletonsRestraint(ssvarcen, listvargenes)
 
-constraints += rvar
+constraints += [rvar]
 
 print('High temp MD..')
 mdstep(m, constraints, 1000000, 500)
@@ -384,15 +384,16 @@ score = cgstep(m, constraints, 1000)
 
 print('angle: %.1f ' % (score))
 # -----------------------
-for i in angle_set:
-    m.remove_restraint(i)
-score = cgstep(1000)
+
+constraints = [evr, br, rcell, rcentro, rt, rvar]
+
+score = cgstep(m, constraints, 1000)
 print('Final score:%.1f' % (score))
 
 X = pdboutput(outname)
 X = np.array(X)
 np.savetxt(outname, X)
-np.savetxt(outname + ".score", score)
+np.savetxt(outname + ".score", np.array([score]))
 
 
 # -------------------------

### 2017-06-26 scripts/volume_exclusion/plasmo_landmark.py
Reordering chromosomes before saving.

@@ -37,7 +37,7 @@ else:
 
 basedir = os.path.dirname(os.path.realpath(__file__))
 outname = os.path.join(basedir, "results", "with_VRMS",
-                       '%s_plasmodium_%004d.pdb' % (state, args.seed))
+                       '%s_plasmodium_%004d.txt' % (state, args.seed))
 print('output pdb:', outname)
 
 if os.path.exists(outname):
@@ -48,32 +48,34 @@ try:
 except OSError:
     pass
 
-chr_seq = {"chr1": 640851, "chr2": 947102, "chr3": 1067971, "chr4": 1200490,
-           "chr5": 1343557, "chr6": 1418242, "chr7": 1445207, "chr8": 1472805,
-           "chr9": 1541735, "chr10": 1687656, "chr11": 2038340,
+chr_seq = {"chr01": 640851, "chr02": 947102,
+           "chr03": 1067971, "chr04": 1200490,
+           "chr05": 1343557, "chr06": 1418242,
+           "chr07": 1445207, "chr08": 1472805,
+           "chr09": 1541735, "chr10": 1687656, "chr11": 2038340,
            "chr12": 2271494,
            "chr13": 2925236, "chr14": 3291936}
 
 # Centromers middle position
-chr_cen = {"chr1": 459191, "chr2": 448856, "chr3": 599144, "chr4": 643179,
-           "chr5": 456668, "chr6": 479886, "chr7": 810620, "chr8": 300205,
-           "chr9": 1243266, "chr10": 936752, "chr11": 833107,
+chr_cen = {"chr01": 459191, "chr02": 448856, "chr03": 599144, "chr04": 643179,
+           "chr05": 456668, "chr06": 479886, "chr07": 810620, "chr08": 300205,
+           "chr09": 1243266, "chr10": 936752, "chr11": 833107,
            "chr12": 1283731,
            "chr13": 1169395, "chr14": 1073139}
 
 # Var genes mean position
 var_genes = {
-    "chr1": [62704, 555068],
-    "chr2": [46388, 892057],
-    "chr3": [52823, 1020563],
-    "chr4": [61257.5, 131090, 159800, 580028,
-             961239, 1140018],
-    "chr5": [31166, 1332682],
-    "chr6": [26775.5,
-             732761.5, 1348167.5],
-    "chr7": [48681, 561236,  1404322.],
[…]

### 2017-06-26 scripts/volume_exclusion/plasmo_landmark_without_VRSM.py

@@ -37,7 +37,7 @@ else:
 
 basedir = os.path.dirname(os.path.realpath(__file__))
 outname = os.path.join(basedir, "results", "without_VRSM",
-                       '%s_plasmodium_%004d.pdb' % (state, args.seed))
+                       '%s_plasmodium_%004d.txt' % (state, args.seed))
 print('output pdb:', outname)
 
 if os.path.exists(outname):
@@ -48,49 +48,49 @@ try:
 except OSError:
     pass
 
-chr_seq = {"chr1": 640851, "chr2": 947102, "chr3": 1067971, "chr4": 1200490,
-           "chr5": 1343557, "chr6": 1418242, "chr7": 1445207, "chr8": 1472805,
-           "chr9": 1541735, "chr10": 1687656, "chr11": 2038340,
-           "chr12": 2271494,
-           "chr13": 2925236, "chr14": 3291936}
+chr_seq = {"chr01": 640851, "chr02": 947102, "chr03": 1067971, "chr04": 1200490,
+           "chr05": 1343557, "chr06": 1418242, "chr07": 1445207, "chr08": 1472805,
+           "chr09": 1541735, "chr010": 1687656, "chr011": 2038340,
+           "chr012": 2271494,
+           "chr013": 2925236, "chr014": 3291936}
 
 # Centromers middle position
-chr_cen = {"chr1": 459191, "chr2": 448856, "chr3": 599144, "chr4": 643179,
-           "chr5": 456668, "chr6": 479886, "chr7": 810620, "chr8": 300205,
-           "chr9": 1243266, "chr10": 936752, "chr11": 833107,
-           "chr12": 1283731,
-           "chr13": 1169395, "chr14": 1073139}
+chr_cen = {"chr01": 459191, "chr02": 448856, "chr03": 599144, "chr04": 643179,
+           "chr05": 456668, "chr06": 479886, "chr07": 810620, "chr08": 300205,
+           "chr09": 1243266, "chr010": 936752, "chr011": 833107,
+           "chr012": 1283731,
+           "chr013": 1169395, "chr014": 1073139}
 
 # Var genes mean position
 var_genes = {
-    "chr1": [62704, 555068],
-    "chr2": [46388, 892057],
-    "chr3": [52823, 1020563],
-    "chr4": [61257.5, 131090, 159800, 580028,
[…]

### 2017-07-03 doc/manuscript/paper.tex
MDS approach

@@
 % Title of paper
-\title{My very cool title}
+\title{Analysing the three-dimensional structure of \em{P. falciparum}}
 
@@
 % E-mail address for correspondence
-{eaheron@tcd.ie}}
+{nelle@berkeley.edu}}
 
@@
 
-Let us first introduce some notations. Hi-C data can be summarized as an $n
-\times n$ symmetric contact count matrix, where each row and column corresponds to a
-genomic window (or loci), and each entry $c_{ij}$ refers to the number of time
-$i$ and $j$ have been seen interacting with one another. The matrix is hollow
-
-Hi-C data suffers from technical and biological biases that need to be
-corrected. \fixme{blha}
-
 \subsection{Analysing contact count data}
@@
 
-\subsection{3D structure inference}
+\subsection{Inferring the 3D structures}
 
-A crucial step of our analysis of the \textit{P. falciparum} genome relied on
-visual introspection of a consensus 3D model of the genome. Being able to
-quickly visualize how chromosome folded, where families of genes lied in the
-3D space or whether regions where enriched helped us integrate with other
-sources of data.
+An important part of analysing the 3D structure of the genome of \textit{P.
+falciparum} relied on having consensus models for each stage of the life
+cycle. The method presented in \citet{ay:three-dimensional} relied on an
+approach called ``multidimensional scaling'' (MDS), widely used in the
+literature for the purpose of 3D structure inference
+\citep{duan:three-dimensional, varoquaux:statistical, bau:, tanizawa:}. Each chromosomes is
+modeled a series of beads, spaced ~10~kb apart. MDS-based method consists of
+two steps: first convert the
+contact counts $c_{ij}$ into ``wish-distances'' $\delta_{ij}$; then attempt
+to place each bead in 3D such that the euclidean distances $d_{ij}$ separating
+beads $i$ and $j$ matches as closely as possible the wish distance
+$\delta_{ij}$.
 
-Briefly, we cast the problem of finding a consensus 3D structure $X \in R^{n
-\times n}$ as an optimization problem, either as minimizing a cost function,
-(often measuring the error between euclidean distance of $\mathbf{X}$ and a
-series of ``wish distances'') or maximizing a likelihood. The optimization
-problem can either be unconstrained or constrained to reflect some prior
-knowledge on the structure (size and shape of the nucleas, distances between
-loci, etc).
+While these approaches are widely used, since better motivated and more
+flexible models to infer 3D consensus structures from contact counts have been
+proposed.
 
-We added constraints to our model based on prior knowledge, such as the size
-of the nucleus, or how far apart two adjacent beads should lie, though we
-showed these constraints where well satisfied even when performing an
-unconstrained optimized. In this section, I will review the inference method
-we used in \citet{ay:three-dimensional} as well as two newer methods, based on
-well-grounded statistical method.
+\subsubsection{Constrained multi-dimensional scaling}
 
-\subsubsection{Multidimensional scaling}
+The method presented in \citet{ay:three-dimensional} relied on an
+approach called ``multidimensional scaling'' (MDS), widely used in the
+literature for the purpose of 3D structure inference
+\citep{duan:three-dimensional, varoquaux:statistical, bau:, tanizawa:}. Each chromosomes is
+modeled a series of beads, spaced ~10~kb apart. MDS-based method consists of
+two steps: first convert the
+contact counts $c_{ij}$ into ``wish-distances'' $\delta_{ij}$; then attempt
+to place each bead in 3D such that the euclidean distances $d_{ij}$ separating
+beads $i$ and $j$ matches as closely as possible the wish distances
+$\delta_{ij}$.
+
+\paragraph{Wish distances}
+
+A crucial step of MDS-based method rely on finding a good set of
+``wish-distances'' from Hi-C contact count data.
+
+\paragraph{Optimization}
+
+Given the resulting wish-distances, we cast the inference of the consensus
+structure $\mathbf{X}^{3 \times n}$ as the following constrained weighted MDS:
 
@@
 \underset{\mathbf{X}}{\text{minimize}} & & 
-\underset{(i,j) \in \mathcal{D}}{\sum} w_{ij} \big(d_{ij}(\mathbf{X}) - \delta_{ij}\big)^2
-\,,&\\ 
+\underset{(i,j) \in \mathcal{D}}{\sum} \frac{1}{\delta^2_{ij}}
+\big(d_{ij}(\mathbf{X}) - \delta_{ij}\big)^2 & \\
+\text{subject to} & & d_{ij} \leq r^2_\text{max} \quad \forall i, j &\,,\\
+& & d_{i, i+1} \leq b_\text{max} \quad i = \{i:n|\text{chr}_i = \text{chr}_{i+1}\,,&\\ 
 \end{array} 
@@
 
-where $w_{ij}$ corresponds to weight.
+The various constraints corresponds to {\em prior} knowledge on the 3D
+structure, and vary across organisms, time-points of the life cycle and
+cell-type. For {\em P. falciparum}, the constraints are as follow.
+
+\begin{itemize}
+\item All beads lie in a spherical nucleus centered on the origin and of a
+certain size.
+\item Two adjacent beads must not be too far apart. Leveraging the work of
+\citet{berger}, we assume that $b^\text{max} = 91~nm$
+\end{itemize}
+
+Note that in practice, the constraints are mostly intrinsic to the data and
+thus can be dropped.
 
@@
 -
+
 \subsubsection{Volume exclusion models}
@@
 \label{sec3}
-
-\subsection{Simulation study application}
 

### 2017-07-03 doc/manuscript/paper.tex
Wish distances

@@
 A crucial step of MDS-based method rely on finding a good set of
-``wish-distances'' from Hi-C contact count data.
+``wish-distances'' from Hi-C contact count data. While several approach have
+been used in the litterature, most leverage bio-physics relationships either
+between the genomic distnaces $s$ and the expected count, or the relationship
+between the euclidean distance $d$ and the expected
+count. \citet{ay:three-dimensional} observe that {\em P. falciparum}'s DNA
+behave like a fractal globule polymer. \citet{grosberg:fractal} show fractal
+globul polymer exhibits a well-defined relationship between its genomic length
+and physical distance $d$:
+
+$$d\sim s^{1/3}$$
+
+The relationship between genomic distance $s$ and $c$ can be derived from the
+data, by fitting a linear model between the log expected count and the genomic
+distances.
 
@@
 
-Note that in practice, the constraints are mostly intrinsic to the data and
-thus can be dropped.
+Note that in practice, the constraints are fulfilled naturally by the data,
+and thus can be dropped.
 
 \subsubsection{Modeling contact counts as a Poisson distribution}
-
-\subsubsection{Modeling contact counts as a Negative Binomial distribution}
 
@@
 \subsection{Integrative analysis of gene expression and 3D structure using KernelCCA}
+
+In order to integrate
 

### 2017-07-03 doc/manuscript/paper.tex
More on publicly available softwares
VE models

@@
 \item Two adjacent beads must not be too far apart. Leveraging the work of
-\citet{berger}, we assume that $b^\text{max} = 91~nm$
+\citet{berger:high TOVERIFY}, we assume that $b^\text{max} = 91~nm$
 \end{itemize}
@@
 Note that in practice, the constraints are fulfilled naturally by the data,
-and thus can be dropped.
+and thus can be dropped. In addition of drastically reducing the computation
+time, this allows to use generic publicly available software such as chromSDE
+(which implements a variant of the model defined here), or Pastis-MDS (which
+implements exactly the unconstrained model defined here).
 
@@
 \subsubsection{Volume exclusion models}
+blha
+
+\citet{tjong:physical} showed \textit{S. cerevisiae}'s structure can be mostly
+explained by a large set of constrained yet random population of structures
+following a simple volume exclusion model. We repeated the experiment on {\em
+P. falciparum}, and showed that, despite recovering similar results on {\em S.
+cerevisiae}, the same set of constraints is not sufficient to explain {\em P.
+falciparum}'s Hi-C contact map.
+
+Briefly, let us recall the methodology presented by \citet{tjong:physical}. We
+construct a population of three-dimensional structures by modeling chromosomes
+as random configuration of beads-on-a-string, subjected to the following constraints.
+
+\begin{itemize}
+\item \textbf{Adjacent Beads Constraints} Two adjacent beads spaced 3.2kb apart are constrained to be 30nm apatr.
+\item \textbf{Volume Exclusion Constraints} Each pairs of beads are spaced at minimum XX apart
+\item \textbf{Nucleus Constraint} All beads lie in a spherical nucleus of specific
+radius (XXX)
+\item \textbf{Centromeres Constraints} Centromeres are localized in a small sphere of
+50nm abutting the nuclear envelop.
+\item \textbf{Telomere Constraints} Telomeres are located at most 50nm from the nuclear envelop.
+\end{itemize}
+
+\citet{tjong:physical} then formulates an optimization problem that includes
+in addition to all the aforementioned constraints a penalty term on three
+adjacent beads accounting for chromatin stiffness.
+
+$$
+\frac{1}{2} FIXME
+$$
 

### 2017-07-03 scripts/population_structures/clustering_across_stages_downsampled_distances.py
Now computing PCA on full dataset

@@ -24,7 +24,7 @@ for i, dataset in enumerate(datasets):
             "results/ay2013/%s_10000_raw_%s_distances_chr%02d.npy" % (
                 dataset, algorithm, args.chromosome))
     x = np.load(filename)
-    x = x[:100]
+    x = x[:1000]
     if labels is not None:
         X[i, :len(x)] = x
         labels = np.concatenate([labels, i * np.ones(len(x))])

### 2017-07-03 doc/manuscript/paper.tex
more on VE model
Now using DGJIB's template

@@
-\documentclass[oupdraft]{bio}
-% \usepackage[colorlinks=true, urlcolor=citecolor, linkcolor=citecolor, citecolor=citecolor]{hyperref}
+%\documentclass[2columns]{article}
+
+\documentclass[letterpaper,12pt]{article}
 \usepackage{url}
+\usepackage{graphicx}
+\usepackage{url}
+\usepackage{wrapfig}
+\usepackage[normalem]{ulem}
+\usepackage{float}
+\usepackage{xr}
 
-% Add history information for the article if required
-\history{Received XXX;
-revised XXX;
-accepted for publication XXX}
+\usepackage{amsmath,amssymb,amsfonts,amscd,latexsym}
+\usepackage{tabularx}
+\usepackage{multirow}
+\usepackage{multicol}
+\usepackage{eurosym}
+\usepackage{geometry}
+\usepackage{fancyhdr}
+\usepackage{enumitem}
+\usepackage{xspace}
+
+% Two column foonotes
+\usepackage{dblfnote}
+
+%% This is the recommended preamble for your document.
+
+%% Load De Gruyter specific settings 
+\usepackage{dgjournal}
+
+%% The mathptmx package is recommended for Times compatible math symbols.
+%% Use mtpro2 or mathtime instead of mathptmx if you have the commercially
+%% available MathTime fonts.
+%% Other options are txfonts (free) or belleek (free) or TM-Math (commercial)
+\usepackage{mathptmx}
+
+%% Use the graphics package to include figures
+\usepackage{graphics}
+
+%% Use natbib with these recommended options
+\usepackage[authoryear,comma,longnamesfirst,sectionbib]{natbib}
+
+%\externaldocument{supp}
 
@@
 % Title of paper
-\title{Analysing the three-dimensional structure of \em{P. falciparum}}
+\title{Analysing the three-dimensional structure of P. falciparum}
+\author{Nelle Varoquaux}
 
-% List of authors, with corresponding author marked by asterisk
-\author{Nelle Varoquaux$^\ast$,\\[4pt]
-% Author addresses
-\textit{Department of Statistics, University of California, Berkeley \\
-Berkeley Data Science Fellow}
-\\[2pt]
-% E-mail address for correspondence
-{nelle@berkeley.edu}}
-
-% Running headers of paper:
-\markboth%
-% First field is the short list of authors
-{Nelle Varoquaux}
-% Second field is the short title of the paper
-{XXX}
-
-\maketitle
-
-% Add a footnote for the corresponding author if one has been
-% identified in the author list
-\footnotetext{To whom correspondence should be addressed.}
+%\maketitle
 
 \begin{abstract}
-{My very cool abstract.}
-{Hi-C; three-dimensional structure; and other cool stuff.
-}
+
 \end{abstract}
@@
 
+We use IMP to generate a population of structures.
+
+\paragraph{Adding constraints on VRSM genes}
+
 \subsection{Integrative analysis of gene expression and 3D structure using KernelCCA}
@@
 
-\bibliographystyle{biorefs}
+\bibliographystyle{DeGruyter}
 \bibliography{refs}

### 2017-07-05 scripts/structures/infer_structures_po.py
Fixes on PO

@@ -25,7 +25,7 @@ else:
     filename = args.filename
 
 outname = filename.replace(
-    ".matrix", "_PO_%02d_structure.txt" % (algo, args.seed))
+    ".matrix", "_PO_%02d_structure.txt" % (args.seed))
 if os.path.exists(outname):
     import sys
     sys.exit(0)

### 2017-07-05 scripts/structures/launch_all_jobs.sh

@@ -15,13 +15,11 @@ for FOLDER in ${FOLDERS[@]}; do
 #    -N 1 \
 #    --mail-type=ALL --mail-user=nelle@berkeley.edu;
 
-  sbatch --export=DATASET=${FOLDER} -v -p RM-shared --array 101-1000 --mem 1500Mb \
+  sbatch --export=DATASET=${FOLDER} -v -p RM-shared --array 11-1000 --mem 1500Mb \
     --job-name ${FOLDER}_PO \
     --time=48:00:00 cluster_scripts/slurm_po.sh \
     -N 1 \
     --mail-type=ALL --mail-user=nelle@berkeley.edu;
-
-
 done;
 
 

### 2017-07-05 scripts/volume_exclusion/launch_all_jobs.sh

@@ -4,11 +4,18 @@ STATES=(rings trophozoites schizonts)
 # FOLDERS=(HP1_KD HP1_tagged pv_sporozoite)
 for STATE in ${STATES[@]}; do
 
-  sbatch --export=STATE=${STATE} -v -p RM-shared --array 1-100 --mem 1500Mb \
+  sbatch --export=STATE=${STATE} -v -p RM-shared --array 101-1000 --mem 1500Mb \
     --job-name ${STATE}_VE \
-    --time=4:00:00 cluster_scripts/slurm_ve.sh \
+    --time=1:00:00 cluster_scripts/slurm_ve.sh \
     -N 1 \
     --mail-type=ALL --mail-user=nelle@berkeley.edu;
+
+  sbatch --export=STATE=${STATE} -v -p RM-shared --array 1-1 --mem 1500Mb \
+    --job-name ${STATE}_VE \
+    --time=1:00:00 cluster_scripts/slurm_ve_with_VRSM.sh \
+    -N 1 \
+    --mail-type=ALL --mail-user=nelle@berkeley.edu;
+
 done;
 
 

### 2017-07-05 scripts/volume_exclusion/plasmo_landmark_without_VRSM.py

@@ -98,7 +98,7 @@ sep = 3200       # 3200 bp separation
 chr_bead = {}    # number of beads for each chromosome
 nbead = 0
 bead_start = {}  # bead label starts of a chr
-for i in chr_seq.keys():
+for i in sorted(chr_seq.keys()):
     n = chr_seq[i] / sep + 1
     chr_bead[i] = n
     nbead = nbead + n
@@ -125,7 +125,7 @@ def bead_id(chr, gpos):
 
 def find_chromosome(bid):
     """ Returns a chromosome id given a bead number"""
-    for i in chr_seq.keys():
+    for i in sorted(chr_seq.keys()):
         if bid < bead_start[i] + chr_bead[i] \
            and bid >= bead_start[i]:
             chrid = i
@@ -138,7 +138,7 @@ def find_bead_in_chr(bid):
     Returns a chromosome and bead_order and mid genome position given a
     beadnum
     """
-    for i in chr_seq.keys():
+    for i in sorted(chr_seq.keys()):
         if bid < bead_start[i] + chr_bead[i] \
            and bid >= bead_start[i]:
             order = bid - bead_start[i] + 1  # order starts from 1
@@ -216,7 +216,7 @@ for i in range(nbead):
 # print 'Setting up restraints'
 # Create bonds for consecutive beads in a string
 bonds = IMP.container.ListSingletonContainer(m)
-for id in chr_seq.keys():
+for id in sorted(chr_seq.keys()):
     istart = bead_start[id]
     iend = istart + chr_bead[id]
     bp = IMP.atom.Bonded.setup_particle(chain.get_particle(istart))
@@ -256,7 +256,7 @@ rcell = IMP.container.SingletonsRestraint(sscell, chain)
 centro_rad = 50.0
 centro = IMP.algebra.Vector3D(nuclear_rad - centro_rad, 0, 0)
 listcentro = IMP.container.ListSingletonContainer(m)
-for k in chr_seq.keys():
+for k in sorted(chr_seq.keys()):
     j = bead_id(k, chr_cen[k])
     pcen = chain.get_particle(j)
     listcentro.add(pcen)
@@ -284,7 +284,7 @@ print('before telo: ', score)
 telo = IMP.container.ListSingletonContainer(m)
 # galBead =  bead_id(galpos[0],galpos[1])
 # telo.add_particle(chain.get_particle(galBead))
[…]

### 2017-07-05 scripts/population_structures/compute_distances.sh
Now plotting results and starting from exact MDS results for PO model

@@ -48,3 +48,25 @@ python compute_pairwise_distances_chr.py \
   structures/ay2013/rings_10000_raw_UNB0 \
   --lengths data/ay2013/rings_10000_raw.bed
 
+python compute_downsampled_distances.py \
+  structures/ay2013/trophozoites_10000_raw_PO \
+  --lengths data/ay2013/rings_10000_raw.bed
+
+python compute_downsampled_distances.py \
+  structures/ay2013/trophozoites_10000_raw_UNB0 \
+  --lengths data/ay2013/rings_10000_raw.bed
+
+python compute_pairwise_distances_chr.py \
+  structures/ay2013/trophozoites_10000_raw_PO \
+  --lengths data/ay2013/rings_10000_raw.bed
+
+python compute_pairwise_distances_chr.py \
+  structures/ay2013/trophozoites_10000_raw_UNB0 \
+  --lengths data/ay2013/rings_10000_raw.bed
+
+
+python compute_downsampled_distances.py \
+  structures/ay2013/schizonts_10000_raw_PO \
+  --lengths data/ay2013/rings_10000_raw.bed
+
+

### 2017-07-05 scripts/structures/infer_structures_po.py

@@ -6,6 +6,7 @@ from pastis import fastio
 from minorswing import dispersion
 from minorswing._inference import poisson_structure
 from pastis.optimization import mds
+from utils import compute_wish_distances
 from minorswing._inference import utils
 import iced
 
@@ -78,7 +79,9 @@ normed = normed.tocoo()
 # Compute starting point
 print "Initializing"
 random_state = np.random.RandomState(args.seed)
-X = mds.estimate_X(normed, random_state=random_state)
+wd = compute_wish_distances(normed)
+X = mds.estimate_X(wd, random_state=random_state,
+                   precompute_distances="precomputed")
 X[np.isnan(X)] = 0
 
 # PM2

### 2017-07-05 scripts/structures/plot_results.py

@@ -1,0 +1,148 @@
+import os
+import numpy as np
+from sklearn.metrics import euclidean_distances
+from mayavi import mlab
+from minorswing import fastio
+import argparse
+
+from utils import interpolate_chromosomes
+from minorswing.validation import realign_structures
+
+
+parser = argparse.ArgumentParser()
+parser.add_argument("filename")
+parser.add_argument("--lengths", default=None, type=str)
+parser.add_argument("--counts", default=None, type=str)
+parser.add_argument("--resolution", default=10000)
+parser.add_argument("--outfile", "-o", default=None, type=str)
+parser.add_argument("--rotate", "-r", default=False, action="store_true")
+parser.add_argument("--nucleus-size", "-n", dest="nucleus_size", default=1000,
+                    type=float)
+parser.add_argument("--show", "-s", default=False, action="store_true")
+parser.add_argument("--no-smooth", dest="smooth", default=True,
+                    action="store_false")
+parser.add_argument("--flipped", "-f", default=False, action="store_true")
+args = parser.parse_args()
+
+if "Pv" in args.filename:
+    gene_clusters = np.loadtxt("files/pv_loci.txt", dtype=int)
+else:
+    gene_clusters = np.loadtxt(
+        'files/VRSSMclusterIDs.txt',
+        skiprows=1,
+        usecols=(0, 2, 3), dtype=int)
+
+if args.rotate and "Pv" not in args.filename:
+    X_ = np.loadtxt(
+        "results/gametocyte_structure.txt")
+
+resolution = args.resolution
+
+if args.lengths is not None:
+    if args.lengths.endswith(".npy"):
+        lengths = np.load(args.lengths)
+    else:
+        lengths = fastio.load_lengths(args.lengths)
+else:
+    lengths = fastio.load_lengths(
+        os.path.join(os.path.dirname(args.filename).replace(
+            "results",
+            "data"), os.path.basename(args.filename)[:13] + ".bed"))
+
+centromeres = np.loadtxt("files/pf.cent")
+
+mlab.figure(bgcolor=(1, 1, 1))
+mlab.clf()
+
+f = mlab.gcf()
+f.scene.light_manager.light_mode = 'vtk'
+
+scale = 5
[…]

### 2017-07-05 scripts/structures/utils.py

@@ -12,3 +12,160 @@ def compute_wish_distances(counts, alpha=-3.0, beta=1., use_zero_counts=True):
     wish_distances[m] = 0
     wish_distances[:, m] = 0
     return sparse.coo_matrix(np.triu(wish_distances, 1))
+
+import numpy as np
+from scipy import interpolate
+from scipy.misc import comb
+
+
+def bernstein_poly(i, n, t):
+    """
+     The Bernstein polynomial of n, i as a function of t
+    """
+
+    return comb(n, i) * (t**(n-i)) * (1 - t)**i
+
+
+def bezier_curve(points, nTimes=100):
+    """
+       Given a set of control points, return the
+       bezier curve defined by the control points.
+
+       points should be a list of lists, or list of tuples
+       such as [ [1,1],
+                 [2,3],
+                 [4,5], ..[Xn, Yn] ]
+        nTimes is the number of time steps, defaults to 1000
+
+        See http://processingjs.nihongoresources.com/bezierinfo/
+    """
+
+    nPoints = len(points[0])
+    xPoints = np.array(points[0])
+    yPoints = np.array(points[1])
+    zPoints = np.array(points[2])
+
+    f_z = interpolate.interp1d(
+        np.arange(len(zPoints)), points[2])
+
+    t = np.linspace(0.0, 1.0, nTimes)
+
+    polynomial_array = np.array(
+        [bernstein_poly(i, nPoints-1, t) for i in range(0, nPoints)])
+
+    xvals = np.dot(xPoints, polynomial_array)
+    yvals = np.dot(yPoints, polynomial_array)
+    zvals = f_z(np.arange(len(xvals)))
+
+    return xvals, yvals, zvals
+
+
+def interpolate_chromosomes(X, lengths, kind="cubic", eps=1e-1):
+    """
+    Return a smoothed interpolation of the chromosomes
+
+    Parameters
+    ----------
+    X : ndarray (n, 3)
+        The 3D structure
+
+    lengths : ndarray (L, )
+        The lengths of the chromosomes. Note that the sum of the lengths
+        should correspond to the length of the ndarray of the 3D structure.
+
+    Returns
+    -------
+    smoothed_X : the smoothed 3D structure
+
+    """
+    smoothed_X = []
+    mask = np.invert(np.isnan(X[:, 0]))
+
+    begin, end = 0, 0
+    for j, length in enumerate(lengths):
+        end += length
+        x = X[begin:end, 0]
+        y = X[begin:end, 1]
+        z = X[begin:end, 2]
[…]

### 2017-07-08 doc/manuscript/paper.tex
Processing lemieux's data

@@
 
-While these approaches are widely used, since better motivated and more
-flexible models to infer 3D consensus structures from contact counts have been
-proposed.
-
 \subsubsection{Constrained multi-dimensional scaling}
@@
 $\delta_{ij}$.
-
-\paragraph{Wish distances}
-
-A crucial step of MDS-based method rely on finding a good set of
-``wish-distances'' from Hi-C contact count data. While several approach have
-been used in the litterature, most leverage bio-physics relationships either
-between the genomic distnaces $s$ and the expected count, or the relationship
-between the euclidean distance $d$ and the expected
-count. \citet{ay:three-dimensional} observe that {\em P. falciparum}'s DNA
-behave like a fractal globule polymer. \citet{grosberg:fractal} show fractal
-globul polymer exhibits a well-defined relationship between its genomic length
-and physical distance $d$:
-
-$$d\sim s^{1/3}$$
-
-The relationship between genomic distance $s$ and $c$ can be derived from the
-data, by fitting a linear model between the log expected count and the genomic
-distances.
 
@@
 
-Note that in practice, the constraints are fulfilled naturally by the data,
-and thus can be dropped. In addition of drastically reducing the computation
-time, this allows to use generic publicly available software such as chromSDE
-(which implements a variant of the model defined here), or Pastis-MDS (which
-implements exactly the unconstrained model defined here).
+\paragraph{Wish distances}
 
-\subsubsection{Modeling contact counts as a Poisson distribution}
+A crucial step of MDS-based method rely on finding a good set of
+``wish-distances'' from Hi-C contact count data. While several approach have
+been used in the litterature, most leverage bio-physics relationships either
+between the genomic distnaces $s$ and the expected count, or the relationship
+between the euclidean distance $d$ and the expected
+count. \citet{ay:three-dimensional} observe that {\em P. falciparum}'s DNA
+behave like a fractal globule polymer. \citet{grosberg:fractal} show fractal
+globul polymer exhibits a well-defined relationship between its genomic length
+and physical distance $d$:
+
+$$d\sim s^{1/3}$$
+
+The relationship between genomic distance $s$ and $c$ can be derived from the
+data, by fitting a linear model between the log expected count and the genomic
+distances and extracting the slope coefficient $\alpha$. We thus obtain, up to
+a scaling factor $\beta$ a function to map contact counts to wish distances:
+
+\begin{equation}
+\label{eq:count-to-distance}
+\delta_{ij} = \beta c^{\alpha/3}
+\end{equation}
+
+The scaling factor depends both on the sequencing depth of the libraries, but
+also on the nucleus size of the organims. \citet{ay:three-dimensional} infers
+3D models for a range of parameters and keeps the models that are the most
+plausible, under the assumption that beads should be as uniformely distributed
+in the nucleus as possible. 0-contact counts can either be ignored, or mapped
+to the largest distance possible. Note that for high coverage data set, the
+amount of non-interacting beads is sufficiently small that including those or
+not in the optimization will make very little difference.
+
+\subsubsection{Reproducing the results using a simpler model}
+
+Two simplifications can be brought to the method proposed by
+\citet{ay:three-dimensional}.
+
+First, the ``count-to-distance'' relationship is based on two steps: (1) find
+the relationship between genomic distances and counts; (2) from the
+coefficient, deduce what type of polymer the DNA of this particular organism
+is closest two (fractal globule or equilibrium); (3) use the relationship
+between genomic distance $s$ and euclidean distance $d$ for that particular
+type of polymer. Yet, it is likely that the probability of contact count
+between two elements depends mostly on the average distance separating the two
+beads. As noted in \citet{varoquaux:statistical}, while the relationship
+between both genomic distance $s$ and expected contact count $c$ and the
+relationship between genomic distance $s$ and euclidean distance $c$,
+combining both yields the same count-to-distance function for both polymer
+type:
+
+\begin{equation}
+\delta_{ij} = \beta c^{-1/3}
+\end{equation}
+
+While the difference may seem subtile, fitting the linear model to find
+$\alpha$ is not as straightforward as it seems, as the range of genomic
+distance used can have a large impact on the final coefficient.
+
+The second simplification is to perform an unconstraint optimization. Indeed,
+in practice, the constraints are fulfilled naturally by the data and thus can
+be dropped from the optimization. In addition of drastically reducing the
+computation time, this allows to use generic publicly available software such
+as chromSDE (which implements a variant of the model defined here), or
+Pastis-MDS (which implements exactly the unconstrained model defined here).
 
 \subsubsection{Structure stability}
-- Structure stability
-- at least 100 iterations
-- maybe compare results with chromSDE, PCA \& ShRec3D 
--
+
+Both the constrained and unconstained problem stated above are not convex. The
+reader may well ask how sensitive the result is to initialize. In short, it is
+not. To study the stability of the results with respect to the initialization,
+I perform for each stage presented here 1000 optimization, starting from
+different initialization.
+
+The first question that arise is simply: are structures from the same time
+points but from a different start ponit more alike than structures from
+different time points? To answer this question, I compute for each structure
+and each time point a feature matrix, composed of a subsampled pairwise
+euclidean distances of each structure. The reason for subsampling is simply so
+that the resulting feature matrix fits in memory on a laptop. I then compute a
+stochastic PCA on this feature matrix and plot the projection in 2D space
+using the 3 first components. The results clearly show that structures from
+the same time point cluster strongly. Repeating this experiment on singly
+chromosomes yields the same result.
+
+The second question one can ask is how do the structures differ. Tackling this
+question is very challenging, but can be reformulated in a much easier way.
+Are the hallmarks of interest conserved across structures of a same stage?
+\citet{ay:three-dimensional} and \citet{lemieux:genome-wide} both identified
+{\em P. falciparum} folded in very specific ways, with VRSSM genes highly
+interacting. \citet{ay:three-dimensional} also observed strong clustering of
+the centromeres, and enrichment in interaction at the telomeres. These
+observation can lead to a rigourous approach at identifying whether families
+of loci clustered in the structures and even quantifying this in a rigourous
+manner.
 
 \subsubsection{Volume exclusion models}
-blha
 

### 2017-07-08 scripts/population_structures/compute_downsampled_distances.py

@@ -26,7 +26,6 @@ distances = []
 
 filenames = glob(args.directory + "*_structure.txt")
 filenames.sort()
-filenames = filenames[:100]
 
 distances = []
 

### 2017-07-08 scripts/prepare_lemieux_data/combine_libraries.py

@@ -1,0 +1,27 @@
+from __future__ import print_function
+from iced import io
+from glob import glob
+
+
+libraries = ["MobI", "HindIII"]
+filenames = glob("data/lemieux/*/*_raw.matrix")
+filenames = [filename for filename in filenames
+             if not filename.endswith("_combined_raw.matrix")]
+filenames.sort()
+basenames = set(
+    [filename.strip("_raw.matrix").rstrip("MobI").rstrip("HindIII")
+     for filename in filenames])
+lengths = io.load_lengths("data/ay2013/rings_10000_raw.bed")
+
+
+for basename in basenames:
+    print("Combining", basename)
+    counts = None
+    for filename in filenames:
+        if filename.startswith(basename):
+            if counts is None:
+                counts = io.load_counts(filename, lengths=lengths)
+            else:
+                counts += io.load_counts(filename, lengths=lengths)
+    io.write_counts(
+        basename + "_combined_raw.matrix", counts)

### 2017-07-08 scripts/prepare_lemieux_data/convert_to_hicpro.py

@@ -1,0 +1,44 @@
+import numpy as np
+import os
+from scipy import sparse
+from iced import io
+import gzip
+import argparse
+
+parser = argparse.ArgumentParser()
+parser.add_argument("filename")
+parser.add_argument("--outname", "-o")
+args = parser.parse_args()
+
+filename = args.filename
+
+lengths = io.load_lengths("data/ay2013/rings_10000_raw.bed")
+counts = np.zeros([lengths.sum(), lengths.sum()])
+
+begin_chr = np.concatenate([[0], lengths.cumsum()])
+
+with gzip.open(filename, "rb") as f:
+    for i, line in enumerate(f):
+        chrom1, loc1, chrom2, loc2, c = line.split("\t")
+        loc1, loc2, c = int(loc1), int(loc2), int(c)
+
+        chrom1 = int(chrom1.strip("chr"))
+        chrom2 = int(chrom2.strip("chr"))
+
+        loc1 /= 10000
+        loc2 /= 10000
+
+        loc1 += begin_chr[chrom1 - 1]
+        loc2 += begin_chr[chrom2 - 1]
+
+        counts[loc1, loc2] += c
+
+counts = counts.T + counts
+
+if args.outname is not None:
+    try:
+        os.makedirs(os.path.dirname(args.outname))
+    except OSError:
+        pass
+
+    io.write_counts(args.outname, sparse.coo_matrix(np.triu(counts)))

### 2017-07-08 scripts/prepare_lemieux_data/combine_libraries.py
Preparing lemieux's data

@@ -8,20 +8,23 @@ filenames = glob("data/lemieux/*/*_raw.matrix")
 filenames = [filename for filename in filenames
              if not filename.endswith("_combined_raw.matrix")]
 filenames.sort()
-basenames = set(
-    [filename.strip("_raw.matrix").rstrip("MobI").rstrip("HindIII")
-     for filename in filenames])
-lengths = io.load_lengths("data/ay2013/rings_10000_raw.bed")
-
+basenames = list(set(
+    [filename.split("-")[0]
+     for filename in filenames]))
+basenames.sort()
 
 for basename in basenames:
     print("Combining", basename)
     counts = None
     for filename in filenames:
         if filename.startswith(basename):
+
+            lengths = io.load_lengths(filename.replace("_raw.matrix", ".bed"))
             if counts is None:
                 counts = io.load_counts(filename, lengths=lengths)
             else:
                 counts += io.load_counts(filename, lengths=lengths)
     io.write_counts(
         basename + "_combined_raw.matrix", counts)
+    io.write_lengths(
+        basename + "_combined_raw.bed", lengths)

### 2017-07-08 scripts/prepare_lemieux_data/convert_to_hicpro.py

@@ -12,7 +12,17 @@ args = parser.parse_args()
 
 filename = args.filename
 
-lengths = io.load_lengths("data/ay2013/rings_10000_raw.bed")
+
+lengths = np.loadtxt("files/lengths").astype(float)
+if "10kb" in filename:
+    resolution = 10000
+else:
+    resolution = 25000
+
+lengths /= resolution
+lengths = np.ceil(lengths).astype(int)
+
+
 counts = np.zeros([lengths.sum(), lengths.sum()])
 
 begin_chr = np.concatenate([[0], lengths.cumsum()])
@@ -25,8 +35,8 @@ with gzip.open(filename, "rb") as f:
         chrom1 = int(chrom1.strip("chr"))
         chrom2 = int(chrom2.strip("chr"))
 
-        loc1 /= 10000
-        loc2 /= 10000
+        loc1 /= resolution
+        loc2 /= resolution
 
         loc1 += begin_chr[chrom1 - 1]
         loc2 += begin_chr[chrom2 - 1]
@@ -42,3 +52,4 @@ if args.outname is not None:
         pass
 
     io.write_counts(args.outname, sparse.coo_matrix(np.triu(counts)))
+    io.write_lengths(args.outname.replace("_raw.matrix", ".bed"), lengths)

### 2017-07-08 scripts/visualizing_counts/plot_counts.py

@@ -1,0 +1,116 @@
+import argparse
+import os
+import numpy as np
+from minorswing import fastio
+import iced
+import matplotlib.pyplot as plt
+from matplotlib import colors
+
+
+parser = argparse.ArgumentParser()
+parser.add_argument("filename")
+parser.add_argument("--lengths")
+parser.add_argument("--no-normalization", dest="no_normalization",
+                    default=False, action="store_true")
+args = parser.parse_args()
+if args.lengths is not None:
+    lengths = fastio.load_lengths(args.lengths)
+else:
+    try:
+        lengths = fastio.load_lengths(args.filename.replace(".matrix", ".bed"))
+    except IOError:
+        try:
+            lengths = fastio.load_lengths(
+                args.filename.replace(".matrix", "_abs.bed"))
+        except IOError:
+            try:
+                lengths = fastio.load_lengths(
+                    args.filename.replace("_normalized.matrix", "_abs.bed"))
+            except IOError:
+                lengths = fastio.load_lengths(
+                   args.filename.replace("_normalized.matrix", ".bed"))
+
+counts = fastio.load_counts(args.filename, lengths=lengths)
+
+centromeres = np.loadtxt("files/pf.cent")
+
+AP2 = np.loadtxt(
+    'files/AP2_all_positions.tab',
+    usecols=(1, 2, 3), dtype=str)
+AP2[:, 0] = [int(i[3:]) for i in AP2[:, 0]]
+AP2 = AP2.astype(float)
+
+if "25kb" in args.filename:
+    resolution = 25000
+else:
+    resolution = 10000
+
+counts = np.array(counts.todense())
+counts = counts + counts.T
+counts[np.arange(len(counts)), np.arange(len(counts))] = 0
+
+if not args.no_normalization:
+    counts = iced.filter.filter_low_counts(counts, lengths=lengths,
+                                           sparsity=False, percentage=0.03)
+    # counts = filter_low_mappability(counts, mappabilities=mappability,
+    #                                 cutoff=0.25)
+    counts, bias = iced.normalization.ICE_normalization(counts,
+                                                        output_bias=True)
+
+    counts[np.isnan(counts)] = 0
+
+c = counts.copy().flatten()
+c.sort()
[…]

### 2017-07-08 scripts/structures/plot_distances.py
Now inferring on lemieux et al as well

@@ -32,7 +32,10 @@ else:
     counts = None
 
 centromeres = np.loadtxt("files/pf.cent")
-resolution = 10000
+if "25kb" in args.filename:
+    resolution = 25000
+else:
+    resolution = 10000
 
 X = np.loadtxt(args.filename)
 if len(X.shape) == 1:

### 2017-07-08 scripts/structures/plot_results.py

@@ -13,7 +13,7 @@ parser = argparse.ArgumentParser()
 parser.add_argument("filename")
 parser.add_argument("--lengths", default=None, type=str)
 parser.add_argument("--counts", default=None, type=str)
-parser.add_argument("--resolution", default=10000)
+parser.add_argument("--resolution", default=10000, type=int)
 parser.add_argument("--outfile", "-o", default=None, type=str)
 parser.add_argument("--rotate", "-r", default=False, action="store_true")
 parser.add_argument("--nucleus-size", "-n", dest="nucleus_size", default=1000,
@@ -114,6 +114,7 @@ for j, (x, y, z) in enumerate(smoothed_X):
         else:
             pos = (b + e) / 2
             pos /= float(length) * resolution / m
+            pos = pos.astype(int)
             xp = 2 * linewidth * np.sin(phi) * np.cos(theta) + x[pos]
             yp = 2 * linewidth * np.sin(phi) * np.sin(theta) + y[pos]
             zp = 2 * linewidth * np.cos(phi) + z[pos]

### 2017-07-08 scripts/prepare_ay_data/subsample_20kb.py
subsampling ay data

@@ -1,0 +1,27 @@
+import argparse
+import numpy as np
+from scipy import sparse
+
+from iced import io
+from iced import utils
+
+parser = argparse.ArgumentParser()
+parser.add_argument("filename")
+parser.add_argument("--outname", "-o")
+args = parser.parse_args()
+
+filename = args.filename
+lengths = io.load_lengths("data/ay2013/rings_10000_raw.bed")
+
+
+counts = io.load_counts(filename, lengths=lengths)
+counts = counts.toarray()
+counts = counts.T + counts
+
+new_counts, new_lengths = utils.downsample_resolution(counts, lengths)
+
+new_counts = sparse.coo_matrix(np.triu(counts))
+
+
+io.write_counts(args.outname, new_counts)
+io.write_lengths(args.outname.replace(".matrix", ".bed"), new_lengths)

### 2017-07-08 scripts/prepare_ay_data/subsample_20kb.py
Small fix in subsampling

@@ -20,7 +20,7 @@ counts = counts.T + counts
 
 new_counts, new_lengths = utils.downsample_resolution(counts, lengths)
 
-new_counts = sparse.coo_matrix(np.triu(counts))
+new_counts = sparse.coo_matrix(np.triu(new_counts))
 
 
 io.write_counts(args.outname, new_counts)

### 2017-07-09 doc/manuscript/paper.tex
Now estimating beta before running PO

@@
 
+A number of methods have been proposed to accomplish such tasks, which I will
+review in this section.
+
 \subsubsection{Volume exclusion models}
@@
 
-In order to integrate
+\citet{ay:three-dimensional}  exploited the 3D models to relate structure to
+gene expression. The authors rely on a technic called ``kernelCCA'' with
+attempts to find components on different view of the data in such a way that
+these are coherent with one another.
 
@@
 review? Can the results be interpretable?
-
-\subsection{Differential analysis}
-\note{XXX only possible if Bunnik 2017 paper published.}
 

### 2017-07-09 scripts/structures/infer_structures_po.py

@@ -9,6 +9,7 @@ from pastis.optimization import mds
 from utils import compute_wish_distances
 from minorswing._inference import utils
 import iced
+from sklearn.metrics import euclidean_distances
 
 """
 Launches the inference of the 3D model on .matrix/.bed files"""
@@ -29,7 +30,7 @@ outname = filename.replace(
     ".matrix", "_PO_%02d_structure.txt" % (args.seed))
 if os.path.exists(outname):
     import sys
-    sys.exit(0)
+    #sys.exit(0)
 
 try:
     os.makedirs(os.path.dirname(filename))
@@ -91,14 +92,26 @@ outname = filename.replace(
 if os.path.exists(outname):
     print "Already computed"
     import sys
-    sys.exit()
+    #sys.exit()
 
 alpha = -3.
 max_iter = 1000000
 
-counts = counts.tocoo()
+bias = bias.reshape(-1, 1)
+counts = counts.toarray()
+m  = (counts.sum(axis=0) + counts.sum(axis=1)) == 0
+counts[m, :] = np.nan
+counts[:, m] = np.nan
+
+m, n = X.shape
+d = euclidean_distances(X)
+mask = (np.invert(np.tri(m, dtype=np.bool)) & np.invert(np.isnan(counts)))
+beta = counts[mask].sum() / ((d[mask] ** alpha) * (bias * bias.T)[mask]).sum()
+counts[np.isnan(counts)] = 0
+
+counts = sparse.coo_matrix(counts)
 X = poisson_structure.estimate_X(
-    counts, alpha=-3., beta=1., bias=bias,
+    counts, alpha=-3., beta=beta, bias=bias,
     verbose=1,
     maxiter=max_iter,
     use_zero_entries=True,

### 2017-07-09 scripts/structures/plot_distances.py

@@ -34,6 +34,8 @@ else:
 centromeres = np.loadtxt("files/pf.cent")
 if "25kb" in args.filename:
     resolution = 25000
+if "20000" in args.filename:
+    resolution = 20000
 else:
     resolution = 10000
 

### 2017-07-09 doc/manuscript/paper.tex
Progress on the paper

@@
 % Title of paper
-\title{Analysing the three-dimensional structure of P. falciparum}
+\title{Analysing the three-dimensional structure of {\em P. falciparum}}
 \author{Nelle Varoquaux}
@@
 \section{Introduction}
-\label{sec1}
+\label{sec:introduction}
 
-\section{Methods}
-\label{sec2}
+With more than 217 millions cases nearly 429 000 deaths in 2015,  malaria
+remains a major disease burden in tropical and subtropical countries and an
+impediment to economic development. In Africa, where more than 90\% of the
+cases and deaths occurs, the effect of malaria extend far beyond direct
+measures of mortality: the disease is thought to cost more than US\$21
+billion a year \citep{onwejkwe:do}.
 
-\subsection{Analysing contact count data}
+In humans, malaria is caused by a unicellular protozoan {\em Plasmodium} of
+the following species: {\em P. vivax}, {\em P. malariae}, {\em P. ovale}, {\em
+P. knowelis}, and with 99\% of the deaths associated to it, by far the most
+deadly {\em P. falciparum}. The parasite has a complex life cycle, with
+multiple stage both in the human and mosquito host. The human host is infected
+with sporozoites through the bite of a infected female  {\em Anopheles}
+mosquito. The sporozoites quickly migrate to the liver, where they start a two
+weeks multiplication process. They are then released as merozoites in the
+bloodstream, and infect red blood cells. The parasites then start their
+"erythrocytic" stage, via another round of replications in red blood cells,
+through an unusual process of cell division called schizogony. The parasite
+first undergoes multiple rounds of nuclear replication and division  into 13
+to 32 daughter cells, until the red blood cell bursts and provokes the release
+of merezoites and the infective cycle starts anew. This asexual replication
+cycle is responsible for the symptoms and the complications of the disease:
+anemia, tertian fever, \dots
+
+While efficacious vaccines still remains a hope and drug resistance to anti
+malaria continues to arise, genomic-based research on malaria seeks new
+avenues for developing novel therapies \citet{kirchner:recent}. One of the
+limiting factors of developments of new drugs is our poor understanding of the
+mechanisms underlying the parasite's complex life cycle. While the development
+of {\em P. falciparum} through the different stages of its life is thought to
+be driven by coordinating changes in gene expression, the relative paucity of
+transcription factors points to unusual gene regulatory mechanisms . Meanwhile,
+the relative abundance of proteins related to chromatin structures, mRNA decay
+and translation rates suggests alternative mechanisms of gene regulation at the
+epigenetic and post-translational levels
+\citep{cui:chromatin-mediated, duffy:role, hoeijmakers:placing,
+horrocks:control, deitsch:mechanisms}.
+
+\fixme{introduction on HiC}
+
+We discuss below the analysis methods used and developed for studying the 3D
+architecture of {\em P. falciparum}'s genome.
+
+\section{Profiling {\em P. falciparum}'s genome architecture during the
+asexual cycle}
+\label{sec:data}
+
+- Difficulty of the protocol on the AT-rich genome of p. falciparum.
+- Summary of the data
+
+\section{Analyzing contact count data}
 
@@
 - Enrichment of counts
+- power-law analysis
+- 
 - Anything else?
@@
 
-\subsection{Inferring the 3D structures}
+\section{Inferring the 3D structures}
 
@@
 
-\subsubsection{Constrained multi-dimensional scaling}
+\subsection{Constrained multi-dimensional scaling}
 
@@
 
-\subsubsection{Reproducing the results using a simpler model}
+\subsection{Reproducing the results using a simpler model}
 
@@
 
-\subsubsection{Structure stability}
+\subsection{Structure stability}
 
@@
 
-A number of methods have been proposed to accomplish such tasks, which I will
-review in this section.
+\section{3D gene set enrichment}
 
-\subsubsection{Volume exclusion models}
+To assess whether groups of genes are colocolized in a 3D model,
+\citet{ay:three-dimensional} leveraged a statistical method developped by
+\citet{witten:XXX}, which requires labeling each pair of loci in two groups:
+"close" or "far".
+
+The authors used varying distance
+thresholds (10\%, 20\% and 40\% of the nuclear diameter) to deem a locus pair
+“close” and labeled all remaining
+pairs in the set as “far”.
+
+\fixme{Re-read Witten and Noble}
+
+This approach dichotomize loci pairs into two groups, and check for the
+enrichment of a label in one of the two groups. \citep{capurso:distance-based}
+presents an approach that avoids this step, and instead directly estimates the
+significance within the 3D model. Briefly, for a group $\mathcal{G}$, MPED
+computes a test statitics:
+\begin{equation*}
+M = \underset{i,j \in \mathcal{G}| c_i != c_j}{\text{median}} d_{ij}\,,
+\end{equation*}
+
+where $d_{ij}$ is the euclidean distance between bead $i$ and bead $j$. The
+null distribution is estimated empirically by resampling $10^5$ with
+preservation of the chromosome structure. If the $M$ statistics is smaller
+than the mean of the null distribution, it is compared to the lower tail of
+the distribution and indicates co-localization. If the $M$ statistics is
+larger than the mean of the null distribution, then it is compared to the
+upper tail of the distribution, and indicates dispersion.
+
+Applying MEPD to the Ring stage, \citet{capurso:distance-based} confirm that
+centromers, telomers, VRSM genes (both overall, subtelomeric and internal)
+colocalize.
+
+\section{Volume exclusion models cannot explain }
 
@@
 
-\subsection{Integrative analysis of gene expression and 3D structure using KernelCCA}
+\section{Integrative analysis of gene expression and 3D structure using KernelCCA}
 
@@
 review? Can the results be interpretable?
-
-\section{Results}
-\label{sec3}
 

### 2017-07-09 scripts/structures/utils.py

@@ -60,7 +60,7 @@ def bezier_curve(points, nTimes=100):
     return xvals, yvals, zvals
 
 
-def interpolate_chromosomes(X, lengths, kind="cubic", eps=1e-1):
+def interpolate_chromosomes(X, lengths, kind="cubic", eps=1e-1, step=0.1):
     """
     Return a smoothed interpolation of the chromosomes
 
@@ -119,7 +119,7 @@ def interpolate_chromosomes(X, lengths, kind="cubic", eps=1e-1):
         m = np.arange(len(x))[indx]
 
         if kind == "bezier":
-            m = np.arange(m.min(), len(x)-1, 0.1)
+            m = np.arange(m.min(), len(x), step)
             f_x, f_y, f_z = bezier_curve(
                 [x[indx], y[indx], z[indx]])
             smoothed_X.append([f_x,
@@ -130,11 +130,11 @@ def interpolate_chromosomes(X, lengths, kind="cubic", eps=1e-1):
             f_y = interpolate.Rbf(m, y[indx])
             f_z = interpolate.Rbf(m, z[indx])
 
-            m = np.arange(m.min(), len(x)-1, 0.1)
+            m = np.arange(m.min(), len(x), 0.1)
 
-            smoothed_X.append([f_x(np.arange(m.min(), m.max(), 0.1)),
-                               f_y(np.arange(m.min(), m.max(), 0.1)),
-                               f_z(np.arange(m.min(), m.max(), 0.1))])
+            smoothed_X.append([f_x(np.arange(m.min(), m.max(), step)),
+                               f_y(np.arange(m.min(), m.max(), step)),
+                               f_z(np.arange(m.min(), m.max(), step))])
 
         else:
             f_x = interpolate.interp1d(m, x[indx],
@@ -144,11 +144,11 @@ def interpolate_chromosomes(X, lengths, kind="cubic", eps=1e-1):
             f_z = interpolate.interp1d(m, z[indx],
                                        kind=kind)
 
-            m = np.arange(m.min(), len(x)-1, 0.1)
+            m = np.arange(m.min(), len(x), step)
 
-            smoothed_X.append([f_x(np.arange(m.min(), m.max(), 0.1)),
-                               f_y(np.arange(m.min(), m.max(), 0.1)),
-                               f_z(np.arange(m.min(), m.max(), 0.1))])
+            smoothed_X.append([f_x(np.arange(m.min(), m.max(), step)),
[…]

### 2017-07-09 doc/manuscript/paper.tex
Adding table

@@
 
+In the past year, two groups have published novel dataset studying {\em P.
+falciparum}'s 3D genome architecture using Hi-C. \citet{lemieux:genome-wide}
+studied several strains associated with populations expression unique
+\textit{var} genes, in order to study key folding properties of the 3D
+structure relating to \textit{var} gene expression.
+\citet{ay:three-dimensional} f three key-point of the development
+cycle.
+
 - Difficulty of the protocol on the AT-rich genome of p. falciparum.
 - Summary of the data
+
+\begin{table}
+\footnotesize
+\begin{tabular}{cccccc}
+\hline
+\textbf{Name} & \textbf{Strain} & \textbf{Stage} & \textbf{Resolution} &
+\textbf{Number of contacts} & \textbf{Reference} \\
+\hline
+\hline
+Ay-\textit{rings} & 3D7 & Late Rings & 10~kb & & \cite{ay:three-dimensional} \\
+Ay-\textit{trophozoites} & 3D7 & Trophozoites &10~kb &  & \cite{ay:three-dimensional} \\
+Ay-\textit{schizonts} & 3D7 & Schizonts & 10~kb & & \cite{ay:three-dimensional} \\
+Lemieux-\textit{A4+} & IT/BC6+ & Rings & 25~kb & & \cite{lemieux:genome-wide} \\
+Lemieux-\textit{A4} & IT/BC6- & Rings & 25~kb & & \cite{lemieux:genome-wide}\\
+Lemieux-\textit{A4} & IT/3G8 & Rings &  25~kb & & \cite{lemieux:genome-wide}\\
+Lemieux-\textit{DCJ\_On} & NF54/DCJ on & Rings & 25~kb &  &\cite{lemieux:genome-wide} \\
+Lemieux-\textit{DCJ\_Off} & NF54/DCF Off & Rings & 25~kb &  &\cite{lemieux:genome-wide} \\
+Lemieux-\textit{B15C2} & NF54/B15C2 & Rings & 25~kb &  & \cite{lemieux:genome-wide}\\
+\hline
+\end{tabular}
+\caption{Summary of the available {\em P. falciparum} Hi-C datasets}{ The
+number}
+\end{table}
 
@@
 \citet{ay:three-dimensional} leveraged a statistical method developped by
-\citet{witten:XXX}, which requires labeling each pair of loci in two groups:
+\citet{witten:assessment}, which requires labeling each pair of loci in two groups:
 "close" or "far".

### 2017-07-23 doc/manuscript/paper.tex
more

--- 2017-07-09 doc/manuscript/paper.tex
@@
 
+\newcommand{\OMIT}[1]{}
+\newcommand{\Xb}{\textbf{X}}
+\newcommand{\RR}{\mathbb{R}}
+\newcommand{\Dcal}{\mathcal{D}}
+\newcommand {\br}[1]{\left(#1\right)}
+
 \begin{document}
@@
 impediment to economic development. In Africa, where more than 90\% of the
-cases and deaths occurs, the effect of malaria extend far beyond direct
-measures of mortality: the disease is thought to cost more than US\$21
+cases and deaths occurs, the effects of malaria extend far beyond direct
+measures of mortality and the disease is thought to cost more than US\$21
 billion a year \citep{onwejkwe:do}.
 
-In humans, malaria is caused by a unicellular protozoan {\em Plasmodium} of
-the following species: {\em P. vivax}, {\em P. malariae}, {\em P. ovale}, {\em
-P. knowelis}, and with 99\% of the deaths associated to it, by far the most
-deadly {\em P. falciparum}. The parasite has a complex life cycle, with
-multiple stage both in the human and mosquito host. The human host is infected
-with sporozoites through the bite of a infected female  {\em Anopheles}
-mosquito. The sporozoites quickly migrate to the liver, where they start a two
-weeks multiplication process. They are then released as merozoites in the
-bloodstream, and infect red blood cells. The parasites then start their
-"erythrocytic" stage, via another round of replications in red blood cells,
-through an unusual process of cell division called schizogony. The parasite
-first undergoes multiple rounds of nuclear replication and division  into 13
-to 32 daughter cells, until the red blood cell bursts and provokes the release
-of merezoites and the infective cycle starts anew. This asexual replication
-cycle is responsible for the symptoms and the complications of the disease:
-anemia, tertian fever, \dots
+Malaria is caused by small unicellular protozoan {\em Plasmodium}, infecting
+its host through a mosquito bite. Out of the 5 species that can infect human,
+{\em P. falciparum} is by the far the most deadly with 99\% of the deaths
+associated to it. The parasite has a complex life cycle, with multiple stages
+both in the human and mosquito host. The human host is infected with
+\textit{sporozoites} through the bite of a infected female  {\em Anopheles}
+mosquito. The \textit{sporozoites} quickly migrate to the liver, where they
+start a two weeks multiplication process. They are then released as
+\textit{merozoites} in the bloodstream and proceed to infect red blood cells. The
+parasites then start their ``erythrocytic'' cycle (through \textit{rings},
+\textit{trophozoites} and \textit{schizonts} stages), via another round of
+replications in red blood cells, through an unusual process of cell division
+called \textit{schizogony}. The parasite first undergoes multiple rounds of nuclear
+replication and division into 13 to 32 daughter cells, until the red blood
+cell bursts and provokes the release of \textit{merezoites} and the infective cycle
+starts anew. This asexual replication cycle is responsible for the symptoms
+and the complications of the disease: anemia, tertian fever, \dots
 
@@
 of {\em P. falciparum} through the different stages of its life is thought to
-be driven by coordinating changes in gene expression, the relative paucity of
+be driven by coordinated changes in gene expression, the relative paucity of
 transcription factors points to unusual gene regulatory mechanisms . Meanwhile,
@@
 
-\fixme{introduction on HiC}
+In recent year, chromosome conformation captures-like method, broadly referred
+to as Hi-C allowed to identify physical interactions between two regions in a
+genome-wide fashion, yielding information on their relative spatial distance
+in the nucleus. Hi-C has opened new avenues for more systematic analysis of
+the three-dimensional folding of the genome, paving the way for a better
+understanding of the relations between 3D structure and gene regulation,
+replication timing, epigenetic changes as well as many other biological
+processes. The 3D genome of a wide variety of organisms has been studied in
+the past year, including many species of yeasts \citet{duan:three-dimensional,
+burton:xx}, bacteria, flies, plants and numerous human and mouse cell lines.
+Within the past few years, two studies of the three-dimensional structure of
+{\em P. falciparum} have been released.  \citet{lemieux:genome-wide} probed
+several strains of {\em P. falciparum} to understand the link between 3D
+structure and the complex regulation of \textit{var} genes, a family of genes
+involved the invasion of red blood cell and also responsible for the
+parasite's great capacity to evade immuno-therapy.
+\citet{ay:three-dimensional} assayed the 3D structure of the genome of the
+parasite during three key stages of the erythrocytic cycle.
+\citet{ay:multiple} reviews the key biological findings of those studies.
 
-We discuss below the analysis methods used and developed for studying the 3D
-architecture of {\em P. falciparum}'s genome.
+An important part of analysing the three-dimensional structure of the
+genome of {\em P. falciparum} (as well as many other organisms) relied on the
+inferring accurate 3D models of the structure. These models were then used as
+a preliminary step to various analysis, such as identifying colocolized
+elements, distinguishing between open and closed chromatin and last but not
+least, intergrating with other sources of data, such as gene expression or
+chromatin modification.
 
-\section{Profiling {\em P. falciparum}'s genome architecture during the
-asexual cycle}
-\label{sec:data}
+The study of chromosome organization based on contact count maps broadly
+falls into two categories: model-based studies and data-driven studies. The
+former methods consider the polymer nature of DNA to leverage the theoretical
+and computational work done in statistical physics of polymers to build with
+as few assumptions as possible many chromosome conformations. Those chromosome
+conformations are then used to compare against experimental data, such as Hi-C
+contact count matrices, in order to iteratively improve the models. These
+models offer mechanistical insights into the folding of DNA. The latter
+approaches use the experimental data to infer 3D models, by typically minizing
+a cost function ensuring the models are as consistent as possible with the
+data. These data driven models and analysis, with a particular focus on the
+one applied to {\em P. falciparum} are the primary focus of this
+review.
 
-- Difficulty of the protocol on the AT-rich genome of p. falciparum.
-- Summary of the data
+% XXX See rosa:computational
 
-\section{Analyzing contact count data}
+Though we here review some of the methods used to study and build models, this
+is a very incomplete view of a blooming field. \citet{rosa:computational}
+provide a more thorough (but again incomplete) overview of computational models
+of genome architectures.
 
-- Normalization
-- Quality assessment
-- Enrichment of counts
-- power-law analysis
-- 
-- Anything else?
-    - Did we do a compartment analysis?
-- Can we extend it?
-  - Fit-Hic ?
+\section*{Three-dimensional models of DNA from Hi-C data}
 
-\section{Inferring the 3D structures}
+In this section, we review ``data-driven'' methods to infer 3D models of DNA
+from Hi-C data. These methods broadly fall into two categories: (i)
+\textit{consensus} approaches, that aim at inferring a unique mean structures
+best representing the contact count data; (ii) \textit{ensemble} methods that
+yield a population of structures.
 
-An important part of analysing the 3D structure of the genome of \textit{P.
-falciparum} relied on having consensus models for each stage of the life
-cycle. The method presented in \citet{ay:three-dimensional} relied on an
-approach called ``multidimensional scaling'' (MDS), widely used in the
-literature for the purpose of 3D structure inference
-\citep{duan:three-dimensional, varoquaux:statistical, bau:, tanizawa:}. Each chromosomes is
-modeled a series of beads, spaced ~10~kb apart. MDS-based method consists of
-two steps: first convert the
-contact counts $c_{ij}$ into ``wish-distances'' $\delta_{ij}$; then attempt
-to place each bead in 3D such that the euclidean distances $d_{ij}$ separating
-beads $i$ and $j$ matches as closely as possible the wish distance
-$\delta_{ij}$.
+Both consensus and ensemble methods have benefits and drawbacks. Ensemble
+approaches are more biologically accurate: Hi-C data are derived from a
+population of cells, each of those with a uniquely folded 3D structure. A
+population of structure thus better may represents the variability of
+structures among ours cells. Yet, in addition of being more complex and
+computationally intensive, ensemble approaches raise the question of
+interpretability. Often, one has to fall back to interpreting the mean
+structure \citep{kalhor:genome}, or a reduced set of structures
+\citep{rousseau:three}. On the other hand, consensus approaches yield a single
+structure recapitulating the rich information provided by Hi-C data, including
+hallmarks of genome architecture shared across all cells despite the
+cell-to-cell variability \citet{nagano:single-cell}. More amenable for
+visualization and analysis, this average structure can be easily integrated
+with other sources of data, such as RNA-seq, chip-seq etc, which are also
+population-based.
 
-\subsection{Constrained multi-dimensional scaling}
+After introducing some notations and common terminology, I will dive both in consensus
+and population inference methods.
 
-The method presented in \citet{ay:three-dimensional} relied on an
-approach called ``multidimensional scaling'' (MDS), widely used in the
-literature for the purpose of 3D structure inference
-\citep{duan:three-dimensional, varoquaux:statistical, bau:, tanizawa:}. Each chromosomes is
-modeled a series of beads, spaced ~10~kb apart. MDS-based method consists of
-two steps: first convert the
-contact counts $c_{ij}$ into ``wish-distances'' $\delta_{ij}$; then attempt
-to place each bead in 3D such that the euclidean distances $d_{ij}$ separating
-beads $i$ and $j$ matches as closely as possible the wish distances
-$\delta_{ij}$.
+\subsection*{Notation}
 
-\paragraph{Optimization}
+A typical 3C/Hi-C experiment can be summarized as a contact
+count matrix $\mathbf{C} \in R^{n \times n}$, where each row and column
+correspond to a genomic-window (or loci), and each entry $c_{ij}$ to the
+number of times loci $i$ and $j$ have been observed interacting. Contact count
+matrices are subjected to biological and technical biases that need to be
+corrected \citep{imakaev:iterative, cournac:normalization, yaffe:probabilistic,
+hu:hicnorm}.
 
-Given the resulting wish-distances, we cast the inference of the consensus
-structure $\mathbf{X}^{3 \times n}$ as the following constrained weighted MDS:
+Methods model chromosomes as a series of beads in 3D, each bead corresponding
+to a specific genomic window. We denote by $\textbf{X} \in \RR^{3 \times n}$
+the coordinate matrix of the structure(s) and $x_i \in \RR^3$ corresponds to
+the coordinate matrix of the i-th beads.
 
-\begin{equation}\label{eq:mds1} 
+\subsection*{Consensus models}
+
+Consensus methods aim at inferring a model that best represents the contact
+count matrix. They model each chromosome by a chain of beads, and attempt to
+place the beads in a 3D space to minimize an objective function
+$\mathcal{O}(\mathbf{O}, \mathbf{C})$, sometimes
+under constraints.
+
+\begin{equation*}
 \renewcommand{\arraystretch}{2}
-\begin{array}{ccll} 
-\underset{\mathbf{X}}{\text{minimize}} & & 
-\underset{(i,j) \in \mathcal{D}}{\sum} \frac{1}{\delta^2_{ij}}
-\big(d_{ij}(\mathbf{X}) - \delta_{ij}\big)^2 & \\
-\text{subject to} & & d_{ij} \leq r^2_\text{max} \quad \forall i, j &\,,\\
-& & d_{i, i+1} \leq b_\text{max} \quad i = \{i:n|\text{chr}_i = \text{chr}_{i+1}\,,&\\ 
+\begin{array}{ccl}
+\underset{\Xb}{\text{minimize}} & & \mathcal{O}(\mathbf{C}, \Xb)
+\end{array}
+\end{equation*}
+
+\subsubsection*{Metric MDS-based methods}
+
+Early methods \citep{dekker:capturing, duan:three-dimensional,
+tanizawa:mapping, ay:three-dimensional} cast the
+3D inference problem as a multidimensional scaling (MDS) problem: chromosomes
+are models as a chain of beads, to place in 3D such that the distance between
+bead $i$ and $j$ matches as closely as possible a ``wish-distance''
+$\delta_{ij}$ derived
+from contact counts.
+
+\begin{equation*}
+\renewcommand{\arraystretch}{2}
+\begin{array}{ccl}
+\underset{\Xb}{\text{minimize}} & & \underset{i, j \in \mathcal{D}}{\sum}
+\frac{1}{w_{ij}} (d_{ij} - \delta_{ij})^2 \\
+\text{subject to} & & \text{Biologically motivated constraints} \\
+\end{array}
+\end{equation*}
+
+The first step is thus to find an adequate count-to-distance mapping to
+convert contact counts into pairwise wish-distances $\delta_{ij}$.
+\citet{dekker:capturing} models the 78 pairwise contact count of {\em S.
+cerevisiae}'s chromosome III as a Worm-like chain polymer,
+\citet{duan:three-dimensional} uses a linear mapping between contact counts
+and wish distances, \citet{ay:three-dimensional} relies on the biophysical
+properties of fractal globule polymers, \citet{tanizawa:mapping} inferred the
+count-to-distance mapping by fitting the relationship using known pairwise
+distances obtained through high-resolution FISH measures. While parameters of
+these counts-to-distance functions vary, most of them can be summarized as a
+power-law relationship : $c_{ij} = \beta d_{ij}^\alpha$.
+
+\citet{lesne:3d} proposes a different approach to convert counts into
+distances: they borrow the concept of shortest path from graph theory. A
+graph is constructed from the contact map by considering each loci as a node.
+Loci seen interacting are connected through an edge of weight
+$\frac{1}{c_{ij}}$. The authors then compute wish-distances $\delta_{ij}$ as
+the shortest path between node $i$ and $j$ in the graph. Constructing
+wish-distances in such a way has two advantages: (i) low contact counts do not
+contribute much to the wish-distances; (ii) the resulting wish-distances form
+a distance matrix, and thus an optimal solution to the MDS can be found.
+
+In addition of having subtle differences in the derivation of wish distances,
+the different methods also vary by the inclusion of weights to reflect
+confidence in ``wish-distances'' \citep{ay:three-dimensional}, or constraints
+to reflect prior knowledge on the structure. Table~\ref{table:mds_detail}
+summarizes the characteristic of each method.
+
+\begin{table*}
+\scriptsize
+\centering
+\begin{tabular}{cccccc}
+\hline
+\textbf{Paper} & Organism & Full or partial genome & Counts-to-distance mapping & \textbf{Constraints}
+& \textbf{Weights}\\
+\hline
+\cite{dekker:capturing} & {\em S. cerevisiae} & Chr III & Worm-like chain behavior & - \\
+\cite{duan:three-dimensional} & {\em S. cerevisiae} & Whole-genome & Linear
+relationship & Adjacent beads, rDNA & - \\
+\cite{tanizawa:mapping} & {\em S. pombe} & Whole-genome &
+FISH-derived relationship & Adjacent beads, rDNA, telomeres, centromeres & - \\
+\cite{ay:three-dimensional} & {\em P. falciparum} & Whole-genome & Fractal
+globule derived relationship & adjacent
+beads & $w_{ij} = \frac{1}{\delta_{ij}^2}$ \\
+\cite{lesne:3d} & - & Whole-genome & \textit{ad hoc} & - & - \
+\end{tabular}
+\caption{Differences between MDS-based methods}{}
+\label{table:mds_detail}
+\end{table*}
+
+\subsubsection*{Non-metric MDS-based method}
+
+A crucial step of MDS-based method is the conversion of counts into
+wish-distances. As described earlier, converting contact counts into
+wish-distances require strong assumptions that may not be met in practice. For
+example, this mapping may change from one organism to another
+\citep{fudenberg:higher-order}, from one resolution to another
+\citep{zhang:inference}, from one genomic distance range to another
+\citep{ay:statistical}, or from one time point to another during the cell
+cycle \citep{le:high-resolution, ay:three-dimensional}.
+
+Two methods have since been proposed to alleviate this problem.
+\citet{ben-elazar:spatial} proposes to cast the inference as a
+\emph{non-metric MDS} \citep{kruskal:multidimensional}, where the 3D structure
+is inferred jointly with the wish-distances. \fixme{except they do a bunch of
+stuff to the count data}.
+
+\citet{zhang:inference} proposes to parametrize the count-to-distance mapping
+as a power-law ($d = \beta c^\alpha)$, and to infer jointly the parameters
+jointly with the structure. In addition, the authors propose to add a
+penalization term for non-interacting beads. The proposed penalization allows
+to cast the problem as a convex optimization and thus guarantees to retrieve
+the optimal structure.
+
+\subsubsection*{Other ad-hoc optimization methods}
+
+A series of methods cast the inference as an optimization method, where the
+objective function is designed by hand to fullfill a number of properties
+\citep{trieu:large, trieu:3D, trieu:MOGEN}.
+
+For example, \citep{trieu:large} formulates an optimization problem that
+attempts to place the beads such that, if a significant contact is observed
+between $i$ and $j$, then $\| x_i - x_j \| < d_c$. On the other hand, if a
+no significant contact is observed, then the two beads should be no closer
+than $d_c$. In addition, the authors also place constraints on the minimum
+distance separating two beads and maximum distances between two beads. To
+satisfy all those properties, the authors propose to minimize the following
+objective function:
+
+\fixme{add damn objective function}
+
+This method is then generalized to include \textit{trans} contact counts data
+\citep{trieu:MOGEN}.
+
+\cite{trieu:MOGEN} proposes to exploit the Lozentzian function, to formulate a
+constraint-based problem. They first convert the contact counts $c_{ij}$ into
+wish-distances, and cast the following optimization problem:
+
+\begin{equation*}
+\renewcommand{\arraystretch}{2}
+\begin{array}{ccl}
+\underset{\Xb}{\text{maximize}} & & \underset{| i - j | = 1}{\sum} \frac{c^2
+c_\text{max}}{c^2 + (d_{ij} - \delta_{ij})} + 
+\underset{| i - j | \neq 1}{\sum} \frac{c^2 c_{ij}}{c^2 + (d_{ij} - \delta_{ij})}
+\end{array}
+\end{equation*}
+
+This objective problem attempts to minimize the difference between the
+wish-distance $\delta_{ij}$ and the euclidean distance $d_{ij}$ between bead
+$i$ and $j$ similarly to MDS-based method, but exploiting the property of the
+Loretzian function to have very high derivative when constraints are broken.
+
+\subsubsection*{Statistical-based models}
+
+\citet{varoquaux:statistical} proposes a novel approach based on
+statistical modeling of contact counts as random Poisson variables where the
+intensity is a function of the distance: $c_{ij} \sim \text{Poisson}(\beta
+d_{ij}^\alpha)$, and some perform the optimization only including a subset of
+contact counts. The authors propose to cast the
+inference of 3D models as maximizing the likelihood:
+
+\begin{equation}
+\renewcommand{\arraystretch}{2}
+\begin{array}{cll}
+\underset{\alpha, \beta, \textbf{X}}{\text{max}} &
+\mathcal{L}(\mathbf{X}, \alpha, \beta) = \underset{i<j\leq n}{\sum}  c_{ij}
+\alpha \log d_{ij} + c_{ij} \log \beta - \beta d_{ij}^\alpha &\\
 \end{array}
@@
 
-The various constraints corresponds to {\em prior} knowledge on the 3D
-structure, and vary across organisms, time-points of the life cycle and
-cell-type. For {\em P. falciparum}, the constraints are as follow.
+\subsection*{Ensemble methods}
 
-\begin{itemize}
-\item All beads lie in a spherical nucleus centered on the origin and of a
-certain size.
-\item Two adjacent beads must not be too far apart. Leveraging the work of
-\citet{berger:high TOVERIFY}, we assume that $b^\text{max} = 91~nm$
-\end{itemize}
+\citet{bau:three-dimensional}
+\cite{umbarger:three-dimensional}
+\cite{hu:bayesian}
 
-\paragraph{Wish distances}
+\subsection*{Single-cell models}
 
-A crucial step of MDS-based method rely on finding a good set of
-``wish-distances'' from Hi-C contact count data. While several approach have
-been used in the litterature, most leverage bio-physics relationships either
-between the genomic distnaces $s$ and the expected count, or the relationship
-between the euclidean distance $d$ and the expected
-count. \citet{ay:three-dimensional} observe that {\em P. falciparum}'s DNA
-behave like a fractal globule polymer. \citet{grosberg:fractal} show fractal
-globul polymer exhibits a well-defined relationship between its genomic length
-and physical distance $d$:
+The last category of data-driven methods to infer the 3D architecture of the
+genome rely on a new protocol to probe single-cell for their 3D structures
+\citep{nagano:single-cell, XXX}. Single-cell Hi-C is still in its early days,
+and only a handful of data sets are today publicly available. The contact maps
+originating from these data sets are very sparse, and specific methods to
+infer 3D structures need to be developed specifically for sc-HiC. The
+challenge of constructing models from this data has been tackled very
+differently by two groups.
 
-$$d\sim s^{1/3}$$
+\citet{nagano:single-cell} models the chromosome as a series of beads linked
+as a flexible fiber. Loci seen interacting in the single-cell contact maps are
+constrained to be in contact. The authors then sample a population of possible
+structures.
 
-The relationship between genomic distance $s$ and $c$ can be derived from the
-data, by fitting a linear model between the log expected count and the genomic
-distances and extracting the slope coefficient $\alpha$. We thus obtain, up to
-a scaling factor $\beta$ a function to map contact counts to wish distances:
+\citet{paulsen:manifold} formulate a \textit{manifold based optimization},
+where a low rank PSD matrix (and thus a distance matrix) is optimized to be as
+close as possible to the sparse contact count matrix. Applying classical MDS
+on this low rank PSD matrix then yields a 3D model of the genome.
 
-\begin{equation}
-\label{eq:count-to-distance}
-\delta_{ij} = \beta c^{\alpha/3}
-\end{equation}
+\begin{table}[ht!]
+\begin{center}
+\footnotesize
+\begin{tabular}{rlccccc}
+\hline
+\multirow{2}{*}{\emph{\footnotesize Publication}} & \multirow{2}{*}{\emph{\footnotesize Name}} &
+\multirow{2}{*}{\emph{\footnotesize Con.}} &
+\multirow{2}{*}{\emph{\footnotesize Pop.}}
+&\multirow{2}{*}{\emph{\footnotesize MDS-based}} & \emph{\footnotesize Statistical} &
+\multirow{2}{*}{\emph{\footnotesize Availability}} \\
+& & & & & \emph{\footnotesize model} & \\
+\hline
+\hline
+\footnotesize{\cite{dekker:capturing}} & & & & & & \\
+\footnotesize{\cite{duan:three-dimensional}} &  & x & & x & & x \\
+\footnotesize{\cite{tanizawa:mapping}} & & x & & x & & \\
+\footnotesize{\cite{ay:three-dimensional}} & & x & & x & & \\
+\footnotesize{\cite{ben-elazar:spatial}} & & x & & x & & x \\
+\footnotesize{\cite{varoquaux:statistical}} & Pastis & x & & & x & x\\
+\footnotesize{\cite{bau:three-dimensional}} & & & x & & & \\
+\footnotesize{\cite{umbarger:three-dimensional}} & & x & & & &\\
+\footnotesize{\cite{zhang:inference}} & chromSDE & x & & x & & x\\
+\footnotesize{\cite{rousseau:three}} & & & x & & x & x\\
+\footnotesize{\cite{hu:bayesian}} & Bach & x & x & & x & x\\
+\footnotesize{\cite{kalhor:genome}} & & & x & & &\\
+\footnotesize{\cite{lesne:3d}} & ShRec3D & x & & x & & x \\
+\footnotesize{\cite{trieu:large}} & & x & & & & \\
+\footnotesize{\cite{trieu:3D}} & & x & & & & \\
+\footnotesize{\cite{trieu:MOGEN}} & MOGEN & x & & & & \\
+\footnotesize{\cite{nagano:single-cell}} & & x & & & & \\
+\footnotesize{\cite{paulsen:manifold}} & & x & & x & & x \\
+\hline
+\end{tabular}
+\end{center}
+\caption{\bf A comparison of 3D inference methods}{In this table, we summarize
+properties of published methods to infer the 3D structure of the genome: (1)
+is it a consensus or a population based inference? (2) Is it an MDS based
+method? (3) or relies on a statistical modeling; (4) is the software
+available or not (to the best of our knowledge).}
+\end{table}
 
-The scaling factor depends both on the sequencing depth of the libraries, but
-also on the nucleus size of the organims. \citet{ay:three-dimensional} infers
-3D models for a range of parameters and keeps the models that are the most
-plausible, under the assumption that beads should be as uniformely distributed
-in the nucleus as possible. 0-contact counts can either be ignored, or mapped
-to the largest distance possible. Note that for high coverage data set, the
-amount of non-interacting beads is sufficiently small that including those or
-not in the optimization will make very little difference.
+\section*{Modeling DNA as a flexible fibers under constraints}
 
-\subsection{Reproducing the results using a simpler model}
+In the previous section, we discussed how from a contact count map one could
+infer a {\em consensus} model or a population of models that best represented
+the contact map. Another set of methods approach the problem of finding 3D
+models of genome architecture very differently. Instead of using data-driven
+approaches, these methods model chromosomes as series of beads or flexible
+fibers, and attempt to find the smallest set of constraints such that a
+generated contact maps from these models match as closely as possible the
+observed contact maps. As these models refine the constraints, more insight on
+the hallmarks of the genome architecture are discovered.
 
-Two simplifications can be brought to the method proposed by
-\citet{ay:three-dimensional}.
+\subsection*{A small set of constraints explain the 3D architecture of \textit{S. cerevisiae}}
 
-First, the ``count-to-distance'' relationship is based on two steps: (1) find
-the relationship between genomic distances and counts; (2) from the
-coefficient, deduce what type of polymer the DNA of this particular organism
-is closest two (fractal globule or equilibrium); (3) use the relationship
-between genomic distance $s$ and euclidean distance $d$ for that particular
-type of polymer. Yet, it is likely that the probability of contact count
-between two elements depends mostly on the average distance separating the two
-beads. As noted in \citet{varoquaux:statistical}, while the relationship
-between both genomic distance $s$ and expected contact count $c$ and the
-relationship between genomic distance $s$ and euclidean distance $c$,
-combining both yields the same count-to-distance function for both polymer
-type:
-
-\begin{equation}
-\delta_{ij} = \beta c^{-1/3}
-\end{equation}
-
-While the difference may seem subtile, fitting the linear model to find
-$\alpha$ is not as straightforward as it seems, as the range of genomic
-distance used can have a large impact on the final coefficient.
-
-The second simplification is to perform an unconstraint optimization. Indeed,
-in practice, the constraints are fulfilled naturally by the data and thus can
-be dropped from the optimization. In addition of drastically reducing the
-computation time, this allows to use generic publicly available software such
-as chromSDE (which implements a variant of the model defined here), or
-Pastis-MDS (which implements exactly the unconstrained model defined here).
-
-\subsection{Structure stability}
-
-Both the constrained and unconstained problem stated above are not convex. The
-reader may well ask how sensitive the result is to initialize. In short, it is
-not. To study the stability of the results with respect to the initialization,
-I perform for each stage presented here 1000 optimization, starting from
-different initialization.
-
-The first question that arise is simply: are structures from the same time
-points but from a different start ponit more alike than structures from
-different time points? To answer this question, I compute for each structure
-and each time point a feature matrix, composed of a subsampled pairwise
-euclidean distances of each structure. The reason for subsampling is simply so
-that the resulting feature matrix fits in memory on a laptop. I then compute a
-stochastic PCA on this feature matrix and plot the projection in 2D space
-using the 3 first components. The results clearly show that structures from
-the same time point cluster strongly. Repeating this experiment on singly
-chromosomes yields the same result.
-
-The second question one can ask is how do the structures differ. Tackling this
-question is very challenging, but can be reformulated in a much easier way.
-Are the hallmarks of interest conserved across structures of a same stage?
-\citet{ay:three-dimensional} and \citet{lemieux:genome-wide} both identified
-{\em P. falciparum} folded in very specific ways, with VRSSM genes highly
-interacting. \citet{ay:three-dimensional} also observed strong clustering of
-the centromeres, and enrichment in interaction at the telomeres. These
-observation can lead to a rigourous approach at identifying whether families
-of loci clustered in the structures and even quantifying this in a rigourous
-manner.
-
-\section{3D gene set enrichment}
-
-To assess whether groups of genes are colocolized in a 3D model,
-\citet{ay:three-dimensional} leveraged a statistical method developped by
-\citet{witten:XXX}, which requires labeling each pair of loci in two groups:
-"close" or "far".
-
-The authors used varying distance
-thresholds (10\%, 20\% and 40\% of the nuclear diameter) to deem a locus pair
-“close” and labeled all remaining
-pairs in the set as “far”.
-
-\fixme{Re-read Witten and Noble}
-
-This approach dichotomize loci pairs into two groups, and check for the
-enrichment of a label in one of the two groups. \citep{capurso:distance-based}
-presents an approach that avoids this step, and instead directly estimates the
-significance within the 3D model. Briefly, for a group $\mathcal{G}$, MPED
-computes a test statitics:
-\begin{equation*}
-M = \underset{i,j \in \mathcal{G}| c_i != c_j}{\text{median}} d_{ij}\,,
-\end{equation*}
-
-where $d_{ij}$ is the euclidean distance between bead $i$ and bead $j$. The
-null distribution is estimated empirically by resampling $10^5$ with
-preservation of the chromosome structure. If the $M$ statistics is smaller
-than the mean of the null distribution, it is compared to the lower tail of
-the distribution and indicates co-localization. If the $M$ statistics is
-larger than the mean of the null distribution, then it is compared to the
-upper tail of the distribution, and indicates dispersion.
-
-Applying MEPD to the Ring stage, \citet{capurso:distance-based} confirm that
-centromers, telomers, VRSM genes (both overall, subtelomeric and internal)
-colocalize.
-
-\section{Volume exclusion models cannot explain }
+The budding yeast \textit{S. cerivisiae}'s 3D structure has been extensively
+studied, both through 3C-type studies \citep{dekker:capturing,
+duan:three-dimensional, burton:} and through bio-imaging experiments
+\citep{XX:}.
 
@@
 
-$$
-\frac{1}{2} FIXME
-$$
+\begin{equation}
+\frac{1}{2} k_{\text{angle}} \sum^{N - 2}_{i = 1} \left( 1 - \frac{x_{i + 1} -
+x_i}{\|x_{i + 1} - x_i\|} \cdot \frac{x_{i + 2} - x_{i + 1}}{\|x_{i + 2} -
+  x_{i + 1}\|} \right)^2\,,
+\end{equation}
 
-We use IMP to generate a population of structures.
+\citet{tokuda:dynamical}
 
-\paragraph{Adding constraints on VRSM genes}
+\citet{wong:how} model the budding yeast chromosomes a semi-flexible fiber
+constrained in a nucleus, and applies 4 sequences specific forces on this
+fiber to obtain certain properties: (1) centromeres are attached to a single
+point of the nucleus by a segment; (2) telomeres are subjected to an outward
+force, that pushes them towards the nuclear membrane; (3) rDNA is thicken; (4)
+apply a random brownian movement. This model recovers many of the known
+hallmarks of the 3D architecture of {\em S. cerevisiae}.
 
-\section{Integrative analysis of gene expression and 3D structure using KernelCCA}
+\subsection*{\textit{P. falciparum}'s 3D structure cannot be explained by a
+constrained volume exclusion model}
 
-\citet{ay:three-dimensional}  exploited the 3D models to relate structure to
-gene expression. The authors rely on a technic called ``kernelCCA'' with
-attempts to find components on different view of the data in such a way that
-these are coherent with one another.
+Running a similar model on {\em P. falciparum}
 
-- Can we extend this to other data types? Chip-seq data available from our
-review? Can the results be interpretable?
+\section*{Downstream analysis using 3D models: a highlight of the study of
+{\em P. falciparum}'s 3D structure}
 
-\section{Discussion}
-\label{sec4}
+\subsection*{Profiling {\em P. falciparum}'s genome architecture during the
+asexual cycle}
+\label{sec:data}
 
-\section{Software}
-\label{sec5}
+In the past year, two groups have published novel dataset studying {\em P.
+falciparum}'s 3D genome architecture using Hi-C. \citet{lemieux:genome-wide}
+studied several strains associated with populations expression unique
+\textit{var} genes, in order to study key folding properties of the 3D
+structure relating to \textit{var} gene expression. On the other hand,
+\citet{ay:three-dimensional} focus three key time-point of the development
+cycle. Table~\ref{tab:data} summarizes the available data sets and their key
+properties after processing the data as described in
+\citet{ay:three-dimensional} and combining all available libraries for each
+life-cycle stage and strain.
 
-Software in the form of R code, together with a sample
-input data set and complete documentation is available on
-request from the corresponding author (eaheron@tcd.ie).
+\subsection*{Structure stability}
 
-\section{Supplementary Material}
-\label{sec6}
+Both the constrained and unconstained problem stated above are not convex. The
+reader may well ask how sensitive the result is to initialize. In short, it is
+not. To study the stability of the results with respect to the initialization,
+I perform for each stage presented here 1000 optimization, starting from
+different initialization.
 
-Supplementary material is available online at
-% \href{http://biostatistics.oxfordjournals.org}%
-% {http://biostatistics.oxfordjournals.org}.
-\url{http://biostatistics.oxfordjournals.org}.
+The first question that arise is simply: are structures from the same time
+points but from a different start ponit more alike than structures from
+different time points? To answer this question, I compute for each structure
+and each time point a feature matrix, composed of a subsampled pairwise
+euclidean distances of each structure (so that the resulting feature matrix
+fits in memory on a laptop). I then compute a stochastic PCA on this feature
+matrix and plot the projection in 2D space using the 3 first components. The
+results clearly show that structures from the same time point cluster
+strongly. Repeating this experiment on singly chromosomes yields the same
+result.
+
+The second question one can ask is how do the structures differ. Tackling this
+question is very challenging, but can be reformulated in a much easier way.
+Are the hallmarks of interest conserved across structures of a same stage?
+\citet{ay:three-dimensional} and \citet{lemieux:genome-wide} both identified
+{\em P. falciparum} folded in very specific ways, with VRSSM genes highly
+interacting. \citet{ay:three-dimensional} also observed strong clustering of
+the centromeres, and enrichment in interaction at the telomeres. These
+observation can lead to a rigourous approach at identifying whether families
+of loci clustered in the structures and even quantifying this in a rigourous
+manner.
+
+\subsection*{3D gene set enrichment}
+
+To assess whether groups of genes are colocolized in a 3D model,
+\citet{ay:three-dimensional} leveraged a statistical method developped by
+\citet{witten:assessment}, which requires labeling each pair of loci in two groups:
+"close" or "far".
+
+\fixme{This is unclear. Introduce notatinos and rewrite this section}
+
+The authors used varying distance
+thresholds (10\%, 20\% and 40\% of the nuclear diameter) to deem a locus pair
+“close” and labeled all remaining
+pairs in the set as “far”.  The authors then compare the enrichment of loci
+pairs of a group being close and far by resampling loci amongst a same
+chromosome.
+
+This approach dichotomizes loci pairs into two groups, and check for the
+enrichment of a label in one of the two groups. \citep{capurso:distance-based}
+presents an approach that avoids this step, and instead directly estimates the
+significance within the 3D model. Briefly, for a group $\mathcal{G}$, MPED
+computes a test statistics:
+\begin{equation*}
+M = \underset{i,j \in \mathcal{G}| c_i != c_j}{\text{median}} d_{ij}\,,
+\end{equation*}
+
+where $d_{ij}$ is the euclidean distance between bead $i$ and bead $j$. The
+null distribution is estimated empirically by resampling $10^5$ with
+preservation of the chromosome structure. If the $M$ statistics is smaller
+than the mean of the null distribution, it is compared to the lower tail of
+the distribution and indicates co-localization. If the $M$ statistics is
+larger than the mean of the null distribution, then it is compared to the
+upper tail of the distribution, and indicates dispersion.
+
+Applying MEPD to the Ring stage, \citet{capurso:distance-based} confirm that
+centromers, telomers, VRSM genes (both overall, subtelomeric and internal)
+colocalize.
+
+\begin{table}
+\scriptsize
+\centering
+\begin{tabular}{cccccccc}
+\hline
+\textbf{Name} & \textbf{Strain} & \textbf{Stage} & \textbf{Resolution} &
+\textbf{Number of contacts} & \textbf{Perc. of \textit{cis}} & \textbf{Perc of
+\textit{trans}}& \textbf{Reference} \\
+\hline
+\hline
+Ay-\textit{rings} & 3D7 & Late Rings & 10~kb & 16711552 & 43\% & 57\% &  \cite{ay:three-dimensional} \\
+Ay-\textit{trophozoites} & 3D7 & Trophozoites &10~kb & 56348498 & 53\% & 47\% & \cite{ay:three-dimensional} \\
+Ay-\textit{schizonts} & 3D7 & Schizonts & 10~kb & 11652832 & 55\% & 45 \% & \cite{ay:three-dimensional} \\
+Lemieux-\textit{A4+} & IT/BC6+ & Rings & 25~kb & 18488252 & 19\% & 81\% & \cite{lemieux:genome-wide} \\
+Lemieux-\textit{A4} & IT/3G8 & Rings &  25~kb & 19674672 & 28\% & 72\% & \cite{lemieux:genome-wide}\\
+Lemieux-\textit{A44} & IT/BC6- & Rings & 25~kb & 18660594 & 25\% & 75\% & \cite{lemieux:genome-wide}\\
+Lemieux-\textit{DCJ\_On} & NF54/DCJ on & Rings & 25~kb & 3098370 & 26\% & 74\% &\cite{lemieux:genome-wide} \\
+Lemieux-\textit{DCJ\_Off} & NF54/DCF Off & Rings & 25~kb & 2533470 & 26\% & 73\% &\cite{lemieux:genome-wide} \\
+Lemieux-\textit{B15C2} & NF54/B15C2 & Rings & 25~kb &  1022996 & 12\% & 88\% & \cite{lemieux:genome-wide}\\
+\hline
+\end{tabular}
+\caption{Summary of the available {\em P. falciparum} Hi-C datasets}
+\end{table}
+
+\subsection*{Integrative analysis of gene expression and 3D structure using KernelCCA}
 
@@
 
-The authors thank XXX
-
-Funding for the project was provided by BIDS.
 {\it Conflict of Interest}: None declared.

### 2017-07-23 scripts/visualizing_counts/plot_counts.py

@@ -42,6 +42,8 @@ AP2 = AP2.astype(float)
 
 if "25kb" in args.filename:
     resolution = 25000
+elif "20000" in args.filename:
+    resolution = 20000
 else:
     resolution = 10000
 
@@ -62,7 +64,7 @@ if not args.no_normalization:
 c = counts.copy().flatten()
 c.sort()
 vmax = c[int(0.99*len(c))]
-c = c[ c!= 0 ]
+c = c[c != 0]
 vmin = c[int(0.05*len(c))]
 fig, ax = plt.subplots(figsize=(12, 12))
 m = ax.matshow(counts, origin="bottom", norm=colors.SymLogNorm(5),
@@ -114,3 +116,34 @@ try:
     fig.savefig(outname, dpi=400)
 except IOError:
     pass
+
+centromeres = np.loadtxt("files/pf.cent")
+centromeres = centromeres.mean(axis=1) / resolution
+b, e = 0, 0
+for i, l in enumerate(lengths):
+    e += l
+    scounts = counts[b:e, b:e]
+    fig, ax = plt.subplots(figsize=(5, 5))
+    m = ax.matshow(scounts, origin="bottom", norm=colors.SymLogNorm(5),
+               extent=(0, l, 0, l),
+               cmap="Blues")
+    ax.axhline(centromeres[i], linestyle="--", linewidth=1, color="#000000")
+    ax.axvline(centromeres[i], linestyle="--", linewidth=1, color="#000000")
+    ax.set_xticks([])
+    ax.set_yticks([])
+
+    ap2 = AP2[:, 1:].mean(axis=1)[AP2[:, 0] - 1 == i] / resolution
+    [ax.axhline(a, linestyle=":", linewidth=1, color="g") for a in ap2]
+    [ax.axvline(a, linestyle=":", linewidth=1, color="g") for a in ap2]
+
+    try:
+
+        fig.savefig(
+            outname.replace(
+                ".pdf", "_%02d.pdf" % (i+1)).replace(
+                    ".png", "_%02d.png" % (i+1)),
+            dpi=400)
+    except IOError:
+        pass
+
+    b = e

### 2017-07-24 doc/manuscript/paper.tex
Finished the ensemble methods section

@@
 \usepackage{xspace}
+
+\usepackage{booktabs}
+\usepackage[svgnames,table]{xcolor}
+\usepackage[tableposition=above]{caption}
+\usepackage{pifont}
+
+\newcommand*\CHECK{\ding{51}}
 
@@
 
+\todo{A revoir}
+
 In this section, we review ``data-driven'' methods to infer 3D models of DNA
@@
 approaches are more biologically accurate: Hi-C data are derived from a
-population of cells, each of those with a uniquely folded 3D structure. A
-population of structure thus better may represents the variability of
-structures among ours cells. Yet, in addition of being more complex and
+population of cells, each of those with a uniquely folded 3D structure. The
+variability of structures among ours cells may thus be better represented by a
+population of structures. Yet, in addition of being more complex and
 computationally intensive, ensemble approaches raise the question of
@@
 
-After introducing some notations and common terminology, I will dive both in consensus
-and population inference methods.
+After introducing some notations and common terminology, I will dive both in
+consensus and population inference methods.
 
@@
 A typical 3C/Hi-C experiment can be summarized as a contact
-count matrix $\mathbf{C} \in R^{n \times n}$, where each row and column
+count matrix $\mathbf{C} \in \RR^{n \times n}$, where each row and column
 correspond to a genomic-window (or loci), and each entry $c_{ij}$ to the
@@
 to a specific genomic window. We denote by $\textbf{X} \in \RR^{3 \times n}$
-the coordinate matrix of the structure(s) and $x_i \in \RR^3$ corresponds to
-the coordinate matrix of the i-th beads.
+the coordinate matrix of the structure(s), $x_i \in \RR^3$ corresponds to
+the coordinate matrix of the i-th beads and $d_{ij} = \|x_i - x_j\|_2$ the
+euclidean distance separating beads $i$ and $j$.
 
@@
 Early methods \citep{dekker:capturing, duan:three-dimensional,
-tanizawa:mapping, ay:three-dimensional} cast the
-3D inference problem as a multidimensional scaling (MDS) problem: chromosomes
-are models as a chain of beads, to place in 3D such that the distance between
-bead $i$ and $j$ matches as closely as possible a ``wish-distance''
-$\delta_{ij}$ derived
-from contact counts.
+tanizawa:mapping, ay:three-dimensional} cast the 3D inference problem as a
+multidimensional scaling (MDS) problem: chromosomes are models as a chain of
+beads, to place in 3D such that the distance between bead $i$ and $j$ matches
+as closely as possible a ``wish-distance'' $\delta_{ij}$ derived from contact
+counts.
 
@@
 these counts-to-distance functions vary, most of them can be summarized as a
-power-law relationship : $c_{ij} = \beta d_{ij}^\alpha$.
+power-law relationship: $c_{ij} = \beta d_{ij}^\alpha$.
 
 \citet{lesne:3d} proposes a different approach to convert counts into
-distances: they borrow the concept of shortest path from graph theory. A
-graph is constructed from the contact map by considering each loci as a node.
-Loci seen interacting are connected through an edge of weight
-$\frac{1}{c_{ij}}$. The authors then compute wish-distances $\delta_{ij}$ as
-the shortest path between node $i$ and $j$ in the graph. Constructing
-wish-distances in such a way has two advantages: (i) low contact counts do not
-contribute much to the wish-distances; (ii) the resulting wish-distances form
-a distance matrix, and thus an optimal solution to the MDS can be found.
+distances: they borrow the concept of shortest path from graph theory to
+create a distance matrix. A graph is constructed from the contact map by
+considering each loci as a node. Loci seen interacting are connected through
+an edge of weight $\frac{1}{c_{ij}}$. The authors then compute wish-distances
+$\delta_{ij}$ as the shortest path between node $i$ and $j$ in the graph.
+Constructing wish-distances in such a way has two advantages: (i) low contact
+counts do not contribute much to the wish-distances; (ii) the resulting
+wish-distances form a distance matrix, and thus an optimal solution to the MDS
+can be found.
 
@@
 \centering
-\begin{tabular}{cccccc}
+\begin{tabular}{rccccccccc}
 \hline
-\textbf{Paper} & Organism & Full or partial genome & Counts-to-distance mapping & \textbf{Constraints}
-& \textbf{Weights}\\
+\multirow{2}{*}{\textbf{Paper}} & \multirow{2}{*}{\textbf{Organism}} &
+\textbf{Full or}
+& \textbf{Counts-to-distance} &
+\multicolumn{4}{c}{\multirow{2}{*}{\textbf{Constraints}}}
+& \multirow{2}{*}{\textbf{Weights}}\\
+ & & \textbf{partial} & \textbf{mapping} &  \\
+ \cmidrule(lr){5-8} 
+ & & & & Adj. & rDNA & C & T & \\
 \hline
-\cite{dekker:capturing} & {\em S. cerevisiae} & Chr III & Worm-like chain behavior & - \\
+\cite{dekker:capturing} & {\em S. cerevisiae} & Chr III & Worm-like chain
+behavior & & & & \\
 \cite{duan:three-dimensional} & {\em S. cerevisiae} & Whole-genome & Linear
-relationship & Adjacent beads, rDNA & - \\
+relationship & \CHECK &  \CHECK & \CHECK &  & \\
 \cite{tanizawa:mapping} & {\em S. pombe} & Whole-genome &
-FISH-derived relationship & Adjacent beads, rDNA, telomeres, centromeres & - \\
+FISH-derived relationship & \CHECK & \CHECK & \CHECK & \CHECK &  \\
 \cite{ay:three-dimensional} & {\em P. falciparum} & Whole-genome & Fractal
-globule derived relationship & adjacent
-beads & $w_{ij} = \frac{1}{\delta_{ij}^2}$ \\
-\cite{lesne:3d} & - & Whole-genome & \textit{ad hoc} & - & - \
+globule derived relationship & \CHECK  & & &
+ & $\frac{1}{\delta_{ij}^2}$ \\
+\cite{lesne:3d} & & Whole-genome & \textit{ad hoc} & & & & & \
 \end{tabular}
@@
 wish-distances require strong assumptions that may not be met in practice. For
-example, this mapping may change from one organism to another
+example, this mapping changes from one organism to another
 \citep{fudenberg:higher-order}, from one resolution to another
-\citep{zhang:inference}, from one genomic distance range to another
-\citep{ay:statistical}, or from one time point to another during the cell
-cycle \citep{le:high-resolution, ay:three-dimensional}.
+\citep{zhang:inference} or even from one time point to another during the cell
+cycle \citep{le:high-resolution, ay:three-dimensional}. Several methods have
+since been proposed to alleviate this problem.
 
-Two methods have since been proposed to alleviate this problem.
-\citet{ben-elazar:spatial} proposes to cast the inference as a
-\emph{non-metric MDS} \citep{kruskal:multidimensional}, where the 3D structure
-is inferred jointly with the wish-distances. \fixme{except they do a bunch of
-stuff to the count data}.
+\citet{ben-elazar:spatial} casts the inference as a \emph{non-metric MDS}
+\citep{kruskal:multidimensional}, where the 3D structure is inferred jointly
+with the wish-distances. The authors first filter the
+interaction counts to keep only the most significant interactions and
+interpolate the missing values to obtain a smooth, symmetric positive definite
+matrix.
 
@@
 A series of methods cast the inference as an optimization method, where the
-objective function is designed by hand to fullfill a number of properties
-\citep{trieu:large, trieu:3D, trieu:MOGEN}.
+objective function is designed by hand to fulfill a number of properties
+\citep{trieu:large, trieu:MOGEN, trieu:3D}. For example, \citep{trieu:large}
+formulates an optimization problem that attempts to place the beads such that,
+if a significant contact is observed between $i$ and $j$, then $\| x_i - x_j
+\| < d_c$. On the other hand, if a no significant contact is observed, then
+the two beads should be no closer than $d_c$. In addition, the authors also
+place constraints on the minimum distance separating two beads and maximum
+distances between two beads. To satisfy all those properties, the authors
+propose to minimize the following objective function:
 
-For example, \citep{trieu:large} formulates an optimization problem that
-attempts to place the beads such that, if a significant contact is observed
-between $i$ and $j$, then $\| x_i - x_j \| < d_c$. On the other hand, if a
-no significant contact is observed, then the two beads should be no closer
-than $d_c$. In addition, the authors also place constraints on the minimum
-distance separating two beads and maximum distances between two beads. To
-satisfy all those properties, the authors propose to minimize the following
-objective function:
+\begin{equation*}
+\begin{aligned}
+\mathcal{O}(X, \delta_{ij}) = & \underset{i, j | c_{ij} \neq 0, |i-j| \neq 1}{\sum} \Bigg(W_1 \text{tanh}(d^2_c -
+d^2_{ij}) c_{ij} + W_2 \frac{\text{tanh}(d^2_{ij} - d^2_\text{min})}{N} \Bigg) \\
+& \underset{i, j | c_{ij} = 0}{\sum} \Bigg(W_3 \frac{\text{tanh}(d^2_\text{max} -
+d^2_{ij})}{N} + W_4 \frac{\text{tanh}(d^2_{ij} -
+d^2_{c})}{N}\Bigg) \\
+& \underset{i, j | |i - j| = 1}{\sum} \Bigg(W_1 \frac{\text{tanh}(d^2_\text{amax} -
+d^2_{ij})}{N} + W_2 \frac{\text{tanh}(d^2_{ij} -
+d^2_\text{min})}{N}\Bigg) \,,\\
+\end{aligned}
+\end{equation*}
 
-\fixme{add damn objective function}
+where $N$ is the total number of interaction, $(W_1, W_2, W_3, W_4)$ weights
+associated to each term, $d_\text{amax}$ the maximum distance separating two
+adjacent beads, $d_\text{min}$ the minimum distance separating two adjacent
+beads and $d_\text{max}$ the maximum distance between two beads. 
+This method is then generalized to include \textit{trans} contact counts data
+\citep{trieu:MOGEN}
 
-This method is then generalized to include \textit{trans} contact counts data
-\citep{trieu:MOGEN}.
-
-\cite{trieu:MOGEN} proposes to exploit the Lozentzian function, to formulate a
-constraint-based problem. They first convert the contact counts $c_{ij}$ into
-wish-distances, and cast the following optimization problem:
+To give another example of carefully crafted optimization, \cite{trieu:3D}
+proposes to exploit the Lozentzian function, to formulate a constraint-based
+problem. They first convert the contact counts $c_{ij}$ into wish-distances,
+and cast the following optimization problem:
 
@@
 
-\citet{varoquaux:statistical} proposes a novel approach based on
-statistical modeling of contact counts as random Poisson variables where the
-intensity is a function of the distance: $c_{ij} \sim \text{Poisson}(\beta
-d_{ij}^\alpha)$, and some perform the optimization only including a subset of
-contact counts. The authors propose to cast the
+\citet{varoquaux:statistical} proposes a novel approach called \texttt{Pastis}
+based on statistical modeling of contact counts as random Poisson variables
+where the intensity is a function of the distance: $c_{ij} \sim
+\text{Poisson}(\beta d_{ij}^\alpha)$, and some perform the optimization only
+including a subset of contact counts. The authors propose to cast the
 inference of 3D models as maximizing the likelihood:
@@
 
+\texttt{Pastis} can automatically adjust the parameters $\beta$, $\alpha$ of
+the counts-to-distance transfer function and infer a genome structure that
+best explains the observed data.
+
 \subsection*{Ensemble methods}
 
-\citet{bau:three-dimensional}
-\cite{umbarger:three-dimensional}
-\cite{hu:bayesian}
+{\em Ensemble approaches} aim at inferring a population of structures
+representative of the contact count map. The methods fall into two distinct
+categories: the first type of problem casts a non-convex or ill-defined
+optimization problem and sample local minima of the function
+\citep{bau:three-dimensional, umbarger:three-dimensional}, while the second
+type proposes a statistical modeling of the problem and samples the posterior
+distribution \citep{rousseau:three, hu:bayesian}.A parallel can be made
+between the first group of methods and {\em Consensus} MDS type of approaches,
+and the second with {\em Consensus} statistical based methods.
+
+\subsubsection*{Sampling local minima}
+
+\citet{umbarger:three-dimensional,bau:three-dimensional, kalhor:genome} models
+chromosomes as a series of beads, linked by restraining oscillators. These
+oscillators can be thought of as a ``force'' between beads so that they come
+into contact or ensure a minimal or maximal distance between those. The models
+includes three differente types of restraints: (i) beads seen interacting are
+restrained with harmonic oscillators of strengths derived from the contact
+counts; (ii) adjacent beads are ensured to be neither too close nor too far
+from one another; (iii) the last oscillators are activated only when
+constraints are not fulfilled to XXX.  This yields an optimization problem
+with a large number of local minima, which the authors sample from by running
+50,000 minimizations starting from random starts.
+
+\subsubsection*{Estimating the posterior distribution}
+
+Similarly to \citet{varoquaux:statistical}, \citet{rousseau:three} and
+\citet{hu:bayesian} proposes to model contact counts with a formal
+probabilistic model. \citet{rousseau:three} models observed contact counts
+$c_{ij}$ as a random Gaussian variable of mean $\beta d_{ij}^{\alpha}$,
+$\alpha \leq 0$ and variance $\sigma_{ij}$ estimated directly from the contact
+count data, while \citet{hu:bayesian} models contact counts as random Poisson
+variables of mean $\beta d_{ij}^\alpha$. The authors then sample from the
+posterior using MCMC.
 
@@
 genome rely on a new protocol to probe single-cell for their 3D structures
-\citep{nagano:single-cell, XXX}. Single-cell Hi-C is still in its early days,
+\citep{nagano:single-cell,ramani:massively}. Single-cell Hi-C is still in its early days,
 and only a handful of data sets are today publicly available. The contact maps
@@
 
-\citet{nagano:single-cell} models the chromosome as a series of beads linked
-as a flexible fiber. Loci seen interacting in the single-cell contact maps are
-constrained to be in contact. The authors then sample a population of possible
-structures.
+Akin to {\em consensus} MDS-like method and {\em ensemble}-local minima
+methods,  the first approach is to consider each contact as a constraint and
+to formulate
+an under constrained optimization problem. A population of structures
+satistifying the constraints can be found by sampling local minimas.
 
-\citet{paulsen:manifold} formulate a \textit{manifold based optimization},
-where a low rank PSD matrix (and thus a distance matrix) is optimized to be as
-close as possible to the sparse contact count matrix. Applying classical MDS
-on this low rank PSD matrix then yields a 3D model of the genome.
+\citet{paulsen:manifold} proposes a very different method, and formulate a
+\textit{manifold based optimization}, where a low rank PSD matrix (and thus a
+distance matrix) is optimized to be as close as possible to the sparse contact
+count matrix. Applying classical MDS on this low rank PSD matrix then yields a
+3D model of the genome.
 
@@
 \hline
-\multirow{2}{*}{\emph{\footnotesize Publication}} & \multirow{2}{*}{\emph{\footnotesize Name}} &
-\multirow{2}{*}{\emph{\footnotesize Con.}} &
-\multirow{2}{*}{\emph{\footnotesize Pop.}}
-&\multirow{2}{*}{\emph{\footnotesize MDS-based}} & \emph{\footnotesize Statistical} &
-\multirow{2}{*}{\emph{\footnotesize Availability}} \\
-& & & & & \emph{\footnotesize model} & \\
+\multirow{2}{*}{\textbf{\footnotesize Publication}} & \multirow{2}{*}{\textbf{\footnotesize Name}} &
+\textbf{\footnotesize Consensus or} 
+&\multirow{2}{*}{\textbf{\footnotesize MDS-based}} & \textbf{\footnotesize Statistical} &
+\multirow{2}{*}{\textbf{\footnotesize Available}} \\
+& & \textbf{\footnotesize Ensemble}
+ & & \textbf{\footnotesize model} & \\
 \hline
 \hline
-\footnotesize{\cite{dekker:capturing}} & & & & & & \\
-\footnotesize{\cite{duan:three-dimensional}} &  & x & & x & & x \\
-\footnotesize{\cite{tanizawa:mapping}} & & x & & x & & \\
-\footnotesize{\cite{ay:three-dimensional}} & & x & & x & & \\
-\footnotesize{\cite{ben-elazar:spatial}} & & x & & x & & x \\
-\footnotesize{\cite{varoquaux:statistical}} & Pastis & x & & & x & x\\
-\footnotesize{\cite{bau:three-dimensional}} & & & x & & & \\
-\footnotesize{\cite{umbarger:three-dimensional}} & & x & & & &\\
-\footnotesize{\cite{zhang:inference}} & chromSDE & x & & x & & x\\
-\footnotesize{\cite{rousseau:three}} & & & x & & x & x\\
-\footnotesize{\cite{hu:bayesian}} & Bach & x & x & & x & x\\
-\footnotesize{\cite{kalhor:genome}} & & & x & & &\\
-\footnotesize{\cite{lesne:3d}} & ShRec3D & x & & x & & x \\
-\footnotesize{\cite{trieu:large}} & & x & & & & \\
-\footnotesize{\cite{trieu:3D}} & & x & & & & \\
-\footnotesize{\cite{trieu:MOGEN}} & MOGEN & x & & & & \\
-\footnotesize{\cite{nagano:single-cell}} & & x & & & & \\
-\footnotesize{\cite{paulsen:manifold}} & & x & & x & & x \\
+\footnotesize{\cite{dekker:capturing}} & & C & & & \\
+\footnotesize{\cite{duan:three-dimensional}} &  & C & \CHECK & & \CHECK \\
+\footnotesize{\cite{tanizawa:mapping}} & & C & \CHECK & & \\
+\footnotesize{\cite{ay:three-dimensional}} & & C & \CHECK & & \\
+\footnotesize{\cite{ben-elazar:spatial}} & & C & \CHECK & & \CHECK \\
+\footnotesize{\cite{varoquaux:statistical}} & Pastis & C & & \CHECK & \CHECK\\
+\footnotesize{\cite{bau:three-dimensional}} & & E & & &  \\
+\footnotesize{\cite{umbarger:three-dimensional}} & & E  & & &\\
+\footnotesize{\cite{zhang:inference}} & chromSDE & C &  \CHECK & & \CHECK\\
+\footnotesize{\cite{rousseau:three}} & & E & & \CHECK & \CHECK\\
+\footnotesize{\cite{hu:bayesian}} & Bach & E/C &  & \CHECK & \CHECK\\
+\footnotesize{\cite{kalhor:genome}} & & E &   & &\\
+\footnotesize{\cite{lesne:3d}} & ShRec3D & C  & \CHECK & & \CHECK \\
+\footnotesize{\cite{trieu:large}} & & C & & & \\
+\footnotesize{\cite{trieu:3D}} & & C & &  & \\
+\footnotesize{\cite{trieu:MOGEN}} & MOGEN & C & & & \\
+\footnotesize{\cite{nagano:single-cell}} & & E & & & \\
+\footnotesize{\cite{paulsen:manifold}} & & C & \CHECK & & \CHECK \\
 \hline
@@
 properties of published methods to infer the 3D structure of the genome: (1)
-is it a consensus or a population based inference? (2) Is it an MDS based
+is it a consensus or a ensemble based inference? (2) Is it an MDS based
 method? (3) or relies on a statistical modeling; (4) is the software
@@
 duan:three-dimensional, burton:} and through bio-imaging experiments
-\citep{XX:}.
+\citep{XX}. The small size of its genome, the well-known hallmarks of its
+genome architecture and the availability of high resolution contact maps and
+FISH data set quickly led several teams to investigate the minimal set of
+constraints needed to reproduce the hallmarks of its genome architecture.
 
-\citet{tjong:physical} showed \textit{S. cerevisiae}'s structure can be mostly
-explained by a large set of constrained yet random population of structures
-following a simple volume exclusion model. We repeated the experiment on {\em
-P. falciparum}, and showed that, despite recovering similar results on {\em S.
-cerevisiae}, the same set of constraints is not sufficient to explain {\em P.
-falciparum}'s Hi-C contact map.
+\cite{tjong:physical, tokuda:dynamical, tjong:physical} model {\em S.
+cerevisiae}'s chromosomes as flexible random fibers under a small set of
+constraints. While the exact modeling proposed by the three groups differ, the
+set of constraints are similar: \textit{(i)} the chromosomes are constrained
+into a spherical ball representing the nucleus; \textit{(ii)} centromeres are
+constrained into a spherical ball tethered to the nuclear membrane;
+\textit{(iii)} telomeres are tethered to the nuclear membraine; \textit{(iv)}
+rDNA is constrained into the nucleolus, represented as a spherical ball
+opposite to the centromeres.
 
-Briefly, let us recall the methodology presented by \citet{tjong:physical}. We
-construct a population of three-dimensional structures by modeling chromosomes
-as random configuration of beads-on-a-string, subjected to the following constraints.
+Let us take a closer look at the model proposed by \citet{tjong:physical}.
+The 16 chromosomes are modeled as beads-on-a-string spaced 3.2kb apart. The
+authors cast an optimization problem including the following restraints:
 
 \begin{itemize}
-\item \textbf{Adjacent Beads Constraints} Two adjacent beads spaced 3.2kb apart are constrained to be 30nm apatr.
-\item \textbf{Volume Exclusion Constraints} Each pairs of beads are spaced at minimum XX apart
-\item \textbf{Nucleus Constraint} All beads lie in a spherical nucleus of specific
-radius (XXX)
+\item \textbf{Adjacent Beads Constraints} Two adjacent beads spaced 3.2kb
+apart are constrained to be 30~nm apart;
+\item \textbf{Volume Exclusion Constraints} Each pairs of beads are spaced at
+minimum $30~nm$ apart;
+\item \textbf{Nucleus Constraint} All beads lie in a spherical nucleus of
+radius $1\mu m$
 \item \textbf{Centromeres Constraints} Centromeres are localized in a small sphere of
@@
 \item \textbf{Telomere Constraints} Telomeres are located at most 50nm from the nuclear envelop.
-\end{itemize}
-
-\citet{tjong:physical} then formulates an optimization problem that includes
-in addition to all the aforementioned constraints a penalty term on three
-adjacent beads accounting for chromatin stiffness.
-
+\item \textbf{rDNA restraint} The repeated rDNA located on chromosome 12 is
+constrained to be inside the nucleolus;
+\item \textbf{Nucleolus exclusion restraint} All beads that isn't rDNA is
+constrained outside of the nucleolus;
+\item \textbf{Chain stiffness}: an angular restraint is imposed on three
+consecutive beads to reproduce chain stiffness:
 \begin{equation}
@@
 \end{equation}
+\end{itemize}
 
-\citet{tokuda:dynamical}
+In the end, the objective function minimized is the following:
 
-\citet{wong:how} model the budding yeast chromosomes a semi-flexible fiber
-constrained in a nucleus, and applies 4 sequences specific forces on this
-fiber to obtain certain properties: (1) centromeres are attached to a single
-point of the nucleus by a segment; (2) telomeres are subjected to an outward
-force, that pushes them towards the nuclear membrane; (3) rDNA is thicken; (4)
-apply a random brownian movement. This model recovers many of the known
-hallmarks of the 3D architecture of {\em S. cerevisiae}.
+\begin{equation}
+\end{equation}
+
+Using IMP, the authors generate a 25,000 structures, all of which respect the
+constraints listed above. The population of structures is then used to created
+a ``volume-exclusion contact map'', considering that two beads that are less
+then 45~nm apart from a contact. The Pearson correlation of the
+volume-exclusion contact map and the Hi-C one are highly correlated,
+demonstrating this small set of constraints fully explain the observed counts.
+In addition, the population of structures also explain FISH experiments
+previously published.
 

### 2017-07-26 doc/manuscript/paper.tex
More on the paper

@@
 %% Use natbib with these recommended options
-\usepackage[authoryear,comma,longnamesfirst,sectionbib]{natbib}
+%\usepackage[authoryear,comma,longnamesfirst,sectionbib]{natbib} 
+\usepackage{natbib}
 
@@
 measures of mortality and the disease is thought to cost more than US\$21
-billion a year \citep{onwejkwe:do}.
-
+billion a year \citep{onwujekwe:do}.
 Malaria is caused by small unicellular protozoan {\em Plasmodium}, infecting
@@
 malaria continues to arise, genomic-based research on malaria seeks new
-avenues for developing novel therapies \citet{kirchner:recent}. One of the
+avenues for developing novel therapies \citep{kirchner:recent}. One of the
 limiting factors of developments of new drugs is our poor understanding of the
@@
 be driven by coordinated changes in gene expression, the relative paucity of
-transcription factors points to unusual gene regulatory mechanisms . Meanwhile,
+transcription factors points to unusual gene regulatory mechanisms. Meanwhile,
 the relative abundance of proteins related to chromatin structures, mRNA decay
@@
 processes. The 3D genome of a wide variety of organisms has been studied in
-the past year, including many species of yeasts \citet{duan:three-dimensional,
-burton:xx}, bacteria, flies, plants and numerous human and mouse cell lines.
+the past year, including many species of yeasts \citep{duan:three-dimensional,
+burton:species-level, mizuguchi:cohesin-dependant}, bacteria
+\citep{umbarger:three-dimensional}, flies \citep{sexton:three-dimensional},
+plants \citep{feng:genome-wide, wang:genome-wide} and numerous human and mouse
+cell lines \citep{lieberman-aiden:comprehensive, rao:3D}.
+
 Within the past few years, two studies of the three-dimensional structure of
@@
 involved the invasion of red blood cell and also responsible for the
-parasite's great capacity to evade immuno-therapy.
+parasite's great capacity to evade our immune system.
 \citet{ay:three-dimensional} assayed the 3D structure of the genome of the
@@
 
-\todo{A revoir}
+The development of genowe-wide and high-throughput protocols to probe samples
+for their 3D genome architecture naturally paved the way to systematic and in
+depth study of the folding mechanisms of DNA.
 
-In this section, we review ``data-driven'' methods to infer 3D models of DNA
-from Hi-C data. These methods broadly fall into two categories: (i)
-\textit{consensus} approaches, that aim at inferring a unique mean structures
-best representing the contact count data; (ii) \textit{ensemble} methods that
-yield a population of structures.
+Over the past 10 years, the challenge of building 3D models from contact maps
+have been tackled by a plethora of methods. These model chromosomes as a
+series of beads, and attempt to place the beads in a 3D euclidean space to
+well represent the contact map. ``data-driven'' methods broadly fall into two
+categories: (i) \textit{consensus} approaches, that aim at inferring a unique
+mean structures best representing the contact count data; (ii)
+\textit{ensemble} methods that yield a population of structures.
 
@@
 
-After introducing some notations and common terminology, I will dive both in
-consensus and population inference methods.
-
-\subsection*{Notation}
-
-A typical 3C/Hi-C experiment can be summarized as a contact
-count matrix $\mathbf{C} \in \RR^{n \times n}$, where each row and column
-correspond to a genomic-window (or loci), and each entry $c_{ij}$ to the
-number of times loci $i$ and $j$ have been observed interacting. Contact count
-matrices are subjected to biological and technical biases that need to be
-corrected \citep{imakaev:iterative, cournac:normalization, yaffe:probabilistic,
-hu:hicnorm}.
-
-Methods model chromosomes as a series of beads in 3D, each bead corresponding
-to a specific genomic window. We denote by $\textbf{X} \in \RR^{3 \times n}$
-the coordinate matrix of the structure(s), $x_i \in \RR^3$ corresponds to
-the coordinate matrix of the i-th beads and $d_{ij} = \|x_i - x_j\|_2$ the
-euclidean distance separating beads $i$ and $j$.
-
 \subsection*{Consensus models}
 
-Consensus methods aim at inferring a model that best represents the contact
-count matrix. They model each chromosome by a chain of beads, and attempt to
-place the beads in a 3D space to minimize an objective function
-$\mathcal{O}(\mathbf{O}, \mathbf{C})$, sometimes
-under constraints.
+Consensus methods aim at inferring a model $\mathbf{X} \in \RR^{3 \times n}$
+that best represents the contact count matrix $\mathbf{X} \in \RR^{3 \times
+n}$, usually through the minimization of an objective function
+$\mathcal{O}(\mathbf{O}, \mathbf{C})$, sometimes under constraints.
 
@@
 \begin{array}{ccl}
-\underset{\Xb}{\text{minimize}} & & \mathcal{O}(\mathbf{C}, \Xb)
+\underset{\Xb}{\text{minimize}} & & \mathcal{O}(\mathbf{C}, \Xb) \,,
 \end{array}
@@
 
-Early methods \citep{dekker:capturing, duan:three-dimensional,
-tanizawa:mapping, ay:three-dimensional} cast the 3D inference problem as a
+Early methods cast the 3D inference problem as a
 multidimensional scaling (MDS) problem: chromosomes are models as a chain of
@@
 as closely as possible a ``wish-distance'' $\delta_{ij}$ derived from contact
-counts.
+counts \citep{dekker:capturing, duan:three-dimensional,
+tanizawa:mapping, ay:three-dimensional} .
 
@@
 count-to-distance mapping by fitting the relationship using known pairwise
-distances obtained through high-resolution FISH measures. While parameters of
-these counts-to-distance functions vary, most of them can be summarized as a
-power-law relationship: $c_{ij} = \beta d_{ij}^\alpha$.
+distances obtained through high-resolution FISH measures.
 
@@
 \citep{zhang:inference} or even from one time point to another during the cell
-cycle \citep{le:high-resolution, ay:three-dimensional}. Several methods have
-since been proposed to alleviate this problem.
-
-\citet{ben-elazar:spatial} casts the inference as a \emph{non-metric MDS}
-\citep{kruskal:multidimensional}, where the 3D structure is inferred jointly
-with the wish-distances. The authors first filter the
-interaction counts to keep only the most significant interactions and
-interpolate the missing values to obtain a smooth, symmetric positive definite
-matrix.
-
-\citet{zhang:inference} proposes to parametrize the count-to-distance mapping
-as a power-law ($d = \beta c^\alpha)$, and to infer jointly the parameters
-jointly with the structure. In addition, the authors propose to add a
-penalization term for non-interacting beads. The proposed penalization allows
-to cast the problem as a convex optimization and thus guarantees to retrieve
-the optimal structure.
+cycle \citep{le:high-resolution, ay:three-dimensional}. To alleviate this
+problem, after filtering interaction counts based on significance and
+interpolation of missing values, \citet{ben-elazar:spatial} casts the
+inference as a \emph{non-metric MDS} \citep{kruskal:multidimensional}, where
+the 3D structure is inferred jointly with the wish-distances. Another idea is
+to parametrize the count-to-distance mapping as a power-law ($d = \beta
+c^\alpha)$, and to infer jointly the parameters jointly with the structure
+\citet{zhang:inference}.
 
@@
 
+\fixme{maybe skip this section}
+
 \subsubsection*{Statistical-based models}
@@
 the counts-to-distance transfer function and infer a genome structure that
-best explains the observed data.
+best explains the observed data. The strength of this methods comes from the
+robustness to low signal-to-noise ratio.
+
+\fixme{add results on the P. falciparum to show pastis's results on 
+super noisy data}
 
@@
 representative of the contact count map. The methods fall into two distinct
-categories: the first type of problem casts a non-convex or ill-defined
+categories: the first type of problem casts a restraint-based 
 optimization problem and sample local minima of the function
@@
 type proposes a statistical modeling of the problem and samples the posterior
-distribution \citep{rousseau:three, hu:bayesian}.A parallel can be made
+distribution \citep{rousseau:three, hu:bayesian}. A parallel can be made
 between the first group of methods and {\em Consensus} MDS type of approaches,
@@
 variables of mean $\beta d_{ij}^\alpha$. The authors then sample from the
-posterior using MCMC.
+posterior using MCMC. Obtaining a consensus structure from such a method can
+be accomplished by selecting the maximum a posteriori.
 
@@
 \begin{center}
-\footnotesize
+\scriptsize
 \begin{tabular}{rlccccc}
@@
 \hline
-\footnotesize{\cite{dekker:capturing}} & & C & & & \\
-\footnotesize{\cite{duan:three-dimensional}} &  & C & \CHECK & & \CHECK \\
-\footnotesize{\cite{tanizawa:mapping}} & & C & \CHECK & & \\
-\footnotesize{\cite{ay:three-dimensional}} & & C & \CHECK & & \\
-\footnotesize{\cite{ben-elazar:spatial}} & & C & \CHECK & & \CHECK \\
-\footnotesize{\cite{varoquaux:statistical}} & Pastis & C & & \CHECK & \CHECK\\
-\footnotesize{\cite{bau:three-dimensional}} & & E & & &  \\
-\footnotesize{\cite{umbarger:three-dimensional}} & & E  & & &\\
-\footnotesize{\cite{zhang:inference}} & chromSDE & C &  \CHECK & & \CHECK\\
-\footnotesize{\cite{rousseau:three}} & & E & & \CHECK & \CHECK\\
-\footnotesize{\cite{hu:bayesian}} & Bach & E/C &  & \CHECK & \CHECK\\
-\footnotesize{\cite{kalhor:genome}} & & E &   & &\\
-\footnotesize{\cite{lesne:3d}} & ShRec3D & C  & \CHECK & & \CHECK \\
-\footnotesize{\cite{trieu:large}} & & C & & & \\
-\footnotesize{\cite{trieu:3D}} & & C & &  & \\
-\footnotesize{\cite{trieu:MOGEN}} & MOGEN & C & & & \\
-\footnotesize{\cite{nagano:single-cell}} & & E & & & \\
-\footnotesize{\cite{paulsen:manifold}} & & C & \CHECK & & \CHECK \\
+\scriptsize{\cite{dekker:capturing}} & & C & \CHECK & & \\
+\scriptsize{\cite{duan:three-dimensional}} &  & C & \CHECK & & \CHECK \\
+\scriptsize{\cite{tanizawa:mapping}} & & C & \CHECK & & \\
+\scriptsize{\cite{ay:three-dimensional}} & & C & \CHECK & & \\
+\scriptsize{\cite{ben-elazar:spatial}} & & C & \CHECK & & \CHECK \\
+\scriptsize{\cite{varoquaux:statistical}} & Pastis & C & & \CHECK & \CHECK\\
+\scriptsize{\cite{bau:three-dimensional}} & & E & & &  \\
+\scriptsize{\cite{umbarger:three-dimensional}} & & E  & & &\\
+\scriptsize{\cite{zhang:inference}} & chromSDE & C &  \CHECK & & \CHECK\\
+\scriptsize{\cite{rousseau:three}} & & E & & \CHECK & \CHECK\\
+\scriptsize{\cite{hu:bayesian}} & Bach & E/C &  & \CHECK & \CHECK\\
+\scriptsize{\cite{kalhor:genome}} & & E &   & &\\
+\scriptsize{\cite{lesne:3d}} & ShRec3D & C  & \CHECK & & \CHECK \\
+\scriptsize{\cite{trieu:large}} & & C & & & \\
+\scriptsize{\cite{trieu:3D}} & & C & &  & \\
+\scriptsize{\cite{trieu:MOGEN}} & MOGEN & C & & & \\
+\scriptsize{\cite{nagano:single-cell}} & & E & & & \\
+\scriptsize{\cite{paulsen:manifold}} & & C & \CHECK & & \CHECK \\
 \hline
@@
 
-\section*{Modeling DNA as a flexible fibers under constraints}
+\section*{Model-driven approach: modeling DNA as a flexible fibers under constraints}
 
@@
 studied, both through 3C-type studies \citep{dekker:capturing,
-duan:three-dimensional, burton:} and through bio-imaging experiments
-\citep{XX}. The small size of its genome, the well-known hallmarks of its
+duan:three-dimensional, burton:species-level} and through bio-imaging experiments
+\citep{berger:high}. The small size of its genome, the well-known hallmarks of its
 genome architecture and the availability of high resolution contact maps and
@@
 constraints. While the exact modeling proposed by the three groups differ, the
-set of constraints are similar: \textit{(i)} the chromosomes are constrained
-into a spherical ball representing the nucleus; \textit{(ii)} centromeres are
-constrained into a spherical ball tethered to the nuclear membrane;
-\textit{(iii)} telomeres are tethered to the nuclear membraine; \textit{(iv)}
-rDNA is constrained into the nucleolus, represented as a spherical ball
-opposite to the centromeres.
+set of constraints can roughly be summarized as: \textit{(i)} the chromosomes
+are constrained into a spherical ball representing the nucleus; \textit{(ii)}
+centromeres are constrained into a spherical ball tethered to the nuclear
+membrane; \textit{(iii)} telomeres are tethered to the nuclear membraine;
+\textit{(iv)} rDNA is constrained into the nucleolus, represented as a
+spherical ball opposite to the centromeres.
 
@@
 
-In the end, the objective function minimized is the following:
-
-\begin{equation}
-\end{equation}
-
 Using IMP, the authors generate a 25,000 structures, all of which respect the
@@
 
-\subsection*{\textit{P. falciparum}'s 3D structure cannot be explained by a
-constrained volume exclusion model}
-
-Running a similar model on {\em P. falciparum}
-
 \section*{Downstream analysis using 3D models: a highlight of the study of
@@
 
-\subsection*{Profiling {\em P. falciparum}'s genome architecture during the
-asexual cycle}
-\label{sec:data}
+In previous section, I have reviewed data-driven methods to infer either
+\textit{consensus} or \textit{ensemble} models of the 3D structure. The reader
+may well ask why go through the effort to obtain such model and not directly
+study the contact maps?  In this section, I will review a number of downstream
+analysis one can do on 3D models, highlighting but limiting myself to results
+on the 3D structure of {\em P. falciparum}.
 
-In the past year, two groups have published novel dataset studying {\em P.
+\subsection{Probing the 3D structure of the {\em P. falciparum}}
+In the past year, two groups have published studies of {\em P.
 falciparum}'s 3D genome architecture using Hi-C. \citet{lemieux:genome-wide}
@@
 
-\subsection*{Structure stability}
+\begin{table}
+\scriptsize
+\centering
+\begin{tabular}{cccccccc}
+\hline
+\multirow{2}{*}{\textbf{Name}} & \multirow{2}{*}{\textbf{Strain}} &
+\multirow{2}{*}{\textbf{Stage}} & \multirow{2}{*}{\textbf{Resolution}} &
+\textbf{Number} & \textbf{Perc} & \textbf{Perc}& \multirow{2}{*}{\textbf{Reference}} \\
+& & & & \textbf{of contacts} & \textbf{of \textit{cis}}& \textbf{of trans}& \\
+\hline
+\hline
+Ay-\textit{rings} & 3D7 & Late Rings & 10~kb & 16711552 & 43\% & 57\% &  \cite{ay:three-dimensional} \\
+Ay-\textit{trophozoites} & 3D7 & Trophozoites &10~kb & 56348498 & 53\% & 47\% & \cite{ay:three-dimensional} \\
+Ay-\textit{schizonts} & 3D7 & Schizonts & 10~kb & 11652832 & 55\% & 45 \% & \cite{ay:three-dimensional} \\
+Lemieux-\textit{A4+} & IT/BC6+ & Rings & 25~kb & 18488252 & 19\% & 81\% & \cite{lemieux:genome-wide} \\
+Lemieux-\textit{A4} & IT/3G8 & Rings &  25~kb & 19674672 & 28\% & 72\% & \cite{lemieux:genome-wide}\\
+Lemieux-\textit{A44} & IT/BC6- & Rings & 25~kb & 18660594 & 25\% & 75\% & \cite{lemieux:genome-wide}\\
+Lemieux-\textit{DCJ\_On} & NF54/DCJ on & Rings & 25~kb & 3098370 & 26\% & 74\% &\cite{lemieux:genome-wide} \\
+Lemieux-\textit{DCJ\_Off} & NF54/DCF Off & Rings & 25~kb & 2533470 & 26\% & 73\% &\cite{lemieux:genome-wide} \\
+Lemieux-\textit{B15C2} & NF54/B15C2 & Rings & 25~kb &  1022996 & 12\% & 88\% & \cite{lemieux:genome-wide}\\
+\hline
+\end{tabular}
+\caption{Summary of the available {\em P. falciparum} Hi-C datasets}
+\end{table}
 
-Both the constrained and unconstained problem stated above are not convex. The
-reader may well ask how sensitive the result is to initialize. In short, it is
-not. To study the stability of the results with respect to the initialization,
-I perform for each stage presented here 1000 optimization, starting from
-different initialization.
+\fixme{Add the fact we have time points}
+
+\subsection*{Model evaluation}
+
+A substantial difficulty in modelling the 3D structure of the genome is that
+model evaluation tends to the subjective. What is the relevant measure?
+``Truth'' is generally not fully available, except for a few pairs of loci or
+in simulations. Is validating the colocalization of a pair or a few pairs of
+loci via FISH experiments enough? Is fit to the contact maps or agreements
+between modeling techniques relevant?
+
+\fixme{Add section on model evaluation}
+
+\subsection*{Structure stability across time points, clustering and other
+variance analysis \dots}
+
+The reader may well ask how sensitive the result is either initialization or
+how. In short, it is not. To study the stability of the results with respect
+to the initialization, I perform for each stage presented here 1000
+optimization, starting from different initialization.
 
 The first question that arise is simply: are structures from the same time
-points but from a different start ponit more alike than structures from
+points but from a different start point more alike than structures from
 different time points? To answer this question, I compute for each structure
@@
 the centromeres, and enrichment in interaction at the telomeres. These
-observation can lead to a rigourous approach at identifying whether families
+observation can lead to a rigorous approach at identifying whether families
 of loci clustered in the structures and even quantifying this in a rigourous
 manner.
+
+\subsection{Chromatin compaction}
+
+Bau et al
 
@@
 To assess whether groups of genes are colocolized in a 3D model,
-\citet{ay:three-dimensional} leveraged a statistical method developped by
-\citet{witten:assessment}, which requires labeling each pair of loci in two groups:
-"close" or "far".
-
-\fixme{This is unclear. Introduce notatinos and rewrite this section}
-
-The authors used varying distance
-thresholds (10\%, 20\% and 40\% of the nuclear diameter) to deem a locus pair
-“close” and labeled all remaining
-pairs in the set as “far”.  The authors then compare the enrichment of loci
-pairs of a group being close and far by resampling loci amongst a same
-chromosome.
+\citet{ay:three-dimensional} leveraged a statistical method developed by
+\citet{witten:assessment}, which requires labeling each pair of loci in two
+groups: "close" or "far".The authors used varying distance thresholds (10\%,
+20\% and 40\% of the nuclear diameter) to deem a locus pair “close” and
+labeled all remaining pairs in the set as “far”.  The authors then compare the
+enrichment of loci pairs of a group being close and far by resampling loci
+among a same chromosome.
 
@@
 Applying MEPD to the Ring stage, \citet{capurso:distance-based} confirm that
-centromers, telomers, VRSM genes (both overall, subtelomeric and internal)
+centromeres, telomeres, VRSM genes (both overall, subtelomeric and internal)
 colocalize.
 
-\begin{table}
-\scriptsize
-\centering
-\begin{tabular}{cccccccc}
-\hline
-\textbf{Name} & \textbf{Strain} & \textbf{Stage} & \textbf{Resolution} &
-\textbf{Number of contacts} & \textbf{Perc. of \textit{cis}} & \textbf{Perc of
-\textit{trans}}& \textbf{Reference} \\
-\hline
-\hline
-Ay-\textit{rings} & 3D7 & Late Rings & 10~kb & 16711552 & 43\% & 57\% &  \cite{ay:three-dimensional} \\
-Ay-\textit{trophozoites} & 3D7 & Trophozoites &10~kb & 56348498 & 53\% & 47\% & \cite{ay:three-dimensional} \\
-Ay-\textit{schizonts} & 3D7 & Schizonts & 10~kb & 11652832 & 55\% & 45 \% & \cite{ay:three-dimensional} \\
-Lemieux-\textit{A4+} & IT/BC6+ & Rings & 25~kb & 18488252 & 19\% & 81\% & \cite{lemieux:genome-wide} \\
-Lemieux-\textit{A4} & IT/3G8 & Rings &  25~kb & 19674672 & 28\% & 72\% & \cite{lemieux:genome-wide}\\
-Lemieux-\textit{A44} & IT/BC6- & Rings & 25~kb & 18660594 & 25\% & 75\% & \cite{lemieux:genome-wide}\\
-Lemieux-\textit{DCJ\_On} & NF54/DCJ on & Rings & 25~kb & 3098370 & 26\% & 74\% &\cite{lemieux:genome-wide} \\
-Lemieux-\textit{DCJ\_Off} & NF54/DCF Off & Rings & 25~kb & 2533470 & 26\% & 73\% &\cite{lemieux:genome-wide} \\
-Lemieux-\textit{B15C2} & NF54/B15C2 & Rings & 25~kb &  1022996 & 12\% & 88\% & \cite{lemieux:genome-wide}\\
-\hline
-\end{tabular}
-\caption{Summary of the available {\em P. falciparum} Hi-C datasets}
-\end{table}
+\subsection*{Integrative analysis of gene expression and 3D structure using KernelCCA}
 
-\subsection*{Integrative analysis of gene expression and 3D structure using KernelCCA}
+Last but not least, an exciting contribution of \citet{ay:three-dimensional}
+is the integrative analysis of the 3D structure with gene expression using an
+unsupervised learning technique called ``kernel Canonical Correlation
+Analysis'' \citep{bach:kernel}. In this particular case,  KernelCCA aims at
+extracting a set of orthogonal gene expression component that exhibit
+coherence with respect (here) to the 3D structure. It can be helpful to
+think of this procedure as performing a Principal Component Analysis on the
+set of gene expression profiles but such that the component extracted are
+correlated with the gene expression.
+
+Let us take a closer look at how formally kCCA is used in this context.
+Consider the set of $n$ genes $g \in \mathcal{G}$. Each gene $g$ is
+represented on the one hand by its gene expression profile $e(g) \in \RR^{p}$
+and its 3D position $x(g) \in \RR^3$. Assume the set of gene expression
+profiles is mean centered and of unit variance.
+
+The goal of the procedure is to extract a gene expression component $v\in
+\RR^p, \|v\|=1$ such that it is both representative of set of gene expression profile,
+but also correlated with the 3D structure. We thus construct two scores to
+assess those properties. First, define $V(v) = \underset{g \in
+\mathcal{G}}{\sum} (v^T e(g))^2$. $V(v)$ measures the percentage of variance
+explained among the gene expression profile when projected onto the component
+$v$. Note that maximizing this score results in finding the first principal
+component. To assess smoothness along the structure,
+\citet{ay:three-dimensional} leverage a standard approach in kernel methods,
+where the smoothness of a score $f$ is quantified by the function:
+
+\begin{equation}
+S(f) = \frac{f^TK^{-1}_{3D}f}{\|f\|}\,,
+\end{equation}
+
+where $K_3D$ is the Gaussian kernel matrix of the genes' 3D coordinates. The
+smaller $S(f)$, the smoother $f$ is with respect to the 3D structure.
+
+In summary, to obtain smooth gene expression components $v$ on the 3D
+structure representative of the variance amongst all gene expression, one
+needs to: (1) maximize $V(v)$; (2) minimize $S(f)$; (3) maximize the
+correlation between $f$ and $v^T e(g)$. In other word, one would need to solve
+the kernelCCA problem, using on one hand a linear kernel over the gene
+expression profile, while on the other the Gaussian kernel over the 3D
+coordinates of each genes.
+
+\citet{ay:three-dimensional} apply this method on the three model obtained,
+and find highly correlated gene expression profiles with the 3D structure.
+Ranking the genes with their projection onto the gene components,
+\citet{ay:three-dimensional} demonstrate that several gene families and GO
+terms are enriched both close to the telomeres and at the opposite end of the
+nucleus.
+
+\section{Discussion}
 

### 2017-07-26 doc/manuscript/paper.tex
Added peng et al

@@
 associated to it. The parasite has a complex life cycle, with multiple stages
-both in the human and mosquito host. The human host is infected with
+both in the human and mosquito host. \sout{The human host is infected with
 \textit{sporozoites} through the bite of a infected female  {\em Anopheles}
@@
 starts anew. This asexual replication cycle is responsible for the symptoms
-and the complications of the disease: anemia, tertian fever, \dots
+and the complications of the disease: anemia, tertian fever, \dots}
 
@@
 
+The objective function $\mathcal{O}$, sometimes called the ``scoring
+function''
+
 \subsubsection*{Metric MDS-based methods}
@@
 \cite{ay:three-dimensional} & {\em P. falciparum} & Whole-genome & Fractal
+globule derived relationship & \CHECK  & & &
+& $\frac{1}{\delta_{ij}^2}$ \\
+\cite{peng:sequencing} & {\em P. falciparum} & Whole-genome & Fractal
 globule derived relationship & \CHECK  & & &
@@
 \scriptsize{\cite{zhang:inference}} & chromSDE & C &  \CHECK & & \CHECK\\
+\scriptsize{\cite{peng:sequencing}} & autochrom3D & C &  \CHECK & & \CHECK\\
 \scriptsize{\cite{rousseau:three}} & & E & & \CHECK & \CHECK\\

### 2017-07-26 doc/manuscript/paper.tex
more figures and schemas - rewording the introduction

@@
 % Title of paper
-\title{Analysing the three-dimensional structure of {\em P. falciparum}}
+\title{The art of modelling the 3D structure of the genome\\
+  Insights in {\em P. falciparum} genome organization}
 \author{Nelle Varoquaux}
@@
 the relative abundance of proteins related to chromatin structures, mRNA decay
-and translation rates suggests alternative mechanisms of gene regulation at the
-epigenetic and post-translational levels
-\citep{cui:chromatin-mediated, duffy:role, hoeijmakers:placing,
-horrocks:control, deitsch:mechanisms}.
+and translation rates suggests alternative mechanisms of gene regulation at
+the epigenetic and post-translational levels \citep{cui:chromatin-mediated,
+duffy:role, hoeijmakers:placing, horrocks:control, deitsch:mechanisms}. Thus,
+a better understanding of genome architecture of {\em P. falciparum}, both at
+a local and global scale may provide new leads for developing new therapies.
 
@@
 the past year, including many species of yeasts \citep{duan:three-dimensional,
-burton:species-level, mizuguchi:cohesin-dependant}, bacteria
+burton:species-level, mizuguchi:cohesin-dependent}, bacteria
 \citep{umbarger:three-dimensional}, flies \citep{sexton:three-dimensional},
 plants \citep{feng:genome-wide, wang:genome-wide} and numerous human and mouse
-cell lines \citep{lieberman-aiden:comprehensive, rao:3D}.
-
-Within the past few years, two studies of the three-dimensional structure of
-{\em P. falciparum} have been released.  \citet{lemieux:genome-wide} probed
-several strains of {\em P. falciparum} to understand the link between 3D
-structure and the complex regulation of \textit{var} genes, a family of genes
-involved the invasion of red blood cell and also responsible for the
-parasite's great capacity to evade our immune system.
-\citet{ay:three-dimensional} assayed the 3D structure of the genome of the
-parasite during three key stages of the erythrocytic cycle.
-\citet{ay:multiple} reviews the key biological findings of those studies.
-
-An important part of analysing the three-dimensional structure of the
-genome of {\em P. falciparum} (as well as many other organisms) relied on the
-inferring accurate 3D models of the structure. These models were then used as
-a preliminary step to various analysis, such as identifying colocolized
+cell lines \citep{lieberman-aiden:comprehensive, rao:3D}. Within the past few
+years, two studies of the three-dimensional structure of {\em P. falciparum}
+have been released.  \citet{lemieux:genome-wide} probed several strains of
+{\em P. falciparum} to understand the link between 3D structure and the
+complex regulation of \textit{var} genes, a family of genes involved the
+invasion of red blood cell and also responsible for the parasite's great
+capacity to evade our immune system. \citet{ay:three-dimensional} assayed the
+3D structure of the genome of the parasite during three key stages of the
+erythrocytic cycle. The relatively small size of the genome yet relatively
+complicated genome architecture, the complex life cycle and the strong link
+between chromatine folding, gene regulation and epigenetic places {\em P.
+falciparum} as an interesting case study for developing new methods for better
+understanding the properties of genome folding and its link to gene
+regulation. An important part of analysing the three-dimensional structure of
+the genome of {\em P. falciparum} (as well as many other organisms) relied on
+the inferring accurate 3D models of the structure. These models were then used
+as a preliminary step to various analysis, such as identifying colocolized
 elements, distinguishing between open and closed chromatin and last but not
@@
 chromatin modification.
+
+\fixme{Add why we are doing a review}
 
@@
 
+\begin{figure}
+\centering
+\includegraphics[width=0.5\linewidth]{schemas/schema.pdf}
+\caption{The art of understanding 3D genome structure}{
+\fixme{write caption}}
+\label{Fig1}
+\end{figure}
+
 \section*{Three-dimensional models of DNA from Hi-C data}
@@
 for their 3D genome architecture naturally paved the way to systematic and in
-depth study of the folding mechanisms of DNA.
-
-Over the past 10 years, the challenge of building 3D models from contact maps
-have been tackled by a plethora of methods. These model chromosomes as a
+depth study of the folding mechanisms of DNA. Over the past 10 years, the
+challenge of building 3D models from contact maps have been tackled by a
+plethora of methods, some developped for a particular organisms, others with
+the hope to be generalizable to any data sets. Methods model chromosomes as a
 series of beads, and attempt to place the beads in a 3D euclidean space to
@@
 \textit{ensemble} methods that yield a population of structures.
+
+\fixme{contrast model-driven studies with data-driven studies}
 
@@
 
+\begin{figure}
+\centering
+\includegraphics[width=0.7\linewidth]{figures/counts_maps.png}
+\caption{Contact maps of {\em P. falciparum}'s chr7}{
+\fixme{write caption}
+\textbf{A.} \citep{lemieux:genome-wide}}
+\label{Fig1}
+\end{figure}
+
 \subsection*{Consensus models}
@@
 The objective function $\mathcal{O}$, sometimes called the ``scoring
-function''
+function'', can be derived from known embedding algorithms (such as
+multidimensional-scaling methods), from statistical modeling of contact
+counts, or simply constructed in an \textit{ad hoc} manner to fulfill a set of
+wished properties.
 
@@
 count-to-distance mapping by fitting the relationship using known pairwise
-distances obtained through high-resolution FISH measures.
+distances obtained through high-resolution FISH measures. In brief, there are
+as many ways to derive the count-to-distance mapping as there are of studies.
 
@@
 best explains the observed data. The strength of this methods comes from the
-robustness to low signal-to-noise ratio.
+robustness to low signal-to-noise ratio, through a direct modeling of the
+noise through the statistical model.
 
@@
 methods,  the first approach is to consider each contact as a constraint and
-to formulate
-an under constrained optimization problem. A population of structures
-satistifying the constraints can be found by sampling local minimas.
+to formulate an under constrained optimization problem. A population of
+structures satistifying the constraints can be found by sampling local
+minimas.
 
@@
 count matrix. Applying classical MDS on this low rank PSD matrix then yields a
-3D model of the genome.
+\textit{consensus} 3D model of the genome.
 
@@
 
-In the previous section, we discussed how from a contact count map one could
-infer a {\em consensus} model or a population of models that best represented
-the contact map. Another set of methods approach the problem of finding 3D
-models of genome architecture very differently. Instead of using data-driven
-approaches, these methods model chromosomes as series of beads or flexible
-fibers, and attempt to find the smallest set of constraints such that a
-generated contact maps from these models match as closely as possible the
-observed contact maps. As these models refine the constraints, more insight on
-the hallmarks of the genome architecture are discovered.
+``data-driven'' methods, as presented in the last section, use the
+experimental contact maps to infer models as consistent as possible with the
+data. Another approach consists in modeling chromosomes as polymers under a
+small number constraints such that a generated contact maps from these models
+match as closely as possible the observed contact maps. These ``model-driven''
+approaches offer mechanistical insights into the genome architecture.
 

### 2017-07-26 scripts/population_structures/clustering_across_stages_downsampled_distances.py

@@ -134,8 +134,11 @@ except OSError:
     pass
 if args.chromosome is None:
     outfile = "images/pca_downsampled_distances_%s.png" % algorithm
+    outfile = "images/pca_downsampled_distances_%s.pdf" % algorithm
 else:
     outfile = "images/pca_downsampled_distances_%s_chr%02d.png" % (
+        algorithm, args.chromosome)
+    outfile = "images/pca_downsampled_distances_%s_chr%02d.pdf" % (
         algorithm, args.chromosome)
 
 fig.savefig(outfile)

### 2017-07-26 scripts/population_structures/compute_pca.sh

@@ -9,5 +9,5 @@ done;
 
 python clustering_across_stages_downsampled_distances.py
 python clustering_across_stages_downsampled_distances.py -a UNB0
-
+python clustering_across_stages_downsampled_distances.py -a PO
 

### 2017-07-26 doc/manuscript/paper.tex
More on the introduction

@@
 complicated genome architecture, the complex life cycle and the strong link
-between chromatine folding, gene regulation and epigenetic places {\em P.
+between chromatin folding, gene regulation and epigenetic places {\em P.
 falciparum} as an interesting case study for developing new methods for better
 understanding the properties of genome folding and its link to gene
-regulation. An important part of analysing the three-dimensional structure of
-the genome of {\em P. falciparum} (as well as many other organisms) relied on
-the inferring accurate 3D models of the structure. These models were then used
-as a preliminary step to various analysis, such as identifying colocolized
+regulation.
+
+An important part of analyzing the three-dimensional structure of the genome
+of {\em P. falciparum} (as well as many other organisms) relied on the
+inferring accurate 3D models of the structure. These models were then used as
+a preliminary step to various analysis, such as identifying colocalized
 elements, distinguishing between open and closed chromatin and last but not
-least, intergrating with other sources of data, such as gene expression or
-chromatin modification.
+least, integrating with other sources of data, such as gene expression or
+chromatin modification. In recent years, a profusion of methods were developed
+for creating 3D models of the genome, either as standalone methods or with the
+specific goal to better understand an organism, a biological process (such as
+the inactivation of chromosome X) or the folding of a chromosome. These
+methods broadly fall into two categories: ``model-based'' and ``data-driven''.
+The former methods consider the polymer nature of DNA to leverage the
+theoretical and computational work done in statistical physics of polymers to
+build with as few assumptions as possible many chromosome conformations. Those
+chromosome conformations are then used to compare against experimental data,
+such as Hi-C contact count matrices, in order to iteratively improve the
+models. These models offer mechanistical insights into the folding of DNA. The
+latter approaches use the experimental data to infer 3D models, by typically
+minimizing a cost function ensuring the models are as consistent as possible
+with the data.
 
-\fixme{Add why we are doing a review}
-
-The study of chromosome organization based on contact count maps broadly
-falls into two categories: model-based studies and data-driven studies. The
-former methods consider the polymer nature of DNA to leverage the theoretical
-and computational work done in statistical physics of polymers to build with
-as few assumptions as possible many chromosome conformations. Those chromosome
-conformations are then used to compare against experimental data, such as Hi-C
-contact count matrices, in order to iteratively improve the models. These
-models offer mechanistical insights into the folding of DNA. The latter
-approaches use the experimental data to infer 3D models, by typically minizing
-a cost function ensuring the models are as consistent as possible with the
-data. These data driven models and analysis, with a particular focus on the
-one applied to {\em P. falciparum} are the primary focus of this
-review.
+This paper provides a view on the work accomplished on building and using 3D
+models using contact maps for better understanding of genomic and epigenetic
+processes, with a particular highlight on the {\em P falciparum}. It dwells on
+modeling challenges: why, how and what do 3D models teach us about gene
+regulation? The first and second sections of this paper discuss respectively
+current data- and model-driven methods to build 3D structures of DNA
+using contact maps.  The last section reviews how these models can be used to
+discover the key role of genome architecture in biological process, with a
+particular focus on discoveries on {\em P. falciparum} genome architecture.
 
 % XXX See rosa:computational
-
-Though we here review some of the methods used to study and build models, this
-is a very incomplete view of a blooming field. \citet{rosa:computational}
-provide a more thorough (but again incomplete) overview of computational models
-of genome architectures.
 

### 2017-07-26 doc/manuscript/paper.tex
small rewording

@@
 associated to it. The parasite has a complex life cycle, with multiple stages
-both in the human and mosquito host. \sout{The human host is infected with
-\textit{sporozoites} through the bite of a infected female  {\em Anopheles}
+both in the human and mosquito host.
+
+\begin{figure}
+\caption{The life cycle of {\em P. falciparum}}{
+\small
+The human host is infected by \textit{sporozoites} through the bite of a infected female  {\em Anopheles}
 mosquito. The \textit{sporozoites} quickly migrate to the liver, where they
@@
 and the complications of the disease: anemia, tertian fever, \dots}
+\end{figure}
 
@@
 the epigenetic and post-translational levels \citep{cui:chromatin-mediated,
-duffy:role, hoeijmakers:placing, horrocks:control, deitsch:mechanisms}. Thus,
+duffy:role, hoeijmakers:placing, horrocks:control, deitsch:mechanisms}:
 a better understanding of genome architecture of {\em P. falciparum}, both at
@@
 \textit{ensemble} methods that yield a population of structures.
-
-\fixme{contrast model-driven studies with data-driven studies}
 
@@
 
-In addition of having subtle differences in the derivation of wish distances,
+In addition of having subtle differences in the derivation of wish distances
+(which can have important effects on the resulting structures),
 the different methods also vary by the inclusion of weights to reflect
@@
 genome rely on a new protocol to probe single-cell for their 3D structures
-\citep{nagano:single-cell,ramani:massively}. Single-cell Hi-C is still in its early days,
-and only a handful of data sets are today publicly available. The contact maps
-originating from these data sets are very sparse, and specific methods to
-infer 3D structures need to be developed specifically for sc-HiC. The
-challenge of constructing models from this data has been tackled very
-differently by two groups.
+\citep{nagano:single-cell,ramani:massively}. Single-cell Hi-C is still in its
+early days, and, despite potential for assessing the variability of
+cell-to-cell genome architecture in a genome-wide fashion, only a handful of
+data sets are today publicly available. The contact maps originating from
+these data sets are very sparse, and specific methods to infer 3D structures
+need to be developed specifically for sc-HiC. Two methods have been developed
+to tackle the challenging task of constructing models from this data.
 
@@
 \end{center}
-\caption{\bf A comparison of 3D inference methods}{In this table, we summarize
+\caption{\bf A comparison of 3D inference methods}{\small In this table, we summarize
 properties of published methods to infer the 3D structure of the genome: (1)
@@
 centromeres are constrained into a spherical ball tethered to the nuclear
-membrane; \textit{(iii)} telomeres are tethered to the nuclear membraine;
+membrane; \textit{(iii)} telomeres are tethered to the nuclear membrane;
 \textit{(iv)} rDNA is constrained into the nucleolus, represented as a
-spherical ball opposite to the centromeres.
-
-Let us take a closer look at the model proposed by \citet{tjong:physical}.
-The 16 chromosomes are modeled as beads-on-a-string spaced 3.2kb apart. The
-authors cast an optimization problem including the following restraints:
-
-\begin{itemize}
-\item \textbf{Adjacent Beads Constraints} Two adjacent beads spaced 3.2kb
-apart are constrained to be 30~nm apart;
-\item \textbf{Volume Exclusion Constraints} Each pairs of beads are spaced at
-minimum $30~nm$ apart;
-\item \textbf{Nucleus Constraint} All beads lie in a spherical nucleus of
-radius $1\mu m$
-\item \textbf{Centromeres Constraints} Centromeres are localized in a small sphere of
-50nm abutting the nuclear envelop.
-\item \textbf{Telomere Constraints} Telomeres are located at most 50nm from the nuclear envelop.
-\item \textbf{rDNA restraint} The repeated rDNA located on chromosome 12 is
-constrained to be inside the nucleolus;
-\item \textbf{Nucleolus exclusion restraint} All beads that isn't rDNA is
-constrained outside of the nucleolus;
-\item \textbf{Chain stiffness}: an angular restraint is imposed on three
-consecutive beads to reproduce chain stiffness:
-\begin{equation}
-\frac{1}{2} k_{\text{angle}} \sum^{N - 2}_{i = 1} \left( 1 - \frac{x_{i + 1} -
-x_i}{\|x_{i + 1} - x_i\|} \cdot \frac{x_{i + 2} - x_{i + 1}}{\|x_{i + 2} -
-  x_{i + 1}\|} \right)^2\,,
-\end{equation}
-\end{itemize}
+spherical ball opposite to the centromeres. A scoring function eng
 

### 2017-07-26 doc/manuscript/paper.tex
more rewording
bullet pointed the discussion

@@
 
+\subsection*{Model evaluation}
+
+A substantial difficulty in modeling the 3D structure of the genome is that
+model evaluation tends to the subjective. What is the relevant measure?
+``Truth'' is generally not fully available, except for a few pairs of loci or
+in simulations. Is validating the colocalization of a pair or a few pairs of
+loci via FISH experiments enough? Is fit to the contact maps or agreements
+between modeling techniques relevant?
+
 \section*{Model-driven approach: modeling DNA as a flexible fibers under constraints}
@@
 
-In previous section, I have reviewed data-driven methods to infer either
-\textit{consensus} or \textit{ensemble} models of the 3D structure. The reader
-may well ask why go through the effort to obtain such model and not directly
-study the contact maps?  In this section, I will review a number of downstream
-analysis one can do on 3D models, highlighting but limiting myself to results
-on the 3D structure of {\em P. falciparum}.
-
-\subsection{Probing the 3D structure of the {\em P. falciparum}}
-In the past year, two groups have published studies of {\em P.
-falciparum}'s 3D genome architecture using Hi-C. \citet{lemieux:genome-wide}
-studied several strains associated with populations expression unique
-\textit{var} genes, in order to study key folding properties of the 3D
-structure relating to \textit{var} gene expression. On the other hand,
-\citet{ay:three-dimensional} focus three key time-point of the development
-cycle. Table~\ref{tab:data} summarizes the available data sets and their key
-properties after processing the data as described in
-\citet{ay:three-dimensional} and combining all available libraries for each
-life-cycle stage and strain.
-
 \begin{table}
@@
 
-\fixme{Add the fact we have time points}
-
-\subsection*{Model evaluation}
-
-A substantial difficulty in modelling the 3D structure of the genome is that
-model evaluation tends to the subjective. What is the relevant measure?
-``Truth'' is generally not fully available, except for a few pairs of loci or
-in simulations. Is validating the colocalization of a pair or a few pairs of
-loci via FISH experiments enough? Is fit to the contact maps or agreements
-between modeling techniques relevant?
-
-\fixme{Add section on model evaluation}
+In previous section, I have reviewed data-driven methods to infer either
+\textit{consensus} or \textit{ensemble} models of the 3D structure. The reader
+may well ask why go through the effort to obtain such model and not directly
+study the contact maps?  \fixme{why?}In this section, I will review a number of downstream
+analysis one can do on 3D models, highlighting but not limiting myself to results
+on the 3D structure of {\em P. falciparum}. \fixme{why}
 
@@
 
-\subsection{Chromatin compaction}
+\subsection*{Chromatin compaction}
 
@@
 
-\section{Discussion}
+\section*{Discussion}
+
+data-driven Plethora of methods, but only handfull that can be used on different
+organisms and have been validated on simulated data set. Ensemble methods are
+more challenging to use, to validate and to explore, but offer more
+biologically accurate view of what is going on in teh nucleus. Overall,
+powerfull tool but that still needs to be extended oto work on challenging
+data set.
+
+model-driven methods are organisms and study specific. For each data set, the
+set of constraints need to be revisited. Yet, once built, they can be used ot
+replace high throughput FISH experiments at low cost.
+
+Downstream analysis gave important insights in the relation between gene
+regulation and genome architecture of teh {\em P. falciparum}. Yet a lot
+remains to be done. How to compare in a meaningful manner structures from
+different time points at a genome-scale level?
 
@@
 
-\begin{figure}[!p]
-\centering
-%\includegraphics{fig1}
-\caption{}
-\label{Fig1}
-\end{figure}
-
 \end{document}

### 2017-07-27 doc/manuscript/paper.tex
First draft of the abstract

@@
-%\documentclass[2columns]{article}
+\documentclass[2columns]{article}
 
-\documentclass[letterpaper,12pt]{article}
+%\documentclass[letterpaper,12pt]{article}
 \usepackage{url}
@@
 %% Load De Gruyter specific settings 
-\usepackage{dgjournal}
+%\usepackage{dgjournal}
 
@@
 % Title of paper
-\title{The art of modelling the 3D structure of the genome\\
-  Insights in {\em P. falciparum} genome organization}
+\title{Unfolding the genome's 3D architecture:\\
+  \Large{Insights in {\em P. falciparum} genome organization}}
 \author{Nelle Varoquaux}
 
-%\maketitle
+\maketitle
 
 \begin{abstract}
+\begin{itemize}
+
+\item {\em P. falciparum} is the deadliest parasite causing human malaria,
+responsible for more than 500,000 death in 2015. Our lack of understanding of
+the parasite gene regulation of its complex life cycle is an impediment to the
+development of novel therapies. Recent evidence points towards unusual means
+of gene regulation, possible through changes in its genome architecture. A
+better understanding of how DNA folds into the nucleus and how it influences
+gene expression could provide new leads for developing new drugs.
+\item In the meantime, the development of new means to probe samples for the
+three-dimensional structure of DNA paves the way for in depth and systematic analysis of genome
+architecture, offering new insights in many biological processes. 3C-like
+methods can now assess physical interactions between pairs of loci in a
+genome-wide fashion, thus leading to the creation of contact maps. The
+democratization of such protocols creates many new opportunities for
+methodological development: how to infer 3D models from these contact maps?
+Can such models help us gain insights in biological process?
+\item Here, I review recent methodological developments \fixme{finish this} 
+\end{itemize}
 
@@
 
-While efficacious vaccines still remains a hope and drug resistance to anti
+While safe and effictive vaccines still remains a hope and drug resistance to anti
 malaria continues to arise, genomic-based research on malaria seeks new
@@
 \includegraphics[width=0.5\linewidth]{schemas/schema.pdf}
-\caption{The art of understanding 3D genome structure}{
-\fixme{write caption}}
+\caption{The art of understanding 3D genome structure}
 \label{Fig1}
@@
 
-The development of genowe-wide and high-throughput protocols to probe samples
+The development of genome-wide and high-throughput protocols to probe samples
 for their 3D genome architecture naturally paved the way to systematic and in
@@
 challenge of building 3D models from contact maps have been tackled by a
-plethora of methods, some developped for a particular organisms, others with
+plethora of methods, some developed for a particular organisms, others with
 the hope to be generalizable to any data sets. Methods model chromosomes as a
@@
 \begin{figure}
-\centering
+\begin{center}
 \includegraphics[width=0.7\linewidth]{figures/counts_maps.png}
-\caption{Contact maps of {\em P. falciparum}'s chr7}{
-\fixme{write caption}
-\textbf{A.} \citep{lemieux:genome-wide}}
-\label{Fig1}
+\end{center}
+\caption{Contact maps of {\em P. falciparum}'s chr7}{ \small Both contact maps show
+enrichment of contact counts in VRSSM clusters (\textbf{A.} Ring stage 
+\citep{lemieux:genome-wide} of IT strains, \textbf{B.} 
+Trophozoite stage of 3D7 \citep{ay:three-dimensional}) 
+Dark blue corresponds to high interactions, light yellow to low interactions.
+The strong diagonal reflects the proximity of adjacent regions of the genome.
+The black line indicates the centromeres. }
 \end{figure}
@@
 c^\alpha)$, and to infer jointly the parameters jointly with the structure
-\citet{zhang:inference}.
+\citep{zhang:inference}.
 
@@
 robustness to low signal-to-noise ratio, through a direct modeling of the
-noise through the statistical model.
+noise through the statistical model (See figure~\ref{fig:3D_structures}.
 
-\fixme{add results on the P. falciparum to show pastis's results on 
-super noisy data}
+\begin{figure}
+\includegraphics[width=\linewidth]{schemas/3D_structures.pdf}
+\label{fig:3D_structures}
+\caption{3D models of {\em P. falciparum}'s genome architecture}{Using
+\texttt{pastis-MDS} (\textbf{A} and \textbf{B}) and \texttt{pastis-PO}
+(\textbf{C} and \textbf{D}) on two
+HiC data set of {\em P. falciparum}. Each color corresponds to a chromosome.
+Large white beads mark centromeres, small blue beads telomeres and green beads
+VRSSM clusters. Centromeric cluster is marked with a black line, telomeric
+cluster is marked with a dashed line.}
+\end{figure}
 
-\subsection*{Ensemble methods}
+\subsection*{Ensemble methods as a mean to infer population of structures}
 
@@
 into contact or ensure a minimal or maximal distance between those. The models
-includes three differente types of restraints: (i) beads seen interacting are
+includes three different types of restraints: (i) beads seen interacting are
 restrained with harmonic oscillators of strengths derived from the contact
@@
 from one another; (iii) the last oscillators are activated only when
-constraints are not fulfilled to XXX.  This yields an optimization problem
+constraints are not fulfilled to \todo{XXX}.  This yields an optimization problem
 with a large number of local minima, which the authors sample from by running
@@
 \textit{(iv)} rDNA is constrained into the nucleolus, represented as a
-spherical ball opposite to the centromeres. A scoring function eng
+spherical ball opposite to the centromeres. One can then simulate a large set
+of random structures fulfilling the constraints, and generate a
+``volume-exclusion contact map'' from this population of structures,
+considering that two beads that are less then 45~nm apart in any of the
+structures form a contact. The Pearson correlation of the volume-exclusion
+contact map and the Hi-C one are highly correlated, demonstrating this small
+set of constraints fully explain the observed counts. In addition, the
+population of structures also explain FISH experiments previously published.
 
-Using IMP, the authors generate a 25,000 structures, all of which respect the
-constraints listed above. The population of structures is then used to created
-a ``volume-exclusion contact map'', considering that two beads that are less
-then 45~nm apart from a contact. The Pearson correlation of the
-volume-exclusion contact map and the Hi-C one are highly correlated,
-demonstrating this small set of constraints fully explain the observed counts.
-In addition, the population of structures also explain FISH experiments
-previously published.
+\subsection*{Volume-exclusion modeling cannot explain {\em P. falciparum}'s
+genome architecture}
+
+For at least some stages, {\em P. falciparum} share a lot of common genomic
+architecture hallmarks with the budding yeast: the centromeres are strongly
+co-localized at one end of the nucleus, telomeres are in physical contacts,
+\dots So could the primary architectural features also arise from a population
+of constrained but otherwise random of structures? Adapting the set of
+constraints to match biological knowledge of {\em P. falciparum},
+\citet{ay:three-dimensional} showed the resulting simulated VE map not only
+yielded lower correlations, but also did not show the same features as the
+original contact count matrix. In particular, the VRSM genes which display
+domain-like enrichment in interactions did not appear in the simulated VE map
+(See Figure~\ref{fig:ve_modeling})
+
+\begin{figure}
+\caption{Volume Exclusion Modeling}{}
+\label{fig_ve_modeling}
+\end{figure}
 
@@
 
-The second question one can ask is how do the structures differ. Tackling this
-question is very challenging, but can be reformulated in a much easier way.
-Are the hallmarks of interest conserved across structures of a same stage?
-\citet{ay:three-dimensional} and \citet{lemieux:genome-wide} both identified
-{\em P. falciparum} folded in very specific ways, with VRSSM genes highly
-interacting. \citet{ay:three-dimensional} also observed strong clustering of
-the centromeres, and enrichment in interaction at the telomeres. These
-observation can lead to a rigorous approach at identifying whether families
-of loci clustered in the structures and even quantifying this in a rigourous
-manner.
+The second question is: ``are models locally consistent''
+\citep{bau:three-dimensional}. A possible answer to this questions is to
+divide the structures into overlapping sub-structures ranging from 5 to 20
+beads, and compute the pairwise RMSD across all structures. Segments
+overlapping within a certain range for a large number of models can be
+assessed as locally consistent, while others labelled as highly variable.
 
-\subsection*{Chromatin compaction}
+The last question one can ask is: ``how do the structures differ?'' . Tackling
+this question is very challenging, but can be reformulated in a much easier
+way. Are the hallmarks of interest conserved across structures of a same
+stage? \citet{ay:three-dimensional} and \citet{lemieux:genome-wide} both
+identified {\em P. falciparum} folded in very specific ways, with VRSSM genes
+highly interacting. \citet{ay:three-dimensional} also observed strong
+clustering of the centromeres, and enrichment in interaction at the telomeres.
+These observation can lead to a rigorous approach at identifying whether
+families of loci clustered in the structures and even quantifying this in a
+rigourous manner.
 
-Bau et al
+\subsection*{Chromatin compaction and chromosome entanglement}
+
+3D models can be used to estimated chromatin compaction and chromosome
+entanglement. The former can be estimated by looking at the number of base
+pairs in a region defined either by volume if the scale of the structure is
+known, or by a percentage of the size of the structure
+\citep{bau:three-dimensional}. Another idea is to sample random beads of a
+certain diameter, and assess how many loci are seen interacting. Applying this
+method on the 3 models of {\em P. falciparum}, \citet{ay:three-dimensional}
+showed the trophozoite stage exhibited a more open chromatin than the ring and
+schizont stage, coherent with the transcriptionally active state of the
+parazite during this moment of the life cycle. A similar analysis, but
+counting the number of inter-chromosomal interaction can help assess the
+chromosome entanglement of the structure.
 
@@
 
-data-driven Plethora of methods, but only handfull that can be used on different
-organisms and have been validated on simulated data set. Ensemble methods are
-more challenging to use, to validate and to explore, but offer more
-biologically accurate view of what is going on in teh nucleus. Overall,
-powerfull tool but that still needs to be extended oto work on challenging
-data set.
+Plethora of methods for inferring 3D models from contact maps have been
+developed: some are very specific to an organism or a region of the genome,
+while others are generalizable easily to many organisms. As 3C-type method are
+democratized, a wider audience of researchers are in need of robust and
+well-implemented algorithms for building models. Surprisingly, only a handfull
+of methods are both publicly available, generalizable and well-validated.
+While \textit{ensemble} methods offer a more biologically accurate view of the
+variety of chromatin folding in a population of cells, their use is more
+challenging, both as a result of a lack of comparison and validation of the
+methods, but also as large population of structures are less amenable to
+exploration and visualization than \textit{consensus} structures.
 
-model-driven methods are organisms and study specific. For each data set, the
-set of constraints need to be revisited. Yet, once built, they can be used ot
-replace high throughput FISH experiments at low cost.
+At the other end of the spectrum, model-driven methods are organisms and study
+specific, and require good understanding of both polymer physics and of the
+particular organism studied: for each data set, the set of constraints to
+apply on the structure need to be revisited and remained. Yet, once built,
+they provide extensive insights in the average location of genes and may be a
+mean to replace high throughput FISH experiments at low cost as a first
+exploration tool.
 
-Downstream analysis gave important insights in the relation between gene
-regulation and genome architecture of teh {\em P. falciparum}. Yet a lot
-remains to be done. How to compare in a meaningful manner structures from
-different time points at a genome-scale level?
+While downstream analysis of the {\em P. falciparum} gave important insights
+in the relation between gene regulation and genome architecture of the {\em P.
+falciparum}, the use of 3D models for downstream analysis remains scarce in
+the literature and many avenues are left opened to methodological development.
+How to compare in a meaningful manner structures from different time points at
+a genome-scale level?
 

### 2017-07-28 doc/manuscript/paper.tex
more fixes

@@
 \begin{abstract}
-\begin{itemize}
-
-\item {\em P. falciparum} is the deadliest parasite causing human malaria,
-responsible for more than 500,000 death in 2015. Our lack of understanding of
-the parasite gene regulation of its complex life cycle is an impediment to the
-development of novel therapies. Recent evidence points towards unusual means
-of gene regulation, possible through changes in its genome architecture. A
-better understanding of how DNA folds into the nucleus and how it influences
-gene expression could provide new leads for developing new drugs.
-\item In the meantime, the development of new means to probe samples for the
-three-dimensional structure of DNA paves the way for in depth and systematic analysis of genome
+The development of new means to probe samples for the three-dimensional
+structure of DNA paves the way for in depth and systematic analysis of genome
 architecture, offering new insights in many biological processes. 3C-like
@@
 Can such models help us gain insights in biological process?
-\item Here, I review recent methodological developments \fixme{finish this} 
-\end{itemize}
 
+Recent studies assessed {\em P. falciparum} (the deadliest of the 5
+human malaria parasites) for its genome organization using 3C-like protocol.
+With its small genomic size, fairly simple yet changing genomic organization
+across its lifecyle and strong correlation between chromatin folding and gene
+expression, this parasite is the ideal case study for applying and developing
+methods to infer 3D models and use them for downstream analysis.
+
+Here, I review a set of methods used to build and analyse three-dimensional
+models from contact maps data with a special highlight findings on {\em P.
+falciparum}'s genome organization.
 \end{abstract}
@@
 measures of mortality and the disease is thought to cost more than US\$21
-billion a year \citep{onwujekwe:do}.
-Malaria is caused by small unicellular protozoan {\em Plasmodium}, infecting
-its host through a mosquito bite. Out of the 5 species that can infect human,
-{\em P. falciparum} is by the far the most deadly with 99\% of the deaths
-associated to it. The parasite has a complex life cycle, with multiple stages
-both in the human and mosquito host.
+billion a year \citep{onwujekwe:do}. The disease is caused by small
+unicellular protozoan parasite {\em Plasmodium} that infects its host through
+a mosquito bite. Out of the 5 species that can infect human, {\em P.
+falciparum} is by the far the most deadly with 99\% of the deaths associated
+to it. The parasite has a complex life cycle, with multiple stages both in the
+human and mosquito host.
 
@@
 
-While safe and effictive vaccines still remains a hope and drug resistance to anti
+While effective vaccines still remains a hope and drug resistance to anti
 malaria continues to arise, genomic-based research on malaria seeks new
@@
 
-\section*{Three-dimensional models of DNA from Hi-C data}
+\section*{``Data-driven'' three-dimensional models of DNA from contact maps}
 
@@
 hallmarks of genome architecture shared across all cells despite the
-cell-to-cell variability \citet{nagano:single-cell}. More amenable for
+cell-to-cell variability \citep{nagano:single-cell}. More amenable for
 visualization and analysis, this average structure can be easily integrated
 with other sources of data, such as RNA-seq, chip-seq etc, which are also
-population-based.
+population-based. On the other hand, interpretation of \textit{consensus}
+model need to be handle with care: these structures should not be considered
+as a true representation of the chromatin folding in the nucleus, but as a
+summary model of the contact maps.
 
@@
 confidence in ``wish-distances'' \citep{ay:three-dimensional}, or constraints
-to reflect prior knowledge on the structure. Table~\ref{table:mds_detail}
-summarizes the characteristic of each method.
+to reflect prior knowledge on the structure. See table~\ref{table:mds_detail}
+for a summary of the characteristic of each method.
 
@@
 c^\alpha)$, and to infer jointly the parameters jointly with the structure
-\citep{zhang:inference}.
+\citep{zhang:inference, varoquaux:statistical}.
 
@@
 
-\subsubsection*{Statistical-based models}
+\subsubsection*{Statistical models for contact counts}
 
-\citet{varoquaux:statistical} proposes a novel approach called \texttt{Pastis}
-based on statistical modeling of contact counts as random Poisson variables
-where the intensity is a function of the distance: $c_{ij} \sim
-\text{Poisson}(\beta d_{ij}^\alpha)$, and some perform the optimization only
-including a subset of contact counts. The authors propose to cast the
-inference of 3D models as maximizing the likelihood:
+\begin{figure}
+\includegraphics[width=\linewidth]{schemas/3D_structures.pdf}
+\caption{3D models of {\em P. falciparum}'s genome architecture}{Using
+\texttt{pastis-MDS} (\textbf{A} and \textbf{B}) and \texttt{pastis-PO}
+(\textbf{C} and \textbf{D}) on two
+HiC data set of {\em P. falciparum}: a low signal-to-noise ratio data set
+at a {\em Ring} stage \citep{lemieux:genome-wide}
+(\textbf{A} and \textbf{C}) and a high signal-to-noise ratio data set
+at the {\em Trophozoit} stage \citep{ay:three-dimensional}
+(\textbf{B} and \textbf{D}). \texttt{pastis-MDS} recovers a plausible
+structure for the high quality data set, but the non-uniform distribution of
+beads on low quality data set suggests that the method is not adequate on such
+data set. In contrast, \texttt{pastis-PO} recovers plausible structures both
+on the high and low quality data sets.
+Each color corresponds to a chromosome.
+Large white beads mark centromeres, small blue beads telomeres and green beads
+VRSSM clusters. Centromeric cluster is marked with a black line, telomeric
+cluster is marked with a dashed line.
+}
+\label{fig:3D_structures}
+\end{figure}
+
+Another approach consists of modeling contact counts as random variables, with
+the 3D model being a latent variables. For example,
+\citet{varoquaux:statistical} proposes an approach called
+\texttt{Pastis}, that models contact counts as random Poisson variables where
+the intensity of the process between $i$ and $j$ is a function of the
+distance: $c_{ij} \sim \text{Poisson}(\beta d_{ij}^\alpha)$. The inference can
+thus be cast as maximizing the likelihood:
 
@@
 robustness to low signal-to-noise ratio, through a direct modeling of the
-noise through the statistical model (See figure~\ref{fig:3D_structures}.
-
-\begin{figure}
-\includegraphics[width=\linewidth]{schemas/3D_structures.pdf}
-\label{fig:3D_structures}
-\caption{3D models of {\em P. falciparum}'s genome architecture}{Using
-\texttt{pastis-MDS} (\textbf{A} and \textbf{B}) and \texttt{pastis-PO}
-(\textbf{C} and \textbf{D}) on two
-HiC data set of {\em P. falciparum}. Each color corresponds to a chromosome.
-Large white beads mark centromeres, small blue beads telomeres and green beads
-VRSSM clusters. Centromeric cluster is marked with a black line, telomeric
-cluster is marked with a dashed line.}
-\end{figure}
+noise through the statistical model (See figure~\ref{fig:3D_structures}).
 
@@
 
-\subsubsection*{Estimating the posterior distribution}
+\subsubsection*{Estimating the posterior distribution of a statistical model}
 
@@
 posterior using MCMC. Obtaining a consensus structure from such a method can
-be accomplished by selecting the maximum a posteriori.
+be accomplished by selecting the maximum {\em a posteriori}.
 
@@
 these data sets are very sparse, and specific methods to infer 3D structures
-need to be developed specifically for sc-HiC. Two methods have been developed
-to tackle the challenging task of constructing models from this data.
+need to be developed specifically for sc-HiC.
 
@@
 to formulate an under constrained optimization problem. A population of
-structures satistifying the constraints can be found by sampling local
+structures satisfying the constraints can be found by sampling local
 minimas.
@@
 \textit{consensus} 3D model of the genome.
+
+\fixme{hirata et al}
 
@@
 
-\subsection*{Model evaluation}
+\subsection*{Model evaluation and comparison}
 
@@
 loci via FISH experiments enough? Is fit to the contact maps or agreements
-between modeling techniques relevant?
+between modeling techniques relevant? Are the conclusion drawn from 3D models
+in agreement when these are inferred from different methods?
 
-\section*{Model-driven approach: modeling DNA as a flexible fibers under constraints}
+First, methods can be validated and compared against contact maps simulated
+from a known ground truth \citep{zhang:inference, varoquaux:statistical,
+lesne:3d, trieu:3D}. Note that while this is a simple and natural first step
+for methods inferring a \textit{consensus} structure, comparing and assessing
+robustness and accuracy of {\em ensemble} of 3D models is much more
+challenging: in fact, it is still an untackled problem. Second, one can assess
+the stability and robustness of the inference with respect to (1) data
+bootstrapping; (2) contact map resolution (the models should not change as the
+resolution of the data varies) \citep{zhang:inference, varoquaux:statistical};
+(3) and in between biological replicates \citep{zhang:inference,
+varoquaux:statistical}. As a cautionary remark, it is important to stress that
+it is not because a method is very stable that is ``good'' with respect to any
+criterion: I can imagine a number of very stable methods, that would yet
+provide absolutely no insights on genome organization. Third, models can be
+compared to other sources of data, such as FISH
+\citep{ay:three-dimensional,duan:three-dimensional}. Fourth, biological
+plausibility of the resulting models can be considered: are the beads
+uniformly distributed in the cell \citep{ay:three-dimensional}? are known
+hallmarks of the genome architecture such as centromeres clustering preserved?
 
-``data-driven'' methods, as presented in the last section, use the
+These are a handfull of ways to assess plausibility and accuracy of 3D
+reconstruction, but many avenues in model evaluation and comparison are yet to
+be explored.
+
+\section*{The art of modeling genome architecture.}
+
+``Data-driven'' methods, as presented in the last section, use the
 experimental contact maps to infer models as consistent as possible with the
-data. Another approach consists in modeling chromosomes as polymers under a
-small number constraints such that a generated contact maps from these models
-match as closely as possible the observed contact maps. These ``model-driven''
-approaches offer mechanistical insights into the genome architecture.
+data. ``Model-driven'' tackle the 3D-modeling challenge exactly in the
+opposite manner. Consider the following task. You have a tens of thousands of
+randomly placed beads-on-strings. Can you find the smallest sets of
+constraints such that these beads-on-a-string interact overall the same way as
+a given contact map? This is the daunting task accomplished by
+``model-driven'' approach: chromosomes are modeled as polymers (or random
+self-excluding fibers) under a small number constraints such that a generated
+contact maps from these models match as closely as possible the observed
+contact maps. These ``model-driven'' approaches offer mechanistical insights
+into the genome architecture.
 
-\subsection*{A small set of constraints explain the 3D architecture of \textit{S. cerevisiae}}
+\subsection*{Building a yeast nucleus}
 
@@
 
-\subsection*{Volume-exclusion modeling cannot explain {\em P. falciparum}'s
-genome architecture}
+\subsection*{Building a {\em P. falciparum} nucleus?}
+
+\begin{figure}
+\caption{Volume Exclusion Modeling}{}
+\label{fig:ve_modeling}
+\end{figure}
 
@@
 architecture hallmarks with the budding yeast: the centromeres are strongly
-co-localized at one end of the nucleus, telomeres are in physical contacts,
+co-localized at one end of the nucleus (for at least some stages),
+telomeres are in physical contacts with one another,
 \dots So could the primary architectural features also arise from a population
@@
 original contact count matrix. In particular, the VRSM genes which display
-domain-like enrichment in interactions did not appear in the simulated VE map
-(See Figure~\ref{fig:ve_modeling})
-
-\begin{figure}
-\caption{Volume Exclusion Modeling}{}
-\label{fig_ve_modeling}
-\end{figure}
+domain-like enrichment in interactions do not appear in the simulated VE map
+(See Figure~\ref{fig:ve_modeling}). So\dots Can we add constraints on VRSM
+gene clusters to improve correlations between true and generated contact maps?
+Adding a constraint on all beads considered as VRSM clusters is not enough:
+running 100 experiments yielded structures that did not fulfill all the
+constraints, demonstrating VRSM genes cannot all cluster together in a cell.
 
@@
 may well ask why go through the effort to obtain such model and not directly
-study the contact maps?  \fixme{why?}In this section, I will review a number of downstream
+study the contact maps?  In this section, I will review a number of downstream
 analysis one can do on 3D models, highlighting but not limiting myself to results
-on the 3D structure of {\em P. falciparum}. \fixme{why}
+on the 3D structure of {\em P. falciparum}.
 

### 2017-07-28 doc/manuscript/paper.tex
More progress on the article

@@
 \begin{abstract}
+
 The development of new means to probe samples for the three-dimensional
 structure of DNA paves the way for in depth and systematic analysis of genome
-architecture, offering new insights in many biological processes. 3C-like
-methods can now assess physical interactions between pairs of loci in a
-genome-wide fashion, thus leading to the creation of contact maps. The
+architecture. 3C-like methods coupled with high-throughput sequencing can now
+assess physical interactions between pairs of loci in a genome-wide fashion,
+thus enabling the creation of genome-by-genome contact maps. The
 democratization of such protocols creates many new opportunities for
@@
 
-Recent studies assessed {\em P. falciparum} (the deadliest of the 5
-human malaria parasites) for its genome organization using 3C-like protocol.
-With its small genomic size, fairly simple yet changing genomic organization
-across its lifecyle and strong correlation between chromatin folding and gene
-expression, this parasite is the ideal case study for applying and developing
-methods to infer 3D models and use them for downstream analysis.
+Several recent studies applied such protocol to {\em P. falciparum} (the
+deadliest of the 5 human malaria parasites), assessing its genome organization
+at different moment of its life cycle. With its small genomic size, fairly
+simple yet changing genomic organization across its lifecyle and strong
+correlation between chromatin folding and gene expression, this parasite is
+the ideal case study for applying and developing methods to infer 3D models
+and use them for downstream analysis.
 
@@
 genome-wide fashion, yielding information on their relative spatial distance
-in the nucleus. Hi-C has opened new avenues for more systematic analysis of
+in the nucleus \citep{lieberman-aiden:comprehensive}. 
+Hi-C has opened new avenues for more systematic analysis of
 the three-dimensional folding of the genome, paving the way for a better
@@
 replication timing, epigenetic changes as well as many other biological
-processes. The 3D genome of a wide variety of organisms has been studied in
+processes \citep{de:dna, dixon:topological, rao:3d}. 
+The 3D genome of a wide variety of organisms has been studied in
 the past year, including many species of yeasts \citep{duan:three-dimensional,
@@
 plants \citep{feng:genome-wide, wang:genome-wide} and numerous human and mouse
-cell lines \citep{lieberman-aiden:comprehensive, rao:3D}. Within the past few
+cell lines \citep{lieberman-aiden:comprehensive, rao:3d, dixon:topological}. Within the past few
 years, two studies of the three-dimensional structure of {\em P. falciparum}
@@
 understanding the properties of genome folding and its link to gene
-regulation.
+regulation \citep{ay:three-dimensional, lemieux:genome-wide, ay:multiple}.
 
@@
 of {\em P. falciparum} (as well as many other organisms) relied on the
-inferring accurate 3D models of the structure. These models were then used as
+inference of accurate 3D models of the structure
+\citep{duan:three-dimensional, ay:three-dimensional, bau:three-dimensional}. These models were then used as
 a preliminary step to various analysis, such as identifying colocalized
@@
 least, integrating with other sources of data, such as gene expression or
-chromatin modification. In recent years, a profusion of methods were developed
+chromatin modification \citep{duan:three-dimensional, bau:three-dimensional,
+ay:three-dimensional}. In recent years, a profusion of methods were developed
 for creating 3D models of the genome, either as standalone methods or with the
-specific goal to better understand an organism, a biological process (such as
-the inactivation of chromosome X) or the folding of a chromosome. These
+specific goal to better understand an organism
+\citep{umbarger:three-dimensional, ay:multiple, duan:three-dimensional} or
+a biological process (such as
+the inactivation of chromosome X) \citep{bau:three-dimensional, deng:bipartite}. These
 methods broadly fall into two categories: ``model-based'' and ``data-driven''.
@@
 \includegraphics[width=0.5\linewidth]{schemas/schema.pdf}
-\caption{The art of understanding 3D genome structure}
+\caption{Understanding 3D genome structure using contact maps}
 \label{Fig1}
@@
 
-\section*{``Data-driven'' three-dimensional models of DNA from contact maps}
+\section*{Inferring three-dimensional models of DNA from contact maps}
 
@@
 plethora of methods, some developed for a particular organisms, others with
-the hope to be generalizable to any data sets. Methods model chromosomes as a
+the hope to be generalizable to any data sets \citep{dekker:capturing,
+duan:three-dimensional, tanizawa:mapping, ay:three-dimensional,
+peng:sequencing, lesne:3d, ben-elazar:spatial, zhang:inference, trieu:large,
+trieu:MOGEN, trieu:3D, bau:three-dimensional, umbarger:three-dimensional,
+rousseau:three, hu:bayesian}. Methods model chromosomes as a
 series of beads, and attempt to place the beads in a 3D euclidean space to
@@
 
-\citet{lesne:3d} proposes a different approach to convert counts into
-distances: they borrow the concept of shortest path from graph theory to
-create a distance matrix. A graph is constructed from the contact map by
+\citet{lesne:3d, hirata:three-dimensional} proposes a different approach to convert counts into
+distances: they borrow concepts from graph theory to
+create a distance matrix (either using shortest-path algorithm or Djikstra
+algorithm). A graph is constructed from the contact map by
 considering each loci as a node. Loci seen interacting are connected through
@@
 confidence in ``wish-distances'' \citep{ay:three-dimensional}, or constraints
-to reflect prior knowledge on the structure. See table~\ref{table:mds_detail}
-for a summary of the characteristic of each method.
+to reflect prior knowledge on the structure \citep{ay:three-dimensional,
+dekker:capturing, duan:three-dimensional} (see Table~\ref{table:mds_detail}
+for a summary of the characteristic of each method).
 
@@
 
-\fixme{maybe skip this section}
-
 \subsubsection*{Statistical models for contact counts}
@@
 
-{\em Ensemble approaches} aim at inferring a population of structures
+{\em Ensemble} approaches aim at inferring a population of structures
 representative of the contact count map. The methods fall into two distinct
@@
 type proposes a statistical modeling of the problem and samples the posterior
-distribution \citep{rousseau:three, hu:bayesian}. A parallel can be made
-between the first group of methods and {\em Consensus} MDS type of approaches,
-and the second with {\em Consensus} statistical based methods.
+distribution \citep{rousseau:three, hu:bayesian}. In short, the former is the
+{\em ensemble} version of MDS-based and {\em ad hoc} methods, while the former the
+{\em ensemble} version of statistical based methods.
 
@@
 into contact or ensure a minimal or maximal distance between those. The models
-includes three different types of restraints: (i) beads seen interacting are
+includes two different types of restraints: (i) beads seen interacting are
 restrained with harmonic oscillators of strengths derived from the contact
 counts; (ii) adjacent beads are ensured to be neither too close nor too far
-from one another; (iii) the last oscillators are activated only when
-constraints are not fulfilled to \todo{XXX}.  This yields an optimization problem
-with a large number of local minima, which the authors sample from by running
-50,000 minimizations starting from random starts.
+from one another.  This yields an optimization problem with a large number of
+local minima, which the authors sample from by running 50,000 minimization
+starting from random starts.
 
@@
 
-Akin to {\em consensus} MDS-like method and {\em ensemble}-local minima
-methods,  the first approach is to consider each contact as a constraint and
-to formulate an under constrained optimization problem. A population of
-structures satisfying the constraints can be found by sampling local
-minimas.
-
-\citet{paulsen:manifold} proposes a very different method, and formulate a
-\textit{manifold based optimization}, where a low rank PSD matrix (and thus a
-distance matrix) is optimized to be as close as possible to the sparse contact
-count matrix. Applying classical MDS on this low rank PSD matrix then yields a
-\textit{consensus} 3D model of the genome.
-
-\fixme{hirata et al}
+Akin to {\em ensemble}-local minima methods,  the first approach is to
+consider each contact as a constraint and to formulate an under constrained
+optimization problem. A population of structures satisfying the constraints
+can be found by sampling local minimas \citep{nagano:single-cell}.
+Akin to {\em consensus method}, one can attempt to construct a distance
+matrix, either through manifold-based optimization \citep{paulsen:manifold}
+(by finding a low rank PSD approximation of the sparse contact map) or
+akin to ShRec3D, through graph-theory considering the weighted graph of
+interactions \citep{hirata:three-dimensional}. A classical MDS method applied
+to such distance matrix then yields a \textit{consensus} 3D model of the genome.
 
@@
 \scriptsize{\cite{nagano:single-cell}} & & E & & & \\
-\scriptsize{\cite{paulsen:manifold}} & & C & \CHECK & & \CHECK \\
+\scriptsize{\cite{paulsen:manifold}} & MBO & C & \CHECK & & \CHECK \\
+\scriptsize{\cite{hirata:three-dimensional}} & RPR & C & \CHECK & & \\
 \hline
@@
 
-\section*{The art of modeling genome architecture.}
+\section*{The art of modeling genome architecture}
 
@@
 opposite manner. Consider the following task. You have a tens of thousands of
-randomly placed beads-on-strings. Can you find the smallest sets of
+randomly placed beads-on-a-string. Can you find the smallest sets of
 constraints such that these beads-on-a-string interact overall the same way as
@@
 contact maps from these models match as closely as possible the observed
-contact maps. These ``model-driven'' approaches offer mechanistical insights
-into the genome architecture.
+contact maps. These ``model-driven'' approaches offer powerful mechanistical
+insights into the genome architecture, but are difficult to build in practice:
+each organism, cell type, and time point require hand crafted sets of
+constrained, build by iteratively improving models.
 

### 2017-07-30 scripts/population_structures/clustering_across_stages_downsampled_distances.py
Deleting the old paper

@@ -9,7 +9,9 @@ parser.add_argument("--algorithm", "-a", default="MDS")
 parser.add_argument("--chromosome", "-c", default=None, type=int)
 args = parser.parse_args()
 
-datasets = ["rings", "schizonts", "trophozoites"]
+datasets = ["ay2013/rings_10000",
+            "ay2013/schizonts_10000", "ay2013/trophozoites_10000",
+            "lemieux2013/25kb/B15C2_combined"]
 algorithm = args.algorithm
 
 labels = None
@@ -17,11 +19,11 @@ X = None
 for i, dataset in enumerate(datasets):
     if args.chromosome is None:
         filename = (
-            "results/ay2013/%s_10000_raw_%s_distances_downsampled.npy" % (
+            "results/%s_raw_%s_distances_downsampled.npy" % (
                 dataset, algorithm))
     else:
         filename = (
-            "results/ay2013/%s_10000_raw_%s_distances_chr%02d.npy" % (
+            "results/%s_raw_%s_distances_chr%02d.npy" % (
                 dataset, algorithm, args.chromosome))
     x = np.load(filename)
     x = x[:1000]
@@ -45,7 +47,7 @@ colors = ['#FFB773', '#FF7C00', "r",  # '#A65100',
           '#04859D', 'skyblue',
           '#015666', 'green', 'darkgreen']
 markers = [".", ".", ".", "D", "D", "D"]
-stages = ["Rin", "Sch", "Trop"]
+stages = ["Ay-Rin", "Ay-Sch", "Ay-Trop", "Lemieux-Rin"]
 
 pca = PCA(n_components=3, copy=False, random_state=1, svd_solver="randomized")
 X = pca.fit_transform(X)

### 2017-07-30 scripts/population_structures/compute_distances.sh

@@ -1,72 +1,16 @@
 python compute_downsampled_distances.py \
-  structures/ay2013/trophozoites_10000_raw_MDS \
+  structures/ay2013/rings_10000_raw_PO \
   --lengths data/ay2013/rings_10000_raw.bed
-
-python compute_downsampled_distances.py \
-  structures/ay2013/trophozoites_10000_raw_UNB0 \
-  --lengths data/ay2013/rings_10000_raw.bed
-
-python compute_pairwise_distances_chr.py \
-  structures/ay2013/trophozoites_10000_raw_MDS \
-  --lengths data/ay2013/rings_10000_raw.bed
-
-python compute_pairwise_distances_chr.py \
-  structures/ay2013/trophozoites_10000_raw_UNB0 \
-  --lengths data/ay2013/rings_10000_raw.bed
-
-
-python compute_downsampled_distances.py \
-  structures/ay2013/schizonts_10000_raw_MDS \
-  --lengths data/ay2013/rings_10000_raw.bed
-
-python compute_downsampled_distances.py \
-  structures/ay2013/schizonts_10000_raw_UNB0 \
-  --lengths data/ay2013/rings_10000_raw.bed
-
-python compute_pairwise_distances_chr.py \
-  structures/ay2013/schizonts_10000_raw_MDS \
-  --lengths data/ay2013/rings_10000_raw.bed
-
-python compute_pairwise_distances_chr.py \
-  structures/ay2013/schizonts_10000_raw_UNB0 \
-  --lengths data/ay2013/rings_10000_raw.bed
-
-
-python compute_downsampled_distances.py \
-  structures/ay2013/rings_10000_raw_MDS \
-  --lengths data/ay2013/rings_10000_raw.bed
-
-python compute_downsampled_distances.py \
-  structures/ay2013/rings_10000_raw_UNB0 \
-  --lengths data/ay2013/rings_10000_raw.bed
-
-python compute_pairwise_distances_chr.py \
-  structures/ay2013/rings_10000_raw_MDS \
-  --lengths data/ay2013/rings_10000_raw.bed
-
-python compute_pairwise_distances_chr.py \
-  structures/ay2013/rings_10000_raw_UNB0 \
-  --lengths data/ay2013/rings_10000_raw.bed
-
-python compute_downsampled_distances.py \
-  structures/ay2013/trophozoites_10000_raw_PO \
-  --lengths data/ay2013/rings_10000_raw.bed
-
[…]

### 2017-07-30 scripts/population_structures/compute_downsampled_distances.py

@@ -12,9 +12,10 @@ from iced.io import load_lengths
 parser = argparse.ArgumentParser()
 parser.add_argument("directory")
 parser.add_argument("--lengths", "-l")
+parser.add_argument("--factor", type=int, default=10)
 args = parser.parse_args()
 
-
+factor = args.factor
 if args.lengths is not None:
     lengths = load_lengths(args.lengths)
 else:
@@ -43,7 +44,7 @@ for i, filename in enumerate(filenames):
     dis = euclidean_distances(X)
     dis[mask] = np.nan
     dis[:, mask] = np.nan
-    dis, lengths_ = downsample_resolution(dis, lengths, factor=10)
+    dis, lengths_ = downsample_resolution(dis, lengths, factor=factor)
     distances.append(dis[np.triu_indices(dis.shape[0])][np.newaxis])
 
 distances = np.concatenate(distances)
